# (PART) FOUNDATIONS {-}

# Mathematics Education Foundations {#ch:MathEdFoundations}

This first chapter focuses on developing the notion of Mathematical Knowledge for Teaching and examining how content knowledge supports the current standards for mathematical practice and the content standards developed by various educational researchers and mathematics education organizations. Understanding these foundations prior to jumping into the rest of the content provides both an organizational framework and a foundation for the topics covered. Some of these ideas may challenge your ways of thinking, while others reinforce what you already know. We encourage you to go beyond what is written here and to read some of the original research articles and organization standards cited in this chapter.


```{r, echo=FALSE, fig.align='center', out.width='70%', fig.cap=''}
knitr::include_graphics(
  'images/Parthenon_Nashville_bw.jpg',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

\newpage

## Mathematical Knowledge for Teaching {#section:MKT}

In his presidential address to the American Educational Research Association, Lee Shulman (@shulman1986) popularized the concept of knowledge for teaching that included a specialized content knowledge where "the teacher need not only understand *that* something is so; the teacher must further understand *why* it is so, on what grounds its warrant can be asserted, and under what circumstances our belief in its justification" (p. 9). Researchers in mathematics education further expanded on this idea in the development of domains of Mathematical Knowledge for Teaching. This textbook focuses on the Subject Matter Knowledge side of the Mathematical Knowledge for Teaching. 


```{r, echo=FALSE, fig.align='center', out.width='100%', fig.cap='Domains of Mathematical Knowledge for Teaching [@Ball2008]'}
knitr::include_graphics(
  'tikz/ballegg1.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```


### Common Content Knowledge

The foundation for much of the other aspects of mathematical knowledge for teaching is **common content knowledge**, defined as "the mathematical knowledge and skill used in settings other than teaching" [@Ball2008, p. 399]. Common content knowledge provides the individual the ability to solve mathematical problems from the related curriculum and to apply the mathematical knowledge to other fields of knowledge outside of mathematics. Common content knowledge also includes understanding how some of the different mathematical subjects build upon one another, but does not require those studying the material to be able to explain broader patterns. Thus, common content knowledge represents the type of knowledge that we expect secondary and introductory post-secondary students to display. 


For instance, common content knowledge related to rational algebraic expressions and functions includes understanding the relationships between rational expressions and rational numbers, how long-division of polynomials relates to long-division of integers, and the properties of the quotient and remainder of a rational expression and its effect on the graph of the related function. While this common content knowledge includes a deeper level of knowledge and a richer web of conceptual understanding by the teacher than is common for most K-12 students, the knowledge still falls within the expectations for the most advanced of these K-12 students. Thus, it is paramount that secondary mathematics teachers develop deep fluency in common content knowledge and the interconnectedness of the many different mathematical concepts in the curriculum. Such a deep and interconnected knowledge base of the teacher provides a requisite foundation in helping K-12 students learn mathematics. However, it is only a first step.

### Specialized Content Knowledge

**Specialized content knowledge** incorporates mathematical knowledge that goes beyond knowledge expected of students, with a focus on the mathematical knowledge that improves the ability of the teacher to assist students learning mathematics [@Ball2008, p. 400]. Thus, specialized content knowledge supplements the common content knowledge that all K-12 students of mathematics need. For example, while common content knowledge for rational algebraic expressions and rational functions include the relationships to the rational numbers, specialized content knowledge could include  knowledge of rings and integral domains, thereby improving the teacher's ability to help students to make the connections between various pieces of related content knowledge. 

A teacher would also use this specialized content knowledge related to rational expressions when explaining the extent to which two rational expressions are equivalent when common factors of the numerator and denominator cancel. For instance the teacher could explain the ways in which the expressions 
$$ \frac{(x-1)(x+2)^2}{(x+1)(x+2)} \quad \mbox{and} \quad \frac{(x-1)(x+2)}{(x+1)}$$ are equivalent and distinct.

### Horizon Content Knowledge

Excellent teachers use more than just a knowledge of the content in the current curriculum when teaching students. They also draw on knowledge of what their students have previously learned in mathematics, what mathematics content will be covered in the next few years, and how the current mathematical topic relates to applications outside of mathematics. That is, excellent mathematics teachers see their instruction as part of a continuum, of which their work is only a small part. We call this domain of knowledge **horizon content knowledge**. 

For example, a teacher with horizon content knowledge might use students' familiarity with rational numbers to help high school students develop knowledge of rational algebraic expressions and functions. She could also use her knowledge from differential equations to know that the rational functions play a pivotal role in the Laplace transform, helping to determine the amount of time and detail appropriate for teaching about the quotient and remainder theorem for polynomials and how to rewrite rational expressions using partial fractions. The teacher could also use knowledge of the physics curriculum and Boyle's law to help students make connections between rational expressions and functions and other fields of study. While each of these pieces of knowledge are not essential for mathematics teachers, the more knowledge one has, the better that teacher can help students learn and apply the critical components of the content.



### Exercises

1. Consider the general quadratic expression $ax^2 + bx+c$, where $a$, $b$, and $c$ are real numbers such that $a\neq0$.

    a. Write a list of everything you **know** about the general quadratic expression.

    b. Write a list of everything you can **do** to the general quadratic expression.

    c. How are quadratic expressions different than quadratic equations?

    d. How does the factorization of a quadratic expression relate to prime numbers?

    e. How much do you know about the ways in which quadratic expressions are used outside of mathematics?

    f. Review the mathematics standards for quadratic expressions for your state. How many of the things you have listed for parts (a) and (b) match up with those standards?

    g. For each of your answers to a. through e., does the content you have listed or described align most with common content knowledge, specialized content knowledge, or horizon content knowledge?

2. Consider a triangle. A particular high school geometry textbook defines a triangle as "a polygon with three sides." A second textbook defines a triangle as "the figure formed by connecting three non-co-linear points with straight segments." A last textbook defines triangles as "A three-sided figure."
    
    a) Are all three definitions accurate, or do some allow for shapes that might not be triangles as you understand them to be included within the category of triangle?   
  
    b) In what ways might each definition be considered sloppy? That is, are there any parts of the definition that might not be well-defined? In what ways might each definition make using triangles in future lessons more difficult? 

    c) What information do you think each textbook has presented prior to giving its definition for a triangle?

    d) How might horizon content knowledge help an instructor preparing a lesson on triangles decide whether a definition is appropriate or not for her students?







## Mathematical Practice Standards

Just like almost all of the content taught in secondary schools, the development of procedural habits and pieces of information are important parts of a secondary education, the heart of learning lies in developing habits of thinking, perseverance techniques, and developing communication skills to improve the ability to interact with the world around them. For example, the United Nations Educational, Scientific, and Cultural Organization (UNESCO) and the United Nations Office on Drugs and Crime (UNODC) [-@UNESCO] described cognitive learning outcomes for secondary education such that a student ``Knows about local, national, and global governance and accountability systems and structures, understands issues affecting interaction and connectedness of communities at local, national and global levels, (and) develops skills for critical inquiry and analysis'' (p. 16). Mathematics, despite it's reputation, aligns well with these goals. Critical thinking, reasoning, and communicating are the most important components of the secondary mathematics curriculum, and are often the most ignored.

The National Council of Teachers of Mathematics [-@PSSM] describes these goals in the -Principles and Standards for School Mathematics-, giving them the name 'Process Standards.' The National Governors Association Center for Best Practices and the Council of Chief State School Officers [@CCSS] expanded upon these to create the 'Common Core State Standards Standards for Mathematical Practice.' These practice standards are listed in Table \@ref(tab:SMPs)  and more details about these standards can be found in their corresponding publications.


```{r SMPs, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
NCTMPS <- c('Problem Solving', 'Reasoning and Proof', 'Communication', 'Connections', 'Representations', '', '', '')
CCSMP <- c('Make sense of problems and persevere in solving them.', 'Reason abstractly and quantitatively.', 'Construct viable arguments and critique the reasoning of others.', 'Model with mathematics.', 'Use appropriate tools strategically.', 'Attend to precision', 'Look for and make use of structure.', 'Look for and express regularity in repeated reasoning.')
SMPs <- data.frame(CCSMP, NCTMPS)
names(SMPs) <- c('Standards for Mathematical Practice', 'Process Standards')
knitr::kable(SMPs, caption = 'NCTM Process Standards and  Common Core Standards for Mathematical Practice')
```

It is worth noting that these practices are expected of students of mathematics at *all* grade levels. In order to help others develop these practices, we must first develop them in ourselves. Only then will we be in a position to create a learning environment that motivates and enables students to grow in these practices.

To support the users of this text in developing these practices, opportunities to practice them are woven into the text, exercises, and projects. However, as any good teacher knows, the learner needs to be actively engaged in order for an objective to be reached. As you work through the text, pay attention to how the arguments are presented and seek to understand the process behind the mathematical content. When completing the exercises and projects, do not just try to get an answer. Instead, take some time to grapple with the ideas, think of better ways to communicate what you do not understand, and seek to understand the deeper connections involved in the task.

For the purpose of this text we group these practice standards into four categories: mathematical problem solving, modeling with mathematics, communicating mathematically, and understanding mathematical structures. We briefly elaborate on each in the following sections.


### Mathematical Problem Solving

Mathematical problem solving is perhaps the most widely cited application of mathematics ``in the real world''. Although generally acknowledge as important, problem solving is a complex process that is difficult to teach. In his book *How  to  Solve  It: A  new  aspect  of  mathematical  method*, George Pólya described four phases of the problem solving process (See Figure \@ref(fig:polya) when approaching mathematical problems [@Polya1945]. While many others have expanded on this problem solving process, much of the thinking around problem solving still traces back to the four phases that Pólya describes.

```{r polya, echo=FALSE, fig.align='center', out.width='45%', fig.cap="Pólya's problem solving process"}
knitr::include_graphics(
  'tikz/polya.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

The first phase is the process of understanding the problem. What this phase looks like varies by problem, but often involves making sure you understand what information is given and relevant. It also includes knowledge of what an answer should look like for the question posed and, where relevant, being able to organize the provided information in a picture or diagram to better understand the situation. So, for example, if the problem is computational, understanding the problem may entail understanding whether positive or negative answers are valid solutions. If the problem is a theorem that needs proving, this phase may involve generating related examples or writing out relevant definitions to make sure you understand the given inputs and implications of the theorem. 

The second phase of the problem solving process is to devise a plan. Many students struggle in the problem solving process because they only want to solve problems for which they have previously been given a template. The origin of this desire has roots in the fact that many teachers only assign students problems that are similar to those discussed in class. True mathematical problem solving involves confronting problems for which the solution is not immediately obvious to the problem solver. In problem solving, one may need to connect the current problem to previously solved problems or restate the problem in a different form with an easier solution process. Sometimes a full plan is not possible at the beginning and the solver needs to just plan initial steps and work through those to gather more information about the problem that will help them create a new plan later in the process.

Once an initial plan is devised, it needs to be carried out. This is often the simplest part of the problem solving process, and often the only component that students complete. With today's technology, computer applications can often complete the details of this part of the problem solving process through statistical analyses, computer algebra systems, or graphical programs.

After the plan is carried out, it is important to look back, completing the final phase of the problem solving process. In this step the problem solver determines whether their solution makes sense in the context of the original problem. For example, in calculus, it is often possible to obtain negative solutions that are not appropriate to the problem posed. The final step of looking back also entails making sure that we have actually addressed the question asked or the problem posed, rather than a different, somewhat related, question or problem.

Although these four phases appear linear, the majority of mathematical problems require various iterations of these phases. In particular, it is often the case that the first plan created to solve a problem does not work, so it has to be revised following an attempt to carry it out. Good problem solvers see these revisions as opportunities to learn more about the problem, rather than as failure at the problem solving process.

### Modeling with Mathematics

Mathematical problem solving often relies on modeling with mathematics, particularly in the context of real-world phenomena. The Guidelines for Assessment and Instruction in Mathematical Modeling Education (GAIMME) defines mathematical modeling as "a process that uses mathematics to represent, analyze, make predictions or otherwise provide insight into real-world phenomena" [@GAIMME, p. 8]. This definition varies from other common connotations of ``modeling'' that are used in education, including in mathematics. For example in mathematics education, using manipulatives to model a mathematical idea; sketching graphs or pictures to communicate or understand concepts; and demonstrating how to solve certain types of problems are all referred to as modeling. However, as a process standard, modeling mathematics does not generally include these activities. Instead, it refers to activities of using mathematics to analyze, predict, and represent real-world data. 

The GAIMME Report breaks the mathematical modeling process into the six components shown in Figure \@ref(fig:GAIMME). Critically, these six components **do not** happen in a progression, but instead are iterative and sometimes run parallel to each other.

```{r GAIMME, echo=FALSE, fig.align='center', out.width='70%', fig.cap="The Math Modeling Process [@GAIMME, p. 13]"}
knitr::include_graphics(
  'tikz/GAIMME.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

The focus on trying to better understand real-world phenomena distinguishes mathematical modeling from application problems or word problems. Actual real-world phenomena are distinctly messy and arriving at a final solution requires interpretation and assumptions. Moreover, an individual engaged in modeling has to confront the challenge of determining whether the question, mathematical model, and data are well-aligned. Teachers face the additional challenge of ensuring that the situations they create for their students are appropriate for their students. For instance, an elementary school classroom might plant some seeds in the soil. The teacher would then assist the students in identifying specific questions that can be answered quantitatively about the seeds and plants, guiding them towards a set of questions and data that the children could collect, analyze, and interpret. A secondary teacher teaching about exponential functions may introduce the class to the concept of population growth and guide the students to questions that can be answered quantitatively with exponential models. In both these examples, the instructor introduces the students to a real-world situation and then guides them to questions that facilitate the students' development of the mathematical modeling they want to discuss.

Another key aspect of the modeling process involves determining the relevance of various quantities in the situation and how to describe them with variables. This cyclical process involves identifying variables, assigning them labels and then assessing how they fit in with other variables. This process allows the modeler to create an idealized version of the original problem in order to create some type of solution. It is important during the process to justify each assumption made and to clearly label each variable, along with its appropriate units. It is through this justification process that the modeler can communicate the applicability and interpretation of the results from the model to the consumer.

As the modeler defines the variables and the relationships between them, she can use the mathematical problem solving techniques to 'do the math' and come up with possible solutions to the problems posed. This section of 'doing the math' reflects the word problems of most math textbooks since the word problems rarely have extraneous information and have obvious variables defined with transparent relationships between them.

Similar to the 'looking back' phase of the mathematical problem solving process, the modeling process includes a stage during which the modeler steps back and assesses the model. During this stage he or she analyzes and tests the model and solution, examining the model created and determining the appropriateness and accuracy of the solution produced. This process often results in a revision to the original assumptions and adjustments to the model with improved variables and relationships between the variables.

Since the mathematical modeling process deals with messy real-world phenomena, an iterative process of reflection that results in the refinement and, where appropriate, extension, of the the model is a key component that almost always appears in a modeling cycle. This process may run parallel to or between any of the other phases of the modeling process.

When the model gets to a state that satisfies the modeler in terms of the model justifications and appropriateness of the solutions, the implementation of the model and the reporting of the results occurs. The presentation of the results must include the justification for the model, a description of the various assumptions made to produce the model, and any limitations the model has in terms of the accuracy of the results. This presentation must be done in a way that is comprehensible to the desired audience. Because modeling deals with real-world situations, the solution is unlikely to be clear or definitive, but will instead be approximate and estimated.

### Communicating Mathematically

Mathematical communication broadly relates to the ways and methods of representing, justifying, and interpreting mathematics. More specifically, the practice of mathematical communication "encompasses both listening and reading (comprehension) and both speaking and writing (expression)... and may also include representation of mathematical ideas in nonlinguistic ways" [@Communication, p. 7]. In order to better understand the practice of communicating mathematically, we examine six overlapping concepts that are at the core of this practice (see Figure \@ref(fig:communication)).

```{r communication, echo=FALSE, fig.align='center', out.width='60%', fig.cap="Components of Mathematical Communication"}
knitr::include_graphics(
  'tikz/communication.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

All types of communication require a language. Within a language there exists a shared meaning behind the words and symbols used when communicating with a particular language. Mathematics communication, in this sense, operates much as a language does. However, unlike a language such as English, mathematics is represented using a mixture of words that have very precise definitions and symbols that compress many different ideas and meanings into a very small amount of space (linguists call mathematics a ``dense'' language because of this property). Moreover, somewhat uniquely to mathematics, the meaning of symbols varies widely depending upon the context of the communication. Consider variables, which are usually represented using any number of symbols. While certain types of symbols will cue a well-informed mathematics student that a variable has some implied meaning in terms of what type of number system to which it belongs, even in this case variables in different circumstances serve different roles. Sometimes the letter represents a fixed value (e.g., $x=4$) and other times it represents a range of values that satisfy a particular condition (e.g., $x\leq 4$). Sometimes it indexes a set, while other times it represents a function. The differences between these uses are often subtle, making explicit training in these nuances mathematics critical.  


Once the language of mathematics has been developed, the grammar is created. In mathematics, logic, argument, and proof form the core patterns of organization and presentation. Together, logic, argument, and proof boil down to the ability to construct viable arguments and critique the arguments of others using mathematical evidence. 

Part of mathematical communication is learning how to consume and understand the ideas of others. In mathematics this usually occurs by reading or listening to the arguments of others. Both reading and listening to mathematics require a great deal of practice and are skills students must hone over many courses. In particular, reading mathematics requires a great deal of practice and discipline. While a novel can usually be consumed at a steady pace without much explicit effort by the reader, written mathematical ideas are often presented symbolically, with the reader left to fill in certain logical steps. Thus, the act of reading mathematics becomes one of translation of the symbolic work into a language he or she understands, followed by a effort to connect current parts of the argument to previously made points, and the absorption and evaluation of those ideas based on previous knowledge. As a result, reading mathematical texts and arguments requires a big investment in time. Sometimes it can take hours just to understand one line of a mathematical text! This slow, methodical way of reading is challenging to students of mathematics, but is a critical skill to develop as it is core to the practice of doing mathematics.

Previous paragraphs have described the difficulties and process of consuming mathematical texts. However, it is also important that students learn to produce their own mathematical communications in both oral and written forms. In communicating to others, it is essential that precise mathematical words are used, rather than slang, in order to reduce the confusion of potential readers. It is also useful to employ a variety of mathematical representations, including graphs, sketches, and diagrams to facilitate the communication of mathematical ideas.

As teachers of mathematics, we must continually emphasize the development of this practice of communicating mathematically by de-emphasizing simple answers to mathematical problems, teaching students how to read and write mathematics (not just about mathematics), and to promote classroom discourse to provide opportunities for students to experience different types of mathematical communication.


### Understanding Mathematical Structures

While the definition of mathematics is not uniform or agreed on, most agree that mathematics is inherently about the study of underlying structures and logical reasoning. For example, the Encyclopædia Britannica -@definition defines mathematics as "the science of structure, order, and relation that has evolved from elemental practices of counting, measuring, and describing the shapes of objects." This means that an important practice in the doing of mathematics is to "look for and make use of structure" and to "look for and express regularity in repeated reasoning" [@CCSS].

The pursuit and use of structure and regularity appear throughout all of mathematics. Mathematical structure informs every part of mathematics teaching, from instruction on common mathematical procedures to searching for and making connections between various, apparently unrelated, mathematical content areas to provide new insight into a problem. As teachers, we need to point out these mathematical structures to the students and help them learn how to discover the structures and regularity on their own.


### Exercises


1. Consider the following mathematical task: 
 <blockquote>An electricity company charges Kelly $\$0.15$ per kWh (kilowatt-hour) of electricity used, plus a basic connection charge of $\$8.00$ per month. Find a function that helps Kelly estimate her monthly electricity bill for a given number of kWhs. Be explicit about the domain of the function you create.</blockquote>

    a) NCTM emphasizes multiple representation of functions using verbal, tabular, graphical, and symbolic representations. Using this word problem as the verbal representation, express the function in each of the remaining ways.

    b) How does the ability to move between different representations of functions support students' development of the mathematical communication process standard? 

    c) The problem solving section discussed the fact that a mathematical task is only considered a problem if the person working on the task does not know in advance a solution method that will produce a correct answer. Would you say that for part (a) you were engaged in mathematical problem solving as described in the standards? Why or why not?

    d) Under what conditions would the task be a problem for students you were teaching?

    e) Based on the description of mathematical modeling, would you say that finding the function for the task in part (a) represents mathematical modeling? Why or why not?

    f) A second part of the task asks students to estimate Kelly's electrical bill for the year, given that her monthly kWh usage ranges between 202 and 254 kWh. 

    g) The textbook you got the problem from lists the answer to the extension as $\$506.40$. What did the author do and what assumptions did she make in order to arrive at that answer? Do you agree with her process and assumptions? Why or why not?

    h) In arriving at your answer for part (a), would you say that you were engaged in mathematical problem solving? Why or why not?

    i) If you were using this task as a modeling activity for your students, what criteria would you use to evaluate whether their answer was reasonable?


2. There are five NCTM Process Standards and eight Common Core Standards for Mathematical Practice. While there is significant overlap between the two sets of standards, they are not the same. Read each set of standards. When there is overlap between the two sets of standards, create a map between them. Also identify the ways in which the two sets of standards differ. These regions are not at the heading level. You'll have to actually dig into the blurbs about each standard in their original documents.

3. A basic theorem that students use from an early age is that the sum of two even integers is even. A typical proof of this theorem might look something like this:
<blockquote>Let $a$ and $b$ be even integers such that $a=2m$ and $b=2n$ where $m$ and $n$ are integers. Then $a+b=2m+2n=2(m+n)$. Thus, the sum of two even integers is even.</blockquote>

    a) The section on mathematical communication emphasizes the reliance of mathematical communication on precise definitions and symbols that compress complex ideas into short phrases. In examining this proof, identify the places where the problem statement or proof rely on a definition or compressed expression.

    b) Identify the places where the proof writer has left the reader to fill in logical steps or rationales. 

    c) Rewrite the proof of the theorem without the use of any symbols, but with the same degree of precision and generalizability. 

    d) In comparing the symbolic proof to your verbal proof, do you think it is easier to understand (i.e., more like a novel?) or do you think it would still be difficult to read if all mathematics were presented without symbols? Why or why not?

    e) If you reflect on the six parts of mathematical communication, which parts do you think you are best at? Which parts do you struggle the most with? What is one goal you have for yourself for mathematical communication that you want to develop during this course?


## Mathematics Content Standards in the U.S.

The current standards for mathematics and assessment in the United States are derived from a variety of sources and past initiatives. David Klein -@Klein2003 gives a good summary of the major events and influential individuals during the 20^th^ century. The current set of standards have origins to a pair of reports from the 1980's: one by NCTM -@NCTM1980, *An Agenda for Action*, and one by the U.S. National Commission on Excellence in Education -@NCEE1983, *A Nation at Risk*. Each provided a different view about what should occur to improve mathematics education in the United States. While these two documents were used as the basis for the 'math wars' of the late 20^th^ century, the biggest difference between them is that the NCTM document focused primarily on standards for teaching methodologies, while *A Nation at Risk* focused on standards of content knowledge.

The NCTM -@NCTM1980 recommended "that

1. problem solving be the focus of school mathematics in the 1980s;

2. basic skills in mathematics be defined to encompass more than computational facility;

3. mathematics programs take full advantage of the power of calculators and computers at all grade levels;

4. stringent standards of both effectiveness and efficiency be applied to the teaching of mathematics;

5. the success of mathematics programs and student learning be evaluated by a wider range of measures than conventional testing;

6. more mathematics study be required for all students and a flexible curriculum with a greater range of options be designed to accommodate the diverse needs of the student population;

7. mathematics teachers demand of themselves and their colleagues a high level of professionalism;

8. public support for mathematics instruction be raised to a level commensurate with the importance of mathematical understanding to individuals and society" (p. 1).

The National Commission on Excellence in Education -@NCEE1983 recommended that 

* "The teaching of *mathematics* in high school should equip graduates to: (a) understand geometric and algebraic concepts; (b) understand elementary probability and statistics; (c) apply mathematics in everyday situations; and (d) estimate, approximate, measure, and test the accuracy of their calculations. In addition to the traditional sequence of studies available for college-bound students, new, equally demanding mathematics curricula need to be developed for those who do not plan to continue their formal education immediately" (p. 25).

* "Standardized tests of achievement (not to be confused with aptitude tests) should be administered at major transition points from one level of schooling to another and particularly from high school to college or work. The purposes of these tests would be to: (a) certify the student's credentials; (b) identify the need for remedial intervention; and (c) identify the opportunity for advanced or accelerated work. The tests should be administered as part of a nationwide (but not Federal) system of State and local standardized tests. This system should include other diagnostic procedures that assist teachers and students to evaluate student progress" (p. 28).

* "Persons preparing to teach should be required to meet high educational standards, to demonstrate an aptitude for teaching, and to demonstrate competence in an academic discipline. Colleges and universities offering teacher preparation programs should be judged by how well their graduates meet these criteria" (p. 30).

* "Substantial nonschool personnel resources should be employed to help solve the immediate problem of the shortage of mathematics and science teachers. Qualified individuals including recent graduates with mathematics and science degrees, graduate students, and industrial and retired scientists could, with appropriate preparation, immediately begin teaching in these fields. A number of our leading science centers have the capacity to begin educating and retraining teachers immediately. Other areas of critical teacher need, such as English, must also be/addressed" (p. 31).

NCTM followed it's 1980 report with the publication of the \textit{Curriculum and Evaluation Standards for School Mathematics} \citep{NCTM1989}. This document focused on the standards for teaching methodologies and mathematical practices. Content standards were sketched out, with overviews of the recommended content knowledge for students in 4-year grade bands. NCTM expanded this work with additional texts focusing on teaching standards and assessment: \textit{Professional Teaching Standards} (\citeyear{NCTM1991}) and \textit{Assessment Standards} (\citeyear{NCTM1995}). In 2000, NCTM released an updated version the 1989 standards with the \textit{Principles and Standards for School Mathematics} \citep{PSSM}. During this same time, many states developed more specific content standards reflecting the guidelines recommended by the 1989 document \citep{Raimi1998}. Although the NCTM document did not make content recomendations by specific grade levels, many of the state standards still resided at the grade-band level \citep[p. 677]{Reys2007}.


### Common Core State Standards-Math

In 2002, Congress passed the No Child Left Behind Legislation. This bill required that states determine measurable content standards for each grade level and develop assessments based on these standards to be given to all students at specific grade levels (generally fourth, eighth, and twelfth grade) in order to receive federal school funding. Since each state developed their own standards, the level and specificity of these standards varied greatly between States [@Reys2007]. 

It was in this environment that the National Governors Association Center for Best Practices and the Council of Chief State School Officers launched an effort in 2009 to develop "a common core of internationally benchmarked standards in math and language arts for grades K-12" [@CCSS]. These standards emphasize content knowledge, while also encouraging some of the teaching methodologies proposed by the NCTM documents.

In order to help teachers better understand how the topics in this book relate to the actual teaching and learning of their students, we include related content standards from the Common Core State Standards for Mathematics throughout the text. While not all states have adopted the Common Core State Standards for their school systems, these standards are a good representation of what is expected from students at different stages in their mathematical education.

>#### Related Content Standards {-}
>- (8.F.1) Understand that a function is a rule that assigns to each input exactly one output. The graph of a function is the set of ordered pairs consisting of an input and the corresponding output.
>- (HSF.IF.7) Graph functions expressed symbolically and show key features of the graph, by hand in simple cases and using technology for more complicated cases.

 Standards that align with the content of each section are provided in blocks with the title "Related Content Standards." One such box is given above as an example. The sequence of numbers and letters proceeding each standard is its standardized abbreviation. The first part of the abbreviation identifies the grade band. Since the Common Core State Standards for high school do not have a recommended grade level for each of the standards, they are denoted by "HS" and the general area of mathematics.  The second part of the abbreviation refers to the general domain of mathematics, and the third part refers to the location of the standard in that grade band and domain.


### NCTM-CAEP Standards

Another requirement of the No Child Left Behind legislation was that states had to create standards for teachers to receive certification. These standards for certification needed to be measured with a blend of standardized assessments, portfolios, and coursework. As a result, certification in secondary mathematics in most states requires a major in mathematics (or its equivalent), the teaching of and about standards during certification coursework, and achievement of certain scores on a content specific standardized test (such as the PRAXIS). In addition, the NCTM partnered with the Council for the Accreditation of Educator Preparation (CAEP) to develop standards for what beginning secondary mathematics teachers should know and be able to do in the teaching of mathematics [@CAEP]. 

The NCTM-CAEP content standards are built upon the standards developed by the Conference Board of Mathematical Sciences in -@MET1 and -@MET2. The content of this book builds upon the contents of the first standard of "Knowing and Understanding Mathematics". In particular: 

>Candidates demonstrate and apply, with the incorporation of mathematics technology, conceptual understanding, procedural fluency, and factual knowledge of major mathematical domains: Number; Algebra and Functions; Statistics and Probability; Geometry, Trigonometry, and Measurement; Calculus; and Discrete Mathematics.


We list the details of the NCTM-CAEP content standards below and discuss how this book connects to each of these standards.



>**Essential Concepts in Number.** Candidates demonstrate and apply conceptual understanding, procedural fluency, and factual knowledge of number including flexibly applying procedures, using real and rational numbers in contexts, developing solution strategies, and evaluating the correctness of conclusions. Major mathematical concepts in Number include number theory; ratio, rate, and proportion; and structure, relationships, operations, and representations.

 
 Number systems are primarily covered in Chapters \@ref(ch:number), \@ref(ch:group-1), and \@ref(ch:rings). We introduce initial concepts of the number systems in Chapter \@ref(ch:number), focusing on operations and the basic properties of common number systems, including whole numbers, integers, rational, real, and complex numbers. Chapters \@ref(ch:group-1) and \@ref(ch:rings) expand these ideas through the added structure of rings and fields.


    
>**Essential Concepts in Algebra and Functions.** Candidates demonstrate and apply understandings of major mathematics concepts, procedures, knowledge, and applications of algebra and functions including how mathematics can be used systematically to represent patterns and relationships including proportional reasoning, to analyze change, and to model everyday events and problems of life and society. Essential Concepts in Algebra and Functions include algebra that connects mathematical structure to symbolic, graphical, and tabular descriptions; connecting algebra to functions; and developing families of functions as a fundamental concept of mathematics. Additional Concepts should include algebra from a more theoretical approach including relationship between structures (e.g., groups, rings, and fields) as well as formal structures for number systems and numerical and symbolic calculations.


Many of these concepts are spread throughout the text, as algebra and functions are fundamental to the secondary curriculum. That said, Chapters \@ref(ch:function), \@ref(ch:group-1), \@ref(ch:rings), and \@ref(ch:real-valued-functions) specifically focus on topics of particular importance in understanding algebra and functions. In particular Chapter \@ref(ch:real-valued-functions) combines the themes of the earlier chapters together using real-valued functions.



    
>**Essential Concepts in Calculus.**  Candidates demonstrate and apply understandings of major mathematics concepts, procedures, knowledge, and applications of calculus including the mathematical study of the calculation of instantaneous rates of change and the summation of infinitely many small factors to determine some whole. Essential Concepts in Calculus include limits; continuity; the Fundamental Theorem of Calculus; and the meaning and techniques of differentiation and integration.


The essential concepts in calculus are currently excluded from this text with the understanding that most pre-service teachers are already required to take at least two semesters worth of calculus courses that focus on this content standard.




    
>**Essential Concepts in Statistics and Probability.**  Candidates demonstrate and apply understandings of statistical thinking and the major concepts, procedures, knowledge, and applications of statistics and probability including how statistical problem solving and decision making depend on understanding, explaining, and quantifying the variability in a set of data to make decisions. They understand the role of randomization and chance in determining the probability of events. Essential Concepts in Statistics and Probability include quantitative literacy; visualizing and summarizing data; statistical inference; probability; and applied problems.


The essential concepts of statistics and probability are the focus of Part IV of the text, Data Analysis. In that section we focus on statistics and a study of variability, looking at different ways to measure, communicate, and understand variability in the context of statistical problems.


>**Essential Concepts in Geometry, Trigonometry, and Measurement.**    Candidates demonstrate and apply understandings of major mathematics concepts, procedures, knowledge, and applications of geometry including using visual representations for numerical functions and relations, data and statistics, and networks, to provide a lens for solving problems in the physical world. Essential Concepts in Geometry, Trigonometry, and Measurement include transformations; geometric arguments; reasoning and proof; applied problems; and non-Euclidean geometries.

 
Part III on Geometry looks at the subject of Geometry from the perspectives of constructional, transformational, analytic, and algebraic. Each perspective helps us to better understand the essential concepts in geometry, trigonometry, and measurement, along with the interactions between geometry, algebra, functions, and number systems. 

### Exercises

1. Reflect on your K-12 mathematics education. Based on the brief history described, what documents were being used to guide your curriculum? 

2. Those who study mathematics curriculum describe the changing focus of what is written in school standards for mathematics as a pendulum. Over time, the pendulum swings between a focus on procedural fluency such as *A Nation at Risk* -@NCEE1983 describes and the more process oriented standards as outlined by NCTM (1989).

    a) How does the Common Core State Standards attempt to unify the two extremes of the pendulum swing?
    b) How does treating the process standards as separate from the content standards provide an opportunity to continue to let the pendulum swing?
    c) Regardless of which documents were currently in vogue while you were in school, do you think your mathematics education was more procedurally focused or more process standard focused? 
    d) Is how you were taught what you hope your own teaching will be like? Why or why not?
    e) K-12 mathematics students often think that mathematics is a set of unrelated procedures that they have to memorize how to do. In contrast, people who do mathematics for a living think mathematics is a conceptual and logical system where everything fits together. As you currently understand them, do you think the Common Core State Standards could support students in moving from thinking about mathematics as a set of unrelated procedures to a more conceptual system? Why or why not?









