# Equality, Order, and Equivalence {#ch:equivalence}


The chapter title may look like just a list, but the notions of equality, order, and equivalence are a critically important trio of related but distinct ideas in mathematics. Together they help us define when two things are the "same." Instruction in these ideas starts on almost the first day of school when students first learn how to count. Students continue study and expand these ideas in their study of mathematics: many of the theorems of mathematics labeled as "Fundamental" are statements of equivalence. These include the Fundamental Theorem of Arithmetic that every integer can be uniquely factored into primes, the Fundamental Theorem of Calculus describing the relationship between antiderivatives and infinite sums, and isomorphism theorems of abstract algebra and homeomorphisms of topology.


## Partitions and Equivalence Relations

```{r, echo=FALSE, fig.align='center', out.width='50%', fig.cap='Sorting Candy'}
knitr::include_graphics(
  'images/m-and-m-1308543_1920.jpg',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```


As a way to better understand the world, humans tend to sort information into categories or orderings. For instance, when some people see a pile of various types of M&M's<sup>TM</sup>  they have the urge to sort them in various ways. Some people sort the candies according to their color. Others may sort them according to their type (plain, peanut, etc.). This process of sorting is a type of "partitioning" the items into distinct categories. With this categorization we want to make sure that each candy is sorted into one, and only one, type. This type of sorting can be formalized with the following definition.

### Partitions

```{definition}
	A **partition** of a set $A$ is a finite or infinite collection of non-empty, mutually-disjoint subsets whose union is $A$. \index{Partition}
```

```{r, echo=FALSE, fig.align='center', out.width='50%', fig.cap='Partition of Sets'}
knitr::include_graphics(
  'tikz/Set_partition.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```


Thus, a partition of a set is a collection of subsets of the original set with several key properties. First, the subsets are non-empty, meaning that each subset in the partition contains at least one element. This allows us to restrict the partition to a collection of non-trivial subsets and means partitions have a type of uniqueness. The subsets in the partition are also mutually-disjoint which means that no element of the original set is in more than one of the subsets in the partition. Lastly, since the union of the subsets that form the partition is $A$, we have that every element in $A$ is in one of these subsets. Thus, each component of the definition of a partition is essential to defining a way to split a set into a set of categories that we find useful.

```{example}
If we have a set
 $$A=\left\{ a, 2, \alpha , \mbox{apple} , \mbox{box}, \beta, 1, 3, \gamma , \mbox{cow}, c, b \right\}$$ there are some natural ways to partition the set. Here are three possible partitions.
 
$$
            A_1 = \{a,b,c\}, \:
            A_2 = \{1,2,3\},\: 
            A_3 = \{\alpha, \beta, \gamma\}, \: 
            A_4 = \{\mbox{apple} , \mbox{box}, \mbox{cow}\} 
$$
$$
            A_1 = \{\mbox{apple} , \mbox{box}, \mbox{cow}\}, \:
            A_2 = \{a,b,c,1,2,3,\alpha,\beta,\gamma\}
$$
$$
            A_1 = \{a,1,\alpha,\mbox{apple}\}, \: 
            A_2 = \{b,2,\beta,\mbox{box}\}, \: 
            A_3 = \{c,3,\gamma, \mbox{cow} \}
$$

There are many other ways to partition the set, but we do not have the space to write out all of the possible partitions of this set.
```

```{example}
	 Early in the K-12 mathematics curriculum we ask students to determine which integers are even and which are odd. In this situation, the set $A$ is the set of integers ($\mathbb{Z}$), while the subsets that form the partition are the subset of even numbers, ($2\mathbb{Z}$), and odd numbers, ($2\mathbb{Z}+1$).
```

```{example}
If we let the set $B$ be defined by the following list of elements,
 $$B = \begin{Bmatrix} 
\mbox{ jar},  & \mbox{ jellybean} ,  & \mbox{ applesauce}, & \mbox{ ape}, & \mbox{ bear}, \\
\mbox{ ball}, & \mbox{ beanbag}, & \mbox{ bag}, & \mbox{ arm} , & \mbox{ skate} , \\
\mbox{ bike}, & \mbox{ rock}, & \mbox{ rowboat},   &  \mbox{ jigsaw}, & \mbox{ saw}
 \end{Bmatrix}$$
take some time to create various partitions of the set $B$, verifying the various properties in the definition of a partition.
```


In this text, we denote the partition $\mathcal{P}$ of a set $A$ as the union of subsets of $A$ over some indexing set, $I$. For example, if the partition is formed using two subsets of $A$, then the indexing set could be $I=\{1,2\}$. The indexing set can also be infinite. For instance, 
$$\mathcal{P} = \left\{ A_n  \right\}_{n\in \mathbb{Z}}, \: \mbox{ where } A_n=\left\{x\in \mathbb{R}\middle \vert n\leq x < n+1  \right\}$$ is a partition of the real numbers with an infinite number of sets.

It is important to note that there are usually many ways to partition a set, but the definition of partition does not impose any constraints on the uniformity or order of the subsets, only that the way we choose our subsets meets the three conditions described previously. 

With our previous discussion and examples in mind, we can translate our verbal definition of a partition into a symbolic one. In particular: if $\mathcal{P}= \left\{ U_i\right\}_{i\in I}$ is a partition of a set $A$, where $I$ is some indexing set that is either finite or infinite, we have the following properties
$$U_i \bigcap U_j = \emptyset \quad \mbox{if } i\neq j, \quad \bigcup_{i\in I} U_i = A \mbox{, and} \quad U_i \neq \emptyset \mbox{ for all } i \in I.$$


### Relations
Up until this point we have focused on building a framework for understanding the notion of partition. In order to understand the main topics of this chapter, we also need to develop the idea of a relation. While we give the general definition of a relation and a few examples, we will focus on those relations that arise from partitions in this section. We will explore other types of relations in later sections.

```{definition}
Let $A$ and $B$ be sets. We define a relation $R$ from $A$ to $B$ as a subset of $A\times B$. We say that for elements $a\in A$ and $b\in B$ that $a$ is related to $b$, ($aRb$), if and only if $(a,b)\in R$. If the two sets are the same set, $A$, then we say that $R$ is a relation on $A$.
```

Here are some examples of relations in the K-12 curriculum.


+ Let $A$ and $B$ be the set of integers. We could say that $$R=\{(a,b)\in A\times B \vert a<b\},$$ which defines the 'less than' relation of $aRb \leftrightarrow a<b$.

+ Let $A$ be the real numbers and $B$ be the non-negative real numbers. We can let $$R=\{(x,y)\in A\times B \vert y=x^2\}.$$ This is the graph of the function $f(x)=x^2$.
    
+ Let $A$ and $B$ be the set of integers. We can define $$R= \{(m,n) \vert m-n=2k, \: \mbox{ for some integer } k\}.$$ This relation is one where are the even numbers are related to each other and all of the odd numbers are related to each other.



### Equivalence Relations Induced by Partitions
For the remainder of this section we will study a specific type of relation that is created by a partition, $\mathcal{P}$, of a set, $A$. If $\mathcal{P}=\left\{U_i\right\}_{i\in I}$ is a partition of $A$, we can write $A$ as the union of a collection of non-empty, mutually disjoint subsets: $$U_i \cap U_j = \emptyset \mbox{ if } i\neq j, \: \mbox{ for each } i \in I, \: U_i\neq \emptyset, \: \mbox{ and } A=\bigcup_{i\in I} U_i.$$ 

Consider one of the individual subsets in the partition, say $U_i$. Then the elements of $U_i$, by virtue of the fact they are in the same subset, have a relationship to each other. This idea extends to the whole partition. That is, the partition induces a relation on $A$ in the following way. For each $a,b\in A$ we say that $aRb$ if, and only if, there is a set $U_i$ in the partition $P$ such that both $a$ and $b$ are in $U_i$. 

The relations we create through a partition behave in ways that are of general interest in the study of mathematics. 

If we choose a generic element $a$ in $A$, the property that $\bigcup_{i\in I} U_i = A$ tells us that there exists $i\in I$ such that $a \in U_i$. So for every element $a\in A$, we have that $aRa$. We will call this property of the relation the **reflexive property**.

For the second property, if we have $a,b\in A$ such that $aRb$, then there must be a $U_i\in P$ with $a,b\in U_i$. This also means that $bRa$. When you have the property that $aRb$ implies that $bRa$, we say that the relation has the **symmetric property**. 

The third property of relations that we define derives from the property of partitions that
$$U_i \bigcap U_j = \emptyset \Leftrightarrow i\neq j.$$ If $a,b,c\in A$ such that $aRb$ and $bRc$, then we have that there are sets $U_i$ and $U_j$ in the partition such that $a$ and $b$ are both in $U_i$ and $b$ and $c$ are both in $U_j$. Since $b$ is in both $U_i$ and $U_j$, $U_i \cap U_j \neq \emptyset$ and so $i=j$. This would then imply that $aRc$. This property that the combination of $aRb$ and $bRc$ implies that $aRc$ is called the **transitive property** of relations. 

Not all relations have these three properties of reflexive, symmetric, and transitive. Those relations that have all three properties are called **equivalence relations**.

We have therefore proven that a partition of a set induces an equivalence relation on that set. 


 
### Partitions Induced by Equivalence Relations

The concept of equivalence relation is one that transcends much of mathematics and we often consider things as being the "same" with respect to some property if there is an equivalence relation under which they are related. One location that this appears in the elementary school curriculum is in regards to rational numbers. We think of $\frac{1}{2}$ and $\frac{2}{4}$ as being the same rational number, but written in a different way. Along these lines we say that two rational expressions $\frac{a}{b}$ and $\frac{c}{d}$ are equivalent if and only if $ad=bc$. This creates and equivalence relation between these rational expressions that are used to create the rational numbers.

Similarly, if a relation, $R$, on a set, $A$, is an equivalence relation, then one can generate an induced partition, $\mathcal{P}$, as follows. For each $a\in A$ we define the set $$[a]_R := \left\{ b\in A \middle \vert aRb\right\}.$$  We call this set, $[a]_R$, the equivalency class of $a$ under the relation $R$. 

Since $R$ is an equivalence relation, it is reflexive and so $a \in [a]_R$ causing $[a]_R \neq \emptyset$. Since $a\in [a]_R$, we have that $$\bigcup_{a\in A} [a]_R = A.$$ Thus to show that $\mathcal{P} = \left\{ [a]_R \middle \vert a \in A\right\}$ is a partition of $A$, we need to show that for $a,b\in A$ that $[a]_R=[b]_R$ or $[a]_R\bigcap [b]_R = \emptyset$. 

 If $[a]_R \bigcap [b]_R \neq \emptyset$, then we need to prove that two sets to be equal.
    
Let us assume that there exists a $c\in [a]_R \bigcap [b]_R$.
    
Let $d\in [a]_R$. So $a R d$. $c$ is also in $[a]_R$ and so $aRc$. Since $R$ is symmetric and $aRd$, $dRa$. We now have that $dR c$ using transitive property. $c$ is also in $[b]_R$ and so $bRc$. Using the symmetric property, we have $cR b$. Combining this with $dRc$ and using the transitive property we have $dR b$. Using the symmetric property we have $bR d$. So $d\in [b]_R$. Therefore, $[a]_R \subseteq [b]_R$.
    
Using very similar arguments we can prove that $[b]_R \subseteq [a]_R$.
    
So if the intersection is nonempty, the two sets are the same set.


We have thus proven the following theorem showing the connection between equivalence classes and partitions.

```{theorem}
If $R$ is an equivalence relation on a set $A$, then $$P = \left\{ [a]_R \middle \vert a \in A\right\}$$ is a partition of $A$.
    
Similarly, if $P= \left\{ U_i \middle \vert i \in I\right\}$ is a partition on a set $A$, then the relation $R$ defined by 
$$aRb \Leftrightarrow \mbox{ there exists } i \in I \mbox{ such that } a,b \in U_i$$ is an equivalence relation.
```


### Exercises

1. Define a relation $\sim$ on $\mathbb{R}\times (\mathbb{R}-\{0\})$ by 
    $$(x,y) \sim (z,w) \Leftrightarrow xw=yz.$$
    a. Show that this relation is an equivalence relation.
    a. Give a description of the partition of $\mathbb{R}\times (\mathbb{R}-\{0\})$ induced by this equivalence relation.
    a. Why do we not allow $0$ in the second part of the ordered pairs?
        
1. Determine which of the following collections of sets form a partition of $\mathbb{R}\times \mathbb{R}$. For those that do form a partition, define the equivalence relation.
    a. For $c\in \mathbb{R}$, let $h_c= \{ (x,y)\in \mathbb{R}^2 \vert y=c\}$, and let $\mathcal{C} = \{h_c\}_{c\in \mathbb{R}}$.
    a. For $m\in \mathbb{R}$, let $s_m=\{ (x,y)\in \mathbb{R}^2 \vert y=mx\}$, and let $\mathcal{D}= \{s_m\}_{m\in \mathbb{R}}$.
    a. For $r\in [0,\infty)$, let $C_r=\{ (x,y) \in \mathbb{R}^2 \vert x^2+y^2=r^2\}$, and let $\mathcal{E}=\{C_r\}_{r\in [0,\infty)}$.
    
1. (For students who have knowledge of linear algebra)  Let $M_{2,2}$ be the set of $2 \times 2$ matrices with real coefficients.  For each of the following, prove that it is an equivalence relation, or show that one of the properties of equivalence relations does not hold.

    a. For matrices $A$ and $B$ in $M_{2,2}$, let $A \sim B$ if and only if $\mathrm{det}(A)=\mathrm{det}(B)$.
    a. For matrices $A$ and $B$ in $M_{2,2}$, let $A \sim B$ if and only if $\mathrm{tr}(A)=\mathrm{tr}(B)$.
    a. For matrices $A$ and $B$ in $M_{2,2}$, let $A \sim B$ if and only if the eigenvalues of $A$ are the same as the eigenvalues of $B$.
        
        




## Ordered Sets and Relations

In the previous section we defined a relation $R$ from $A$ to $B$ as a subset of $A\times B$, and that $a$ is related to $b$, ($aRb$), if and only if $(a,b)\in R$. If the two sets are the same set, $A$, then we say that $R$ is a relation on $A$.

We also defined the following three properties of a relation, $R$, on a set $A$:

+ **Reflexive.** The relation $R$ is reflexive if $aRa$ for all $a\in A$.
+ **Symmetric.** The relation $R$ is symmetric if $aRb$ implies that $bRa$ for all $a,b\in A$.
+ **Transitive.** The relation $R$ is transitive if $aRb$ and $bRc$ imply that $aRc$ for $a,b,c\in A$.


A relation is an equivalence relation if it satisfies all three of these properties. Another type of relation is one that creates an order, or ranking, of the set. A common mathematical topic in this regards are the concepts of "less than" and "less than or equal to".

We will first explore the relation of "less than". Since a number cannot be less than itself we see that the relation is not reflexive. We also know that $a<b$ and $b<a$ cannot both be true and so the relation is not symmetric. However, if $a<b$ and $b<c$ we do have that $a<c$. So this relation of "less than" is transitive. Thus the familiar relation of "less than" is one which is transitive but not reflexive or symmetric.

When we expand the relation to that of "less than or equal to", we add that $a\leq a$, resulting in the relation being reflexive. So this is a common relation that is reflexive and transitive, but not symmetric. For example, $1\leq 2$, but $2$ is not less than or equal to $1$.



```{definition}
We define an **order** on a set, $A$, to be a relation, denoted by $<$, with the following properties.

1. If $x$ and $y$ are two elements of $A$, then one, and only one, of the statements $$x<y, \quad x=y, \quad, y<x$$ is true.
2. If $x,y,z \in A$, and if $x<y$ and $y<z$, then $x<z$.

If a set has an order defined on the set, then we say that it is an **ordered set**.
```

Such an order forms a relation that is transitive, but not symmetric or reflexive. As a convenient notation, we will write $x\leq y$ as the negation of $y<x$. We then see that it is equivalent to the statement that ($x<y$ or $x=y$).

```{definition, well-ordering, name="Well-Ordering Property"}
A set, $A$, together with an order, $<$, satisfies the well-ordering property if every non-empty subset of $A$ has a least element. Equivalently, if $S\subseteq A$ with $S\neq \emptyset$, then there exists an element $s_0\in S$ such that $s_0 \leq s$ for all $s\in S$. 
```

We will see in Chapter \@ref(ch:number) that the natural numbers, $\{0,1,2,3,\ldots\}$, satisfy the well-ordering property, but the real numbers do not. For instance, $\{x\in \mathbb{R}\vert x>1\}$ does not have a least element. This means that in some way, a set needs to be discrete in order to satisfy a well-ordering property.

For sets that are not discrete, we would like to have a similar property.


```{definition}
Let $A$ be an ordered set and let $S\subseteq A$. If there exists an $\alpha \in A$ such that $x\leq \alpha$ for all $x\in S$, then we say that $S$ is **bounded above** and that such an $\alpha$ is an **upper bound** of $S$.
```

 We can similarly define **bounded below** and **lower bound** using opposite inequalities.
 
```{definition, name="Least-Upper-Bound Property"}
An ordered set $A$ is said to have the **least-upper-bound property** if for every non-empty subset $S\subseteq A$ that is bounded above, there exists an element $s_0\in S$ such that $s_0$ is an upper bound of $S$ and if $\alpha$ is an upper bound of $S$, then $s_0\leq \alpha$. Such an $s_0$ is called the **least-upper-bound** of $S$.
```

We will see in Chapter \@ref(ch:number) that the real numbers satisfy the least-upper-bound property, while the rational numbers do not.



### Exercises

1. For each of the following relations, determine if the relation is reflexive, symmetric, and/or transitive.
    a. Let $S$ be a relation defined on $\mathbb{R}$ by $$xSy \Leftrightarrow x^2=y^2.$$
    a. Let $T$ be a relation defined on $\mathbb{R}$ by $$xTy \Leftrightarrow x<y^2.$$
    a. Let $U$ be a relation defined on $\mathbb{Z}$ by $$mTn \Leftrightarrow mn>0.$$
    a. Let $V$ be a relation defined on $\mathbb{R}^2$ by $$ (x,y)V(w,z) \Leftrightarrow x^2+y^2 \leq w^2+z^2.$$
        
1. Let $S=\{a,b\}$. List out all of the possible relations on $S$ and determine which relations are reflexive, symmetric, and/or transitive.
        




## Equality and Equivalence in the K-12 Curriculum

We can see from the discussion in the previous two sections that the concept of two things being equivalent mathematically is a very difficult and abstract concept. It is no surprise then that students of all grade levels have many different conceptions and misconceptions revolving around this concept of equality and equivalence.

Since all of the number systems that arise in the elementary and early secondary curriculum have an inherent order described by the $\leq$ and $<$ symbols, the concept of equality is often described in terms of a balanced scale. Another conception around the equal sign is that two sets of objects have the same number of items, determined by counting the number of objects. This is likely since the concept of equality arises in the first grade where students are only working with whole numbers.

```{r, echo=FALSE, fig.align='center', out.width='35%', fig.cap='Sample problems', label="equality-operation"}
knitr::include_graphics(
  'tikz/sample_addition_problem.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```


One of the common misconceptions surrounding the equal sign is that the symbol implies an action of doing some type of computation. For instance, students usually see the equal sign in the context of the type of problems seen in Figure \@ref{fig:equality-operation}.

These type of problems promote an operational understanding of the equal sign, rather than a relational or substitutional understanding that is essential for later work [@Cognition, pp. 145-150]. Because secondary mathematics students maintain many misconceptions about the concept of equality and the meaning of the equal sign, we will take some time to reflect on some key content standards from first grade that have a very strong influence on student success of learning mathematics in the middle school.

>#### Related Content Standards {-}
>- (1.OA.7) Understand the meaning of the equal sign, and determine if equations involving addition and subtraction are true or false. For example, which of the following equations are true and which are false? $6 = 6$, $7 = 8 - 1$, $5 + 2 = 2 + 5$, $4 + 1 = 5 + 2$.
>- (1.OA.8) Determine the unknown whole number in an addition or subtraction equation relating three whole numbers. For example, determine the unknown number that makes the equation true in each of the equations $8 + ? = 11$, $5 = ? - 3$, $6 + 6 = ?$.


```{r, echo=FALSE, fig.align='center', out.width='40%', fig.cap='Pan balances', label="pan-balance"}
knitr::include_graphics(
  'images/balance-scales.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

We should notice that in standard (1.OA.D.8) that the unknown should not always be at the end of a string of mathematical symbols and is based on educational research that compares students from different cultures and their understanding of the equal sign [@Cognition, pp. 148-150]. Instead it should be at many different locations in order to make sure that students do not internalize the misconception that the equal sign means to compute. They should instead develop the concept of the equal sign representing two things that have the same value in some way. It is this understanding of two things looking different but representing the same underlying value that causes the equal sign to represent an equivalence. 


Another method that may possibly contribute to a more relational understanding of the equal sign is the introduction of the equal sign alongside the greater than ($>$) and less than ($<$) symbols to emphasize the role of the equal sign in the comparison of the size of numbers. This comparison concept can be further reinforced with a pan balance (See Figure \@ref(fig:pan-balance)).



This concept of a pan-balance can give students a concrete idea to attach to the meaning of the equal and inequality signs. However, like all good analogies, the pan balance is challenging to use with negative numbers and breaks down if you are comparing numbers in an unordered number system such as the complex numbers.

 



### Exercises
A student is given the following task.

>- Sophia goes to the store with $\$5.00$ and buys two packs of gum for $\$1.20$. She then gives $\$2.00$ to her friend to buy a drink. How much money does she have left? 

1. The student then writes the following on their paper.
$$5.00-1.20=3.80-2.00=1.80, \mbox{ so Sophia has } \$1.80 \mbox{ left.}$$

    a. What ideas from this section regarding the equal sign and equivalence would you use to provide feedback to this student?
    
    a. What are other points of feedback that you would provide to this student?
    
1. If a student asks you what the equal sign means in the context of $\frac{1}{2}=\frac{2}{4}$ since those symbols are not the same.
    
1. What are the meanings of the equal signs in the task?
>Evaluate the function $f(x)=3x-4$ for $x=2$. 


## Expressions, Equations, and Inequalities

A **numerical expression** is a meaningful string of numbers, operation symbols, and/or grouping symbols. In early elementary grades, examples of numerical expressions include $$2+6+4, \quad 13-3+2, \quad \mbox{ or } \quad (4+6)+3.$$ In upper elementary grades, students are expected to incorporate rational numbers and the operations of multiplication, division, and exponents in order to be comfortable working with such expressions as 
$$3 \times (10+4), \quad \frac{2}{3} \times (12+8), \quad \mbox{ or } \quad  2^3 + \frac{3+2}{7}.$$

In the transition from elementary school to secondary schools, students learn to add **variables**, symbols or letters that stand for any number within a specified range of numbers, to these expressions and thus work with **algebraic expressions** such as
$$100-10\cdot (P-15)^3,  \quad 3xy + 5(x-4)^2-7y^2, \quad 6\cdot (4x+3y), \quad \mbox{ or }\quad  \frac{3x+2}{2x-4}.$$


>#### Related Content Standards {-}
>- (6.EE.2) Write, read, and evaluate expressions in which letters stand for numbers.
      a. Write expressions that record operations with numbers and with letters standing for numbers.
  	  b. Identify parts of an expression using mathematical terms (sum, term, product, factor, quotient, coefficient); view one or more parts of an expression as a single entity. 
	    c. Evaluate expressions at specific values of their variables. Include expressions that arise from formulas used in real-world problems. Perform arithmetic operations, including those involving whole-number exponents, in the conventional order when there are no parentheses to specify a particular order (Order of Operations).



In more advanced courses in secondary school mathematics, algebraic expressions can also include functions giving rise to expressions such as 
$$f(x-h)+k, \quad \sin(3x)-2, \quad \mbox{ or } \quad  \ln(x+3).$$ 

The equivalence of numerical expressions is a key concept in the early elementary grades where students learn to compose and decompose numbers in different ways as they add and subtract numbers using different algorithms. In sixth grade, students are asked to determine the equivalence of two mathematical expressions. This definition that two algebraic expressions are equivalent if they generate the same number regardless of which number is substituted for the variable is one of the first places where students are pushed to move to the abstract realm of equivalence relations. We can verify that this definition of equivalent expressions is an equivalence relation using the fact that equality on the real numbers is an equivalence relation.


>#### Related Content Standards {-}
>- (6.EE.3) Apply the properties of operations to generate equivalent expressions. %{\em For example, apply the distributive property to the expression $3 (2 + x)$ to produce the equivalent expression $6 + 3x$; apply the distributive property to the expression $24x + 18y$ to produce the equivalent expression $6 (4x + 3y)$; apply properties of operations to $y + y + y$ to produce the equivalent expression $3y$.}
>- (6.EE.4) Identify when two expressions are equivalent (i.e., when the two expressions name the same number regardless of which value is substituted into them).
>- (6.EE.5) Understand solving an equation or inequality as a process of answering a question: which values from a specified set, if any, make the equation or inequality true? Use substitution to determine whether a given number in a specified set makes an equation or inequality true.


While substituting in values to determine if two expressions are equivalent is the definition of the equivalence, it is almost always impossible to substitute in all possible values for a variable, as there are infinitely many possibilities in most situations. Instead, we determine the equivalency of two expressions by applying various properties of the operations. For example, the distributive property shows us that $3(x+y)$ is equivalent to $3x+3y$. However, students have to be careful to consider the set of possible numbers for an expression when determining the equivalence of $\sqrt{x^2}$ and $x$, or $\frac{3x^2}{x}$ and $3x$.

An **equation** is a statement that a number or expression is equivalent to a different number or expression. An equation that is true for all values of the variable is called an **identity**. The associative, commutative, and distributive properties of the number systems 
\begin{align*}
    a+(b+c) &= (a+b)+c \\
    a+b &= b+a \\
    a \times (b+c) &= (a\times b)+(a\times c)
\end{align*}
are all common identities. Some other common identities in this regard are the trigonometric identity
\begin{align*}
    \cos^2(\theta) + \sin^2(\theta) &= 1  \\
    \sin(\alpha \pm \beta) &= \sin(\alpha)\cos(\beta)\pm \cos(\alpha)\sin(\beta) \\
    \cos(2\theta) &= \cos^2(\theta)-\sin^2(\theta)
\end{align*}
that are used to simplify trigonometric expressions.

An equation that describes how multiple quantities vary together is usually called a **formula**. These can include the area formula for rectangles defined by $\mbox{Area}=\mbox{width} \times \mbox{height}$ or Boyle's law where pressure ($P$) is inversely proportional to volume ($V$), i.e.  $PV =k$ for some constant $k$.

While the identity equations hold true irrespective of the values of the variables, it is often the case that equations hold for only some of the possible values for the variables, or possibly even none. We can then **solve** such equations by determining which values, called **solutions to the equation**, for the variables cause the equation to be a true statement.

Similar to equations, **inequalities** are statements that a number or expression is related to a different number or expression in a certain order relation. Some of these inequalities are identities or formulas, such as $x\leq |x|, x\in\mathbb{R}$, while some have a solution set, such as $x^2+y^2<1$.


```{definition}
    Two equations (inequalities) are equivalent if they are true statements for the same values of the variables. 
```

In the process of finding solutions to equations or inequalities, it is almost always helpful to rewrite an equation or inequality as an equivalent equation or inequality. For instance in an effort to find the solutions to $$\frac{x^2-10x+21}{3x-12} = \frac{x-5}{x-4}$$
we want to find an expression that is equivalent to the left side of the equation. Such an expression could be $$\frac{(x-3)(x-7)}{3(x-4)},$$ with the equivalency of the expressions verified using the distributive property. Since these two expressions are equivalent, they have the same values for every value of the variable. This means that the original equation is a true statement if, and only if, $$\frac{(x-3)(x-7)}{3(x-4)}=\frac{x-5}{x-4}$$ is a true statement. We also know that if two numbers are the same, then $3$ times each of the numbers are also the same. So the original statement is true if, and only if, the equivalent equation
$$\frac{(x-3)(x-7)}{x-4}=\frac{3(x-5)}{x-4}$$ is true. If $x=4$, then the statement would have no meaning, as division by zero has no meaning. This means that we can eliminate $4$ from the set of possible solutions. With $x\neq 4$ we can multiply both sides of the equation by $(x-4)$ to generate the equivalent equation, $$(x-3)(x-7)=3(x-5).$$ We can use the distributive property again to create an equivalent equation of $$x^2-10x+21=3x-15.$$ Since we can also add the same expression, $-3x+15$, to both sides of the equation and create an equivalent equation, we see that the original equation is equivalent to $$x^2-13x+36=0$$ as long as $x\neq 4$. Using the distributive property, we see that this equation is equivalent to $$(x-4)(x-9)=0.$$ So the original equation is a true statement if, and only if, $(x-4)(x-9)=0$ is a true statement and $x\neq 4$. This only occurs when the variable, $x$, is equal to $9$.

Through this process we can see how equivalence of equations is essential in the process of finding solutions to equations. This means that as students make the transition from elementary to secondary, it is important for teachers to be understanding of the challenge of thinking about equivalent expressions and to be explicit about this relationship.





### Exercises

1. Most of this project will be in the context of algebraic reasoning and these questions should be answered with the expectations of your target student population.
    a. What is the difference between an expression and equation?
    a. Why can we not solve an expression?
    a. What does it mean to "solve an equation"?
    a. What makes two expressions equivalent to each other?
    a. What is the difference between how the equality sign is used when solving an equation versus when working with expressions?

    
1. Prove that the definition of equivalent equations forms an equivalency relation.
















