# Number Systems {#ch:number}

K-12 mathematics is rich in the study of number systems, with particular focus on the Natural Numbers, Integers, Rational Numbers, Real Numbers, and Complex Numbers. In this chapter we use the properties of set theory and equivalence classes discussed in Chapters \@ref(ch:sets) and \@ref(ch:equivalence) to construct these numbers systems. We follow this discussion with a study of some of the properties of number systems, connecting them to the K-12 curriculum. Lastly, we examine different representations of these number systems helpful for developing numerical fluency and conceptual understanding of the numerical operations.

This process of constructing the number systems is very abstract and challenging for those who have not previously studied these constructions. The process is included because it helps teachers to better understand the abstract nature of the number systems and some of the challenges of their students. While working with many of these number systems has become second nature for math teachers, the initial learning by students is much harder than expected for new teachers and so a deeper knowledge of the properties of the number system helps in this teaching situation. 



$$\mathbb{N} \subseteq \mathbb{Z} \subseteq \mathbb{Q} \subseteq \mathbb{R} \subseteq \mathbb{C}$$
As we are working with the various number systems, we will use varying notation for our operations. With all of the constructions, subtraction and division are not defined. We instead define the additive inverse and multiplicative inverse and so we can think of subtraction $a-b$ as $a+(-b)$ and division $a\div b$ as $a \times \frac{1}{b}$. We will also use various symbols to represent multiplication so that $a\times b$, $a\cdot b$, and $ab$ all represent the same thing.


## Natural Numbers and Integers {#sec:Integers}

As part of the effort to axiomatize mathematics in the late 19^th^ and early 20^th^ centuries, individuals such as Richard Dedekind (1831-1916), Georg Cantor (1845-1918), Giuseppe Peano (1858-1932), Alfred Whitehead (1861-1947), David Hilbert (1862-1943), Ernst Zermelo (1871-1953), Bertrand Russell (1872-1970), and Abraham Fraenkel (1891-1965) followed a program to start with the most basic of assumptions and build a solid foundation for the field of mathematics. One of the greatest works in this program is the *Principia Mathematica* by Whitehead and Russell [-@Principia1; -@Principia2; -@Principia3] that built such a foundation on logic and set theory that it took until page 86 of the second volume to prove that $1+1=2$.  While we will not go to the full extent of abstraction, we will build up the natural numbers and the integers from the axioms of Zermelo-Fraenkel set theory, together with the axiom of choice (ZFC axioms).

### Natural Numbers
We will study the concepts of counting and related ideas of the natural numbers in Chapter \@ref(ch:function), but initially we will use John von Neumann's [-@vonNeumann23] definition for the natural numbers which are defined inductively by defining the symbols 
$$0:=\emptyset, \quad 1:=\{\emptyset\}, \quad 2:=\{\emptyset, \{\emptyset\}\},$$ and so on. So for any natural number $n$, the next natural number is defined inductively as $$n+1:=S(n) := n\cup \{n\},$$ where you think of $S(n)$ as the successor of $n$. 

This definition of the natural numbers is highly dependent upon the axiom of infinity from the ZFC axioms. In the formal language of the ZFC axioms, the axiom of infinity is stated as 
$$\exists I ( \emptyset\in I \wedge \forall x\in I((x\cup \{x\})\in I).$$ Or equivalently, there is a set $I$ that contains the empty set as an element and that if $x\in I$, then $x\cup \{x\}$ is also in $I$. The set $I$ is a larger set than $\mathbb{N}$, but allows for the existence of the natural numbers.

With our definition of the natural numbers, we also have that $S(n)\neq n$, $$S(n)=S(m) \Leftrightarrow n=m,$$ and we can create an order on the set of symbols $\mathbb{N} = \{0, 1, 2, 3, \ldots \}$ by defining that
$$n \leq m \Longleftrightarrow n \subseteq m.$$ One can prove that this set of symbols satisfies the well-ordering property (\@ref(def:well_ordering)) using the ZFC axioms, which enables us to prove the principle of mathematical induction creating the mathematical machinery necessary for the study of the natural numbers. 

```{theorem, name="Principle of Mathematical Induction"}
Suppose that $S$ is a subset of $\mathbb{N}$ such that $0\in S$, and for all $n\in \mathbb{N}$, $n\in S$ implies that $S(n) \in S$.

Then $S$ must be the set of natural numbers.
```

We will leave the detailed proof for a text on symbolic logic. However, the application of this theorem is that if we have a mathematical statement that varies along the natural numbers, $P(n)$. If the statement $P(0)$ is true, and if $P(k)$ being true implies that $P(k+1)$ is true, then $P(n)$ must be true for all natural numbers.

We inductively define the operation of addition on these symbols so that for any integer $a$, 
\begin{align*}
    a+0 &:=a, \\
    a+1 := a+S(0) &= S(a+0), \\
    a+2:=a+S(1) &= S(a+1), \\
    \vdots \\
    a+S(b) &:= S(a+b)
\end{align*}
We similarly define multiplication inductively by
$$a\times 0 := 0, \quad a\times 1 := (a \times 0)+a = a, \quad \mbox{ and } \quad a\times S(b) := (a\times b)+a.$$
With these operations we have the following properties of the natural numbers.

```{theorem}
Let $a$, $b$, and $c$ be natural numbers. Then

+ $a+b$ and $a\times b$ are natural numbers (closure)
+ $a+(b+c)=(a+b)+c$ and $a\times (b\times c)=(a\times b)\times c$ (associative property)
+ $a+b=b+a$ and $a\times b=b\times a$ (commutative property)
+ $a+0=a$ and $a\times 1 =a$ (additive and multiplicative identities)
+ $a \times (b+c)= (a\times b) + (a\times c)$ (distributive property)
+ If $a\times b=0$, then $a=0$ or $b=0$ (or both). (no non-zero divisors)

```


In order to get a flavor for the techniques involved in proving these properties, we will prove the associative property of addition for the natural numbers. The proofs of the remainder of the statements will be left to the reader to work through.
```{proof}    
Let $a,b \in \mathbb{N}$. Then, for each $c\in \mathbb{N}$, we let $P(c)$ be the statement that 
$$a+(b+c) = (a+b) + c.$$
  
If $c=0$, then $b+c=b$ and $(a+b)+c=a+b$, and so 
$$a+(b+c)=a+b= (a+b)+c.$$ This implies that $P(0)$ is true.

Assume by that for some value of $c'\in \mathbb{N}$ that $P(c')$ is true, i.e. that $a+(b+c')=(a+b)+c'$ ($c'=0$ is true and so such a $c'$ exists). Then for $c=S(c')$, 
\begin{align*}
    a+(b+c) &= a+(b+S(c'))=a+S(b+c') \\
    &= S(a+(b+c')) = S((a+b)+c') \: \mbox{ (by the induction hypothesis)}\\
    &= (a+b)+S(c') = (a+b)+c.
\end{align*}

Therefore, P(c) is true. and by the principle of mathematical induction  we have that $a+(b+c)=(a+b)+c$ for all values of $a,b,c\in \mathbb{N}$.
```

We previously defined the order on the natural numbers using set inclusions. This order can be rewritten using operations by noting that $m<n$ if, and only if, $n=m+k$ for some natural number $k$. We will explore how this order interacts with the operations of addition and multiplication.

```{lemma, order-addition-naturals}
Let $m,n,k\in \mathbb{N}$. Then $$ (1) \: (m=n) \Leftrightarrow (m+k=n+k) \quad \mbox{ and } \quad (2) \: (m<n) \Leftrightarrow (m+k<n+k).$$
```

```{proof}
We will prove part $(1)$ using an induction argument on $k$ using the statement that given $m,n\in \mathbb{N}$, $P(k)$ is the statement that $(m=n) \Leftrightarrow (m+k=n+k)$. Since $0$ is the additive identity, we know that $P(0)$ is true. We also have from the definition of $S(n)$ that $P(1)$ is true. If we assume for some $l\in \mathbb{N}$ that $P(l)$ is true, then $m+(l+1) = n+(l+1)$ is equivalent to $(m+l)+1 = (n+l)+1$, and since $P(1)$ is true this is equivalent to $(m+l)=(n+l)$. So $P(l)$ being true implies that $P(l+1)$ is true. Therefore, $P(k)$ is true for all $k\in \mathbb{N}$.
    
In order to prove part $(2)$, we note that
\begin{align*}
    (m<n) &\Leftrightarrow (\mbox{There exists } j\in \mathbb{N} \mbox{ such that } (n=m+j)) \\
    &\Leftrightarrow (\mbox{There exists } j\in \mathbb{N} \mbox{ such that } ((n+k)=(m+j)+k)) \\
    &\Leftrightarrow (\mbox{There exists } j\in \mathbb{N} \mbox{ such that } ((n+k)=(m+k)+j)) \\
    &\Leftrightarrow (m+k<n+k).        
\end{align*}
```

```{lemma, order-multiplication-naturals}
Let $m,n,k\in \mathbb{N}$. Then $$ (1) \: (m=n) \Leftrightarrow (m\times k=n\times k) \quad \mbox{ and } \quad (2) \: (m<n) \Leftrightarrow (m\times k<n\times k), \mbox{ for } k>0.$$
```

The proof of this lemma is very similar to the previous lemma and is left as an exercise.


### Integers

Now that we have defined the set of the natural numbers ($\mathbb{N}$) and the operations on $\mathbb{N}$ of addition and multiplication, we would like to have additive inverses. However, we do not assume that the set of additive inverses exist, but instead construct the integers from the natural numbers.

Subsection \@ref(subsec:cartesian) gives the existence of the set $$\mathbb{N}\times \mathbb{N} = \{ (a,b)  \vert a,b \in \mathbb{N}\}.$$ On this set of ordered pairs, we will define an equivalence relation that 
$$(a,b)\sim (c,d) \quad \Leftrightarrow \quad  a+d=b+c.$$

Since this is an equivalence relation we can define $\mathbb{Z}$ to be the set of equivalence classes under this relation. We will denote by $[(a,b)]$ the equivalence class containing $(a,b)$.

We can now define the operations of addition and multiplication on this set of equivalence classes.
$$[(a,b)]+[(c,d)]:= [(a+c,b+d)] \quad \mbox{and} \quad [(a,b)]\times [(c,d)] := [(ac+bd,ad+bc)]$$

We need to make sure that these definitions are well-defined in that the operations do not depend on which element of the equivalence class is chosen for the representation. Assume that $(a,b)$ and $(a',b')$ are both elements of $[(a,b)]$ and that $(c,d)$ and $(c',d')$ are both elements of $[(c,d)]$. This is equivalent to the statements that $$a+b'=b+a' \quad \mbox{ and } c+d'=d+c'.$$ So, using properties of equivalent equations involving natural numbers we can add the two equations together and see that $$(a+b') + (c+d') = (b+a')+(d+c').$$ Using the associative and commutative properties of addition for the natural numbers we see that this equation is equivalent to $$(a+c) + (b'+d') = (b+d) + (a'+c')$$ and so $[(a+c,b+d)]$ is equivalent to $[(a'+c',b'+d')]$, proving that addition is well-defined. The proof that multiplication is well-defined is left as an exercise.

We can also put an order on this set by defining that $$[(a,b)] < [(c,d)] \Leftrightarrow a+d<b+c.$$


We can think of the natural numbers, $\mathbb{N}$, as the subset of $\mathbb{N}\times \mathbb{N}$ of the form
$$\mathbb{N} := \mathbb{N} \times \{0\} = \{ [(a,0)] \vert a\in \mathbb{N}\},$$   since $[(a,0)]+[(b,0)] = [(a+b,0)]$ and $[(a,0)] \times [(b,0)] = [(ab,0)]$.

With this identification of the natural numbers with $\mathbb{N}\times \{0\}$, where $a=[(a,0)]$, we are able to see that the negative integers can be associated with $\{0\} \times \mathbb{N}$, since $[(a,0)]+[(0,a)] = [(a,a)] = [(0,0)]$. We then label $[(0,a)]$ as $ - [(a,0)]$, and usually write it as $-a$. We can then identify every element of $\mathbb{N}\times \mathbb{N}$ with an integer,
$$[(a,b)] \leftrightarrow a+(-b).$$



Thus we have constructed a new set $\mathbb{Z}$ with the operations of addition and multiplication that satisfy the properties for $a,b,c\in \mathbb{Z}$ in Table \@ref(tab:intprops).


```{r intprops, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
addition <- c("$a+b$ is an integer", "$a+(b+c)  = (a+b)+c$", "$a+b=b+a$", "$a+0=a$", "$a \\times (b+c) = (a\\times b) + (a \\times c)$", "$a + (-a) =0$", " If $a,b>0$, then $ab>0$.", "$ab=0$ if, and only if, $a=0$, $b=0$, or both")
multiplication <- c("$a \\times b$ is an integer", "$a \\times (b \\times c) = (a \\times b) \\times c$", "$a\\times b = b\\times a$", "$a \\times 1 = a$", "$(a+b) \\times c = (a\\times c) + (b\\times c)$", "", "If $a,b<0$, then $ab>0$.", "")
property <- c('Closure', 'Associative Property', 'Commutative Property', 'Identities', 'Distributive Property', 'Additive Inverses', '', 'No zero divisors')
intprops <- data.frame(addition, multiplication, property)
names(intprops) <- c('', '', 'Property')
knitr::kable(intprops, format='pandoc', caption = 'Properties of Integers')
```

While the integers do not satisfy the Well-Ordering Property (\@ref(def:well-ordering)), since the negative numbers are a non-empty set without a least element, it does satisfy the least-upper-bound property that any non-empty set that is bounded above has a least upper bound.

The integers do however, have generalizations of Lemmas \@ref(lem:order-addition-naturals) and \@ref(lem:order-multiplication-naturals), with the proofs following directly from various properties listed in Table \@ref(tab:intprops).


```{theorem}
Let $m,n,k\in \mathbb{Z}$. Then $$(1) \: (m=n) \Leftrightarrow (m+k=n+k),  \quad (2) \: (m<n) \Leftrightarrow (m+k<n+k),$$
$$ (3) \: (m=n) \Leftrightarrow (m\times k=n\times k), \mbox{ with } k\neq 0 \quad \mbox{ and } \quad (4) \: (m<n) \Leftrightarrow (m\times k<n\times k), \mbox{ with } k> 0 .$$
```

### Properties of Exponents

Let $a$ be an integer. We define $a^n$ for each natural number $n$ using the following recursive definition.
$$a^0:=1, \quad a^1 := a, \quad \mbox{ and } \quad a^{n+1}:=a \cdot a^n\, \: \forall n\in \mathbb{N}.$$

Notice that we have defined $0^0=1$. This is not a universally accepted definition, but is widely accepted in the context of algebraic and combinatoric contexts, while left as indeterminate in a limit context of analysis, since
$$\lim_{x\rightarrow 0^+} x^0 = 1 \quad \mbox{ and } \quad \lim_{x\rightarrow 0^+} 0^x = 0.$$ 

>#### Related Content Standards {-}
>- (6.EE.1)] Write and evaluate numerical expressions involving whole-number exponents.

With this definition, we are able to prove many of the properties of exponents of natural numbers with a base of an integer. It is already given that $a^0=1$ and $a^1=a$. 

```{lemma, exponent-addition}
Let $a$ be an integer, and let $m$ and $n$ be natural numbers. Then 
$$a^m\cdot a^n=a^{m+n}.$$
```

```{proof}
That we are trying to prove something to be true for all natural numbers points us to a proof by induction argument. In this case, we will fix $n\in \mathbb{N}$ and let $P(m)$ be the statement that $a^m\cdot a^n=a^{m+n}$. 
    
Then $P(0)$ is the statement that $a^0\cdot a^n = a^{0+n}$. Since $a^0=1$ and $0+n=n$, we have that $1\cdot a^n = a^n$, which is a true statement.
    
If we assume that $P(k)$ is true for some value of $k$ (we already know that it is true for $k=0$), then $a^k\cdot a^n = a^{k+n}$. This means that
\begin{align*}
    a^{(k+1)} \cdot a^n & = (a \cdot a^k) \cdot a^n \mbox{ (by the definition of the exponent)} \\
    &= a \cdot (a^k\cdot a^n) \mbox{ (since multiplication is associative for integers)} \\
    &= a \cdot a^{k+n} \mbox{ (by the assumption that $P(k)$ is true)} \\
    &= a^{(k+n+1)} \mbox{ (by the definition of the exponent)} \\
    &= a^{(k+1)+n} \mbox{ (since addition is associative for the natural numbers)}.
\end{align*}
So, $P(k+1)$ is true. 
    
Therefore, by induction, $P(m)$ is true for all natural numbers $m$.
```




```{lemma}
Let $a$ and $b$  be any integers, and let $n$ be any natural number. Then $$(ab)^n=a^n \cdot b^n.$$
```

```{proof}
Since we are proving a statement to be true for all natural numbers, we will use a proof by induction with the statement $P(n)$ being that $(ab)^n=a^n\cdot b^n$.
    
Since $P(0)$ is the statement that $(ab)^0=a^0 \cdot b^0$, we have that this is equivalent to the statement that $1=1$, and so is true.
    
If we assume that $P(k)$ is true for some value of $k$, then we have that $(ab)^k=a^k\cdot b^k$. We can then use properties of addition and multiplication of integers, and the definition of exponents, to see that $$(ab)^{(k+1)}= (ab)^k\cdot (ab) = a\cdot (ab)^k \cdot b = (a\cdot a^k) \cdot (b^k\cdot b)  = a^{(k+1)} \cdot b^{(k+1)}$$ and so $P(k+1)$ is true.
    
Therefore, by induction, $P(n)$ is true for all $n\in \mathbb{N}$.
```

```{lemma, exponent-product}
Let $a$ be an integer and $n$ and $m$ be any natural numbers. Then $$(a^m)^n = a^{mn}.$$
```


The proof of this lemma is similar to the proof of Lemma \@ref(lem:exponent-addition) and so will be left as an exercise.




```{lemma}
Let $a$ and $b$ be integers with $0<a<b$. Then for each natural number $n\geq 1$, $0<a^n<b^n$.
```

```{proof}
We will use an induction argument for fixed integers $a$ and $b$, with $a<b$. We let $P(n)$ be the statement, $0<a^n<b^n$. Since $a^1=a$ and $b^1=b$, we see that $P(1)$ is true. If for some $k\in \mathbb{N}$, with $k\geq 1$, $P(k)$ is true, then 
$$ a^{(k+1)}=a^k\cdot a^1 < b^k \cdot a^1 < b^k \cdot b^1 = b^{(k+1)}$$ using the induction hypothesis and Lemma \@ref(lem:order-multiplication-naturals). Since the natural numbers are closed under multiplication, and therefore also under exponents, we have that $P(k+1)$ is true. Therefore, $P(n)$ is true for all $n\in \mathbb{N}$, with $n\geq 1$.
```

We will summarize these results in the following theorem and extend these properties to other number systems throughout the remainder of the chapter.

```{theorem, exponents-integers}
Let $a$ and $b$ be integers.
    
a. $a^0=1$

a. $a^1=a$
  
a. $a^m\cdot a^n = a^{m+n}$ for each $m,n\in \mathbb{N}$
    
a. $(ab)^n=a^n\cdot b^n$ for each $n\in \mathbb{N}$
    
a. $(a^m)^n = a^{mn}$ for each $m,n\in \mathbb{N}$
    
a. If $0<a<b$, $0<a^n<b^n$ for each $n\in \mathbb{N}$

  
```


### Exercises

1. After introducing the associative property for addition, a student asks if there is something similar for subtraction so that
    $$ A-(B-C) = (A-B)-C.$$ How would you respond to the student?
    
1. Prove that the relation on $\mathbb{N}\times\mathbb{N}$ defined by $$(a,b)\sim (c,d) \quad \Leftrightarrow \quad a+d=b+c$$ is an equivalence relation.

1. Prove Lemma \@ref(lem:order-multiplication-naturals).
    
1. Prove Lemma \@ref(lem:exponent-product).



## Representations of Integers {#sec:integer-representation}


Implicit to our discussion of number systems is the idea of representation. While the previous chapter focused on the concept of "3" in terms of equivalence and equality, representation is concerned with how we choose to communicate that we are currently using three in our mathematical process. As advanced students of mathematics, it is easy to take for granted the ways that we communicate fundamental notions such as number or operation using symbols like "3." However, the numbers and symbols we take for granted are choices that have been made and agreed upon. 

While the choices often allow for efficient communication, they are often not the most obvious choices for developing conceptual understanding of the abstract objects they represent. For example, when was the last time you saw negative 3 things in the natural world? While we understand the notion of "I owe 3 dollars," the concept of owing money is an abstract concept, not a physical one. It takes a long time for students to develop the conceptual understanding of how to work with mathematical objects they cannot physically combine or otherwise manipulate. Fortunately, because the negative sign is a choice of one among many, we can choose different representations of negative numbers and other abstract mathematical objects to facilitate our teaching. This section examines some of the other choices we can make for representing our numeration system and the four common operations.


### Numeration Systems

When we write the number $13,436$, we are communicating how many of something we have. To understand just how many things $13,436$ is, we need two fundamental ideas: digit and place value. For example, in $13,435$ we used the \textit{digit}\index{Digit} $3$ twice: once in the tens place value and once in the thousands place value. In these cases, the digit $3$ communicates how many of each place value we have. Together, the set of digits and a shared understanding of place value gives us our current numeration system.




The Hindu-Arabic Numeration System is the numeration system we use in most of the world today. It is composed of three main ideas:

1. A set of $10$ digits: $\{0,1,2,3,4,5,6,7,8,9\}$
1. A base $10$ place value system
1. Reliance on a digit that means "none"

Note that these are choices: the base of a place value system determines how many digits you need. While our culture has settled on base $10$ (based on the number of digits on our hands), the Mayan's developed a base-$20$ system (based on the number of digits on hands and feet) and computer machine language uses a base $2$ system (based on a signal being on or off). 

The place value base is used to build the place value by using exponential powers of the base value. For example: 
$$1234.567 = 1 \cdot 10^3 + 2\cdot 10^2 + 3\cdot 10^1 + 4\cdot 10^0 + 5\cdot 10^{-1} + 6 \cdot 10^{-2} + 7 \cdot 10^{-3}.$$

Although this probably all sounds like review and very obvious, these are not obvious choices. It took a long time for human civilizations to develop the base number systems and children have to be indoctrinated into it. It is not an easy idea that "3" in one position does not have the same meaning as a "3" in another position. Although this section focuses mostly on representation of number, the use of symbols taking on multiple meanings in mathematics in different contexts is one with which mathematics teachers must repeatedly grapple. 

To better understand the power of these systems, we will examine a couple numeration systems that have different combinations of the properties of the Hindu-Arabic system.




```{example, name="Egyptian Numeration"}
The Egyptian numeration system uses individual Hieroglyphs to hold value, with new symbols needed for each power of $10$). To communicate multiples of a particular power of ten, the Egyptians used repeated Hieroglyphs. For example, the water lily or lotus Hieroglyphs means $1,000$. A sequence of three water lily Hieroglyphs would mean $3,000$. There are no requirements on how Hieroglyphs are presented, so the Hieroglyphs communicating "$1$" could appear at any place in the expression for the number, and did not even have to appear together. A clear con of this system is that new symbols are needed for each power of $10$ would require a new Hieroglyph. In addition, the lack of an ordering or organizational system would make large numbers potentially cumbersome to read.

```{r, echo=FALSE, fig.align='center', out.width='90%', fig.cap=''}
knitr::include_graphics(
  'tikz/hieroglyphs.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```







```{example, name="Roman Numeration"}
The Roman numeration system is similar to the Egyptian system, in that the Roman numbers are each unique to a particular value (for example, X means $10$). The Roman system improved on the Egyptian numeration system by imposing an order to how these numbers could be presented and partially solved the potential for a large number of repeated numbers by adding the subtraction element (so, IX would mean take $1$ away from $10$, while XI means add one to $10$). However, as students of the Super Bowl know, reading these numbers still takes a fair amount of decoding and there is a still a problem with needing a new number for each new grouping.

```{r roman, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
arabic <- c("1", "10", "100", "1,000")
rom <- c("I", "X", "C", "M")
roman <- data.frame(arabic, rom)
names(roman) <- c("Arabic Numeral", "Roman Numeral")
knitr::kable(roman, caption = 'Roman Numeration')
```







```{example, name="Base $5$"}
Base $5$ is like our current standard numeration system, except it uses a set of five digits, $\{0,1,2,3,4\}$, and place values are based on powers of $5$, not $10$. In the United States we use base $5$ in some of our coins, in that there are $5$ pennies in a nickel and $5$ nickels in a quarter. Then the number $14_{5}$ (read as fourteen base five) is saying $1$ nickel and $4$ pennies, or $9$ cents. Similarly,
$$231_{5} = 2\cdot 5^2 + 3\cdot 5^1 + 1 \cdot 5^0 = 66_{10}.$$
```

While there is no obvious use for students to learn operations in a different base, working outside of our base 10 system can help teachers recognize and understand the inner workings of the base 10 system and its associated algorithms. As these algorithms are introduced throughout the section, you are encouraged to use the algorithms to perform operations on numbers in these additional base systems.



```{example, name="Base $12$"}
The duodecimal (base 12) number system is a component of some languages in West Africa and Nepal and is used in our everyday measurements of dozen and gross. A duodecimal system requires twelve digits, $0,1,2,3,4,5,6,7,8,9,t,e$, and place values are based on powers of $12$. In this system $6+5=e$, $7+8=13$, and 
$$3t \times e4 = 2te4.$$
    
One of the implications of a base 12 system is additional tricks for multiplying numbers due to a larger number of factors of twelve than ten.
```

### Representations of Number

In talking about numeration systems, we have exclusively been using one form of number representation. However, numeration systems are only one way that we communicate ideas about number. In this section, we explore different ways of representing numbers, each which helps develop different properties of numbers. For this introduction to representation, we focus first on the positive natural numbers, adding additional elements to these representations as we develop the various number systems.





#### Sets.{-}
As discussed in Chapter \@ref(ch:equivalence), we can think of 3 as an equivalence class for all sets containing 3 objects. (We will go into this in more detail in Section \@ref(sec:cardinality).) As such, sets can be used to represent numbers, either pictorially as a circle containing generic objects or as a set with general elements. Using sets offers the most general representation of number we can develop, as it does not impose any structure on the representation beyond the property of ``how many''. Relationships between numbers (such as inequality or equality) are fairly straightforward to model, but comparing a large number at one time would be difficult using this representation. As a result, most middle and high school textbooks do not represent numbers this way, although it is fairly common in the elementary grades.


```{r, echo=FALSE, fig.align='center', out.width='20%', fig.cap=''}
knitr::include_graphics(
  'tikz/Set-representations.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```
or $\{a,b,c\}$.



    


#### Number Line. {-}
The number line imposes more structure on the number representation than the use of sets. In particular, it organizes numbers along a line, with integers spaced evenly along the line. This representation allows multiple numbers to be visualized at the same time while also showing relationships between the numbers represented. In this representation, any number that falls at the location of 3 is equivalent to three. Numbers of the left of 3 are less than it, while numbers above are greater.
```{r, echo=FALSE, fig.align='center', out.width='80%', fig.cap=''}
knitr::include_graphics(
  'tikz/number-line-representation.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```


#### Base-10 Blocks. {-}
Base-10 blocks are usually a physical way (although virtual and pictorial versions exist) to represent numbers specifically designed to promote understanding of place value. The blocks come in four sizes: units  (small, approximately 1-centimeter cubes), rods (10, represented by a stack of 10 unit cubes), flats (100, represented by a 10 by 10 grid of unit cubes), and big cubes (1000, represented by a cube with side lengths equal to 10 unit cubes). Representing a number using base-10 blocks means drawing or collecting the correct number of blocks for each place value.

```{r, echo=FALSE, fig.align='center', out.width='80%', fig.cap=''}
knitr::include_graphics(
  'tikz/base10blocks.jpg',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```





In order to write down the reasoning these blocks are sometimes transitioned into dots, lines, and squares. So that $13+24$ is represented by 
```{r, echo=FALSE, fig.align='center', out.width='60%', fig.cap=''}
knitr::include_graphics(
  'tikz/base10block-addition.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```






#### Cuisenaire Rods. {-}
Cuisenaire rods are another physical way that numbers are represented. They are a collection of long prisms of different colors and lengths. Rods of the same length are the same color, and each length is a natural number multiple of the smallest rod. These rods are used for a variety of reasons including factoring of natural numbers, but one of the most common is to help students understand concepts related to rational numbers and the operation of division.

```{r, echo=FALSE, fig.align='center', out.width='30%', fig.cap='Cuisenaire Rods'}
knitr::include_graphics(
  'images/Cuisenaire-Rods-2.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```





### Addition Models and Algorithms

Much like the concept of number, it is easy to overlook how quickly addition becomes a complex endeavor for young students of mathematics. Fundamentally, addition represents the notion of ``combine'' or finding the union of mutually disjoint sets.

```{r, echo=FALSE, fig.align='center', out.width='30%', fig.cap=''}
knitr::include_graphics(
  'tikz/addition-model-set.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

Typically, we quickly move away from representing addition using sets and instead use the number line. For natural number addition, this process is fairly straight forward: Students start by identifying the first addend. Then, count up the number that is being added.  The number you land on is the result.

```{r, echo=FALSE, fig.align='center', out.width='60%', fig.cap=''}
knitr::include_graphics(
  'tikz/addition-model-numberline.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

Thinking of addition as the union of sets or on a number line of the natural numbers are good, conceptual starting places for understanding addition. However, doing so will lead some students to think that “when two numbers are added, the result will be the same size or bigger.” In the natural numbers this thinking will work, but in the integers the reasoning begins to fall apart. Note that idea of “combine” does not say anything about the size of the result relative to the others, so the common error is a result of students focusing on the result rather than the concept of addition.

When asked to add something like $125+258$ without the use of a calculator, most adults are well-trained to rewrite the sum and apply the standard algorithm.


```{r, echo=FALSE, fig.align='center', out.width='20%', fig.cap=''}
knitr::include_graphics(
  'tikz/addition-model-algorithm.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

While the work of teaching children these algorithms is usually done in elementary school, it is worth taking a little time to appreciate the ways these traditional methods take advantage of the properties of natural numbers to allow for quick and efficient calculation. To facilitate this discussion, we present several alternative algorithms. It is not our goal to instruct our readers in each of these algorithms: there is not a way to present all possible ways an addition problem can be solved. Rather, we choose algorithms that maximize the opportunity to explore the hidden work of the standard algorithm. A mathematics teacher who studies these examples in relation to the standard algorithm should be able to teach themselves unfamiliar algorithms that they encounter in the wild. For each of these, we use the example  $125+258$.

#### Expanded form. {-}
We can rewrite our numbers using place value as $(100+20+5)$ and $(200+50+8)$ and then use associative and commutative properties of addition for the natural numbers.
    
\begin{align*}
    125+258 &= (100+20+5) + (200+50+8) \\
    &= (100+200) + (20+50) + (5+8)  \\
    &= (100+200) + (20+50) + (13) \\
    &= (100+200) + (20+50) + (10 + 3) \\
    &= (100+200) + (20+50+10) + (3) \\
    &= 300 + 80 + 3 =383 \\
\end{align*}

#### Lattice Algorithm. {-}

The lattice algorithm described below provides another method of adding two numbers by keeping track of place value through some implicit methods.


+ Stack the two numbers so that corresponding place values are in the same column. Draw the traditional equals bar.

```{r, echo=FALSE, fig.align='center', out.width='20%', fig.cap=''}
knitr::include_graphics(
  'tikz/lattice-addition1.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

+ Starting with the right-most column, add within each column. The result of the digit addition will be a number between 0 and 19. The resulting number is written into the square below the column, with the higher place value going into the higher triangle. 

```{r, echo=FALSE, fig.align='center', out.width='20%', fig.cap=''}
knitr::include_graphics(
  'tikz/lattice-addition2.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

+ Add down each diagonal.  You are in effect adding the total number of items in each place value.  

```{r, echo=FALSE, fig.align='center', out.width='20%', fig.cap=''}
knitr::include_graphics(
  'tikz/lattice-addition3.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

#### Expanded algorithm. {-}

The expanded algorithm makes the implicit structure of the lattice algorithm more explicit. 

```{r, echo=FALSE, fig.align='center', out.width='10%', fig.cap=''}
knitr::include_graphics(
  'tikz/addition-model-algorithm-expanded.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

You may have noticed that regardless of which method we used, there were several common elements that all of the algorithms created in one form or another. Perhaps most obviously, $13$ occurred in all three calculations. The intermediate sums an algorithm produces are called **partial sums**. Note that although every algorithm used partial sums, they are not always the same between algorithms, even for the same addition problem. These differences in partial sums are a result of where the algorithm does the **regrouping** of the partial sums. Regrouping, generally, is the process of converting numbers expressed in terms of one place value into a different place value. For example, in the standard algorithm example at the beginning of this section, we got $13$ when we added the ones digits in the right most column. Since $13$ ones is not useful for continued adding, we regrouped $13$ as $1$ ten and $3$ ones. We wrote the $3$ in the ones place value and the $1$ above the $10$'s place value column. You may have learned this process by the name of "carrying the one". This language is no longer encouraged in teaching as it masks the fact that the value of the one changes as a person works through every place value.



### Subtraction Models and Algorithms

Conceptually, we described addition as the process of combining. Subtraction has two different physical interpretations: the process of removal (often referred to as taking away) or the inverse operation of addition. These are clearly related, but are developed separately, which we explore a bit more in the following paragraphs. Although both interpretations of subtraction seem straight forward when modeled in the physical world, students struggle with this operation a great deal more than addition. This may be at least partially because while the natural numbers have some great properties under addition (closed, commutative, and associative) that make addition easy to set up, but none of these properties under subtraction. 

Using our different representations, we can model subtraction a couple different ways: the take away model or the missing addend model. 

#### Take-Away Model. {-}
The first interpretation of subtraction is as removal or the process of taking away. In the mathematics education literature, this is called the "take-away model". This conceptualization of subtraction is a direct interpretation of the calculation $a-b$, where the translation of this expression is "$a$ take away $b$."

To model $a-b$ using sets and the take-away model, we start with a set containing $a$ elements and remove $b$ elements from it. The composition of the set after the removal process is the resulting difference.

```{r, echo=FALSE, fig.align='center', out.width='30%', fig.cap=''}
knitr::include_graphics(
  'tikz/subtraction-model-sets.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

To model $a-b$ using the number line and the take-away model, we find $a$ on the number line and then move back $b$ units. The place where you land is the resulting difference.

```{r, echo=FALSE, fig.align='center', out.width='70%', fig.cap=''}
knitr::include_graphics(
  'tikz/subtraction-model-numberline.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

Once we have introduced the integer number system, we extend the number line and think of $a-b$ as $a+(-b)$ where $-b$ is viewed as a vector of length $|b|$ pointed in the negative direction.

```{r, echo=FALSE, fig.align='center', out.width='80%', fig.cap=''}
knitr::include_graphics(
  'tikz/subtraction-model-numberline2.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

>#### Related Content Standards {-}
>- (6.NS.6) Understand a rational number as a point on the number line. Extend number line diagrams and coordinate axes familiar from previous grades to represent points on the line and in the plane with negative number coordinates.
>     a. Recognize opposite signs of numbers as indicating locations on opposite sides of $0$ on the number line; recognize that the opposite of the opposite of a number is the number itself, e.g., $-(-3) = 3$, and that $0$ is its own opposite.


#### Missing Addend Approach. {-}
The missing addend approach interpretation of subtraction is an explicit model of subtraction as the inverse operation of addition, while converting the computationally more nuanced subtraction problem into an addition problem where the number facts and such are easier. In particular, in the missing addend approach, the subtraction expression  $a-b$ is assumed to equal some unknown (so $a-b=?$), which we can rewrite as $b+?=a$. 

Using the set model and the missing addend approach, we take the statement $b+?=a$ to mean that we know that we want a set to contain $a$ elements and we have $b$. We must determine how many additional elements we need in order to bring the total number of things in our set up to $a$.


```{r, echo=FALSE, fig.align='center', out.width='40%', fig.cap=''}
knitr::include_graphics(
  'tikz/subtraction-model-missingaddend-set.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

Using the number line, we reconceptualize $a-b=?$ As $b+?=a$ which means we start at $b$ on the number line and then find $a$. We then ask, how much do I have to move (and in what direction) from $b$ to get to $a$?. The answer to this question is the answer to the original subtraction problem. This process when formalized into an algorithm that counts up by place value is called **counting up**. 

```{r, echo=FALSE, fig.align='center', out.width='80%', fig.cap=''}
knitr::include_graphics(
  'tikz/subtraction-model-missingaddend-numberline.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

Note that in none of the conceptual work of this section have we used the language of larger number and smaller number, instead focusing on the expression of the difference, $a-b$. Early instruction in subtraction only presents cases of subtraction problems $a-b$ when $a\geq b$ and both are natural numbers. As a result of this presentation bias, students sometimes develop the impression that the results of subtraction will always be smaller than the original number. They may also believe that subtraction when $a<b$ cannot be done. The conceptual work we have developed in this section for the models of subtraction works for most sets of numbers students encounter in the K-12 mathematics curriculum.  

As with addition, we have a standard subtraction algorithm. For example, consider the subtraction problem $423-368$. A well-trained student asked to do this problem without a calculator will produce work that looks something like the following:

```{r, echo=FALSE, fig.align='center', out.width='30%', fig.cap=''}
knitr::include_graphics(
  'tikz/subtraction-model-algorithm.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

As with addition, the traditional subtraction algorithm uses regrouping strategies and the properties of place value to allow for quick calculation in a small amount of space. (You may have learned the regrouping strategy under the name "borrowing" but, as with addition, this name is no longer used for similar reasons.) 

We next provide several alternative algorithms for subtraction designed to help develop conceptual understanding of the subtraction algorithm. 


#### Equal Addition. {-}
The equal addition algorithm repeatedly adds zero to the problem in a way that makes the subtraction easier to compute. The goal is to write an equivalent problem that has at least one zero place value.

\begin{align*}
423-368 &= 423-368 + (2-2) = 425-370 \\
  &= (425-370) + (30-30) = 455-400 \\
  &= 55
\end{align*}



#### Partial Differences. {-}
Another method of subtraction is to subtract from left to right with partial differences. In this case, we subtract the smaller digit from the larger digit, keeping track of positive or negative.

```{r, echo=FALSE, fig.align='center', out.width='80%', fig.cap=''}
knitr::include_graphics(
  'tikz/subtraction-model-partial-differences.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

In the section on subtraction, we commented that when students first learn subtraction in the natural numbers, they are usually only given examples of the form $a-b$ where $b\leq a$. Part of the reason for only having students work on these types of problems is that children are not born knowing about negative numbers. The only place we ``see'' negative numbers in the real world is when humans have imposed a scale on something, such as temperature. However, this is not a true physical representation, since we have simply imposed a scale on a naturally occurring thing, where for one reason or another, we want to have a zero reference point. 

To allow us to find answers to subtraction problems where $a<b$, we expand our number system to the integers. 
You may have noticed that we use "$-$" in the above paragraph to mean additive inverse of a positive number. In the paragraph before that we used the same symbol to mean subtraction. In reality, these are fairly different concepts. Subtraction is an operation.  Additive inverse is a relation. Long-time students of mathematics know that these interpretations are related and will move between them fairly seamlessly, but they are not the same thing.


### Multiplication Models and Algorithms

#### Repeated Addition. {-}
Most commonly, multiplication is described as **repeated addition**. For example, the product $2\times 4$ can be understood as adding $2$ four times ($2+2+2+2$), while the product $4\times 2$ can be understood as adding $4$ two times ($4+4$). While conceptually, this link between addition and multiplication is important, it quickly loses efficiency in practical applications (e.g., it would be cumbersome but possible to multiply $23\times 345$ using repeated addition). that said, it is not uncommon to see secondary students apply this approach to multiplication problems.


#### Rectangular Array. {-}
Another way to represent multiplication is through a **rectangular array**. One way to think about this is counting the number of chairs in $a$ rows and $b$ columns. Such arrays are discrete sets. We can see below how one can use two arrays to represent $3\times 5$ and $5 \times 3$.


```{r, echo=FALSE, fig.align='center', out.width='50%', fig.cap=''}
knitr::include_graphics(
  'tikz/arrays.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```




We can see that the two arrays are related to each other by a rotation. In other words, the order of multiplication is simply a different perspective on the same array.

#### Area Model. {-}
While arrays work well for representing multiplication of natural numbers, they do not work well for integers, rational numbers, or real numbers. So we generalize the array model to an **area model**. If we change the dots in the array to blocks of 1 square unit measure, the above arrays become rectangles with square tiles helping students see the connection between the area of a rectangle and multiplication.

```{r, echo=FALSE, fig.align='center', out.width='50%', fig.cap=''}
knitr::include_graphics(
  'tikz/area-model-3by5.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

This area model can then be generalized to help students multiply multi-digit natural numbers and the relationship to the distribution property.
```{r, echo=FALSE, fig.align='center', out.width='40%', fig.cap=''}
knitr::include_graphics(
  'tikz/area-model-13-16.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```


#### Lattice Multiplication. {-}
The area model can become unwieldy with larger numbers. The lattice multiplication algorithm is a more abstract view of the area model by using a lattice to keep track of place values. Here we can see the progression of using a lattice model to multiply $54 \times 68 = 3672$.

```{r, echo=FALSE, fig.align='center', out.width='50%', fig.cap=''}
knitr::include_graphics(
  'tikz/lattice-multiplication-13-16-1.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```
```{r, echo=FALSE, fig.align='center', out.width='50%', fig.cap=''}
knitr::include_graphics(
  'tikz/lattice-multiplication-13-16-2.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```
```{r, echo=FALSE, fig.align='center', out.width='50%', fig.cap=''}
knitr::include_graphics(
  'tikz/lattice-multiplication-13-16-3.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

After setting up the lattice with diagonals, we multiply all of the partial products and enter the product in the corresponding square with the one's digit in the bottom right and the ten's digit in the upper left. After all of the partial products are complete, we add along the diagonals in order to find the number of units of that place value. For some products, the sum along the diagonal is larger than $9$ and so we have to regroup as is the case for $1564 \times 63$.
```{r, echo=FALSE, fig.align='center', out.width='60%', fig.cap=''}
knitr::include_graphics(
  'tikz/lattice-multiplication-large.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```




#### Partial Products. {-}
We continue the progression of creating more abstract algorithms for multiplication by removing some of the explicit structure using the **partial products algorithm**.

```{r, echo=FALSE, fig.align='center', out.width='60%', fig.cap=''}
knitr::include_graphics(
  'tikz/partial-products.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```





#### Standard U.S. Multiplication Algorithm. {-}
If we continue down the path of more abstract algorithms we come to what is considered the standard algorithm for multiplication. In this algorithm we compress the information in the partial products algorithm into many fewer lines.

```{r, echo=FALSE, fig.align='center', out.width='60%', fig.cap=''}
knitr::include_graphics(
  'tikz/multiplication-algorithm-standard.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

However, with this reduction in the number of lines comes an increase in the number of things that one must keep track of implicitly.
    


### Division Models and Algorithms

Division is the fourth major operation. Conceptually, the division problem $a \div b$ can be viewed several ways: as repeated subtraction, as the inverse of multiplication, and as a partition. Regardless of the model, all three of these representations draw on the notion that when we divide $a$ by $b$ ($a\div b$) the division operation produces two unique numbers: the **quotient**, $q$, and the **remainder**, $r$, where $0\leq r <b$. (We will go into this division algorithm more in-depth in Section \@ref(sec:Integer-Ring).) We call $b$ the **divisor** and $a$ the **dividend**.  These four quantities are related by $a=bq+r$.


#### Repeated Subtraction Model. {-}

In the repeated subtraction model of division, the division problem $a\div b$ conceptually is thought of as repeatedly subtracting $b$ from $a$ until you no longer have a number bigger than $b$. The quotient is the number of times you could subtract and the remainder is what is left over after the quotient has been determined.

#### Missing Factor Model. {-}
In the missing factor model, the division problem defines division as the inverse operation of multiplication. The quotient and remainder are not separated. Instead, we reorganize the quantities in their reciprocal relationships in the equation $a=bq+r$. Generally, we use the missing factor approach when $r=0$. In particular, if $a\div b = q$, where $q$ is a unique value, then we rewrite the problem as the product $bq=a$.

#### Set (Partition) Model. {-}
The set (partition) model frames the quotient as a partition of the whole, $a$, into $b$ even parts. In particular, $a\div b$ is thought of as asking if I must evenly distribute $a$ between $b$ entities, how much will each entity get? While in practice, the finding of the quotient for both the partition and repeated subtraction models are similar, conceptually, the two methods handle the remainder differently. The repeated subtraction approach reports the remainder, while the partition model invites the person completing the problem to keep going until all parts of $a$, including $r$, have been evenly distributed.
    
#### Division Algorithms

The long division algorithm is the last major algorithm students learn. If we examine the work of the long division algorithm, presented below, a few things are apparent. 
First, the algorithm is the only one that does not begin work with the ones place value, making it an anomaly in students' learning. Secondly, it relies on students to have mastered the subtraction algorithm. And lastly, when we have multi-digit divisors, estimation is an important prerequisite skill. These factors all make the division algorithm particularly tricky to master.

```{r, echo=FALSE, fig.align='center', out.width='60%', fig.cap=''}
knitr::include_graphics(
  'images/Long_division.jpg',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

With the widespread availability of cheap and accurate calculators, it is worth asking why we still teach the division algorithm (or really, most advanced arithmetic). And indeed, long division is not emphasized in the curriculum as much as it used to be for this very reason. We do provide a brief treatment here as the division algorithm, while not widely used in daily life, does lay a foundation for polynomial division, discussed in later chapters. In partial products algorithm presented below, pay particular attention to how the various division models are codified into the structure of the algorithm and how the partial quotients differ or are similar to those of the standard algorithm.

```{r, echo=FALSE, fig.align='center', out.width='60%', fig.cap=''}
knitr::include_graphics(
  'tikz/division-algorithm-standard.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

### Exercises
1. Use the definition of base $5$ in this section to answer the following questions.
    a. What is the base $5$ number following $4_{5}$?
    a. Express $234_{5}$ in base $10$.
    a. Express $234_{10}$ in base $5$.

1. Add $342_5+134_5$ using each of the four addition algorithms presented in this section.
 
1. Complete the product $234_5 \times 32_5$ using each of the three multiplication algorithms presented in this section (lattice, partial products, standard).
    
1. Complete the division problem $341_5 \div 41_5$ using any division algorithm discussed above.
    
1. Construct the addition and multiplication tables for the first twelve digits in a base $12$ system. What patterns do you notice?
    
1. Read up on the Mayan numeration system, which is a base $20$ system.
    a. Describe the patterns used to develop the 20 digits.
    a. How is this system comparable to the numeration systems discussed in this section.

1. For the two-number lattice addition, will adding along a diagonal ever require a student to regroup? Why or why not?
    
1. The standard addition algorithm requires that a person adds columns right to left. Many students just learning the algorithm will make the mistake of starting with the left most column and working towards right.
    a. What conceptual understanding is a student missing when they make this type of error?
    a. Which of the three non-standard algorithms will result in a correct sum regardless of whether a person adds right to left or left to right?
    a. Which of the four algorithms presented in this section might help students develop the conceptual understanding you identified in part (a)? 
    
1. Another common error students make with the standard addition algorithm is that students will line up the first digit of a number with the first digit of the second number, regardless of whether or not the numbers have the same number of digits. 
    a. What conceptual understanding is a student missing when they make this type of error?
    a. Which of the four algorithms presented in this section might help students develop the conceptual understanding you identified in part (a)? 
1. Using a set of base-ten blocks (or a virtual base-ten block simulation), model the addition problem $348+753$. 
    a. Write out the arithmetic steps you went through in completing the addition using the blocks.
    a. Describe how base ten blocks might support conceptual understanding of the standard addition algorithm.

1. How does the standard multiplication algorithm utilize each of the four properties of the natural numbers (commutative, associative, distributive, identity)?
    
1. Why, in multiplication by multi-digit numbers in the standard multiplication algorithm, do you sometimes include zeros in the right most digit places?
    
1. For each model of division (e.g., repeated subtraction, partitions, missing factor), explain why the quotient $a\div 0$ does not make sense.
    
## Rational Numbers {#sec:rationals}

Since the only integers that have a multiplicative inverse are $1$ and $-1$, it would be beneficial to expand our number system to include symbols that can operate as multiplicative inverses. In this process we allow the ordered pairs in $\mathbb{Z}\times (\mathbb{Z}-\{0\})$ to be written as a fraction, $$\frac{a}{b}\cong (a,b).$$
But we also want the set to be closed under addition and multiplication. This results in a set that contains all possible rational expressions of integers whose denominator is non-zero. Therefore we create the set $$\widehat{\mathbb{Q}}= \left\{ \frac{m}{n}\middle \vert m,n \in \mathbb{Z}, n\neq 0\right\} \cong \mathbb{Z}\times (\mathbb{Z}-\{0\}).$$  Then, similar to our process of constructing the integers from the natural numbers, we define a relation of two rational expressions by 
$$\frac{a}{b} \sim \frac{c}{d} \Leftrightarrow ad=bc.$$ 


We will now verify that this relation of rational expressions represents an equivalence relation. 

#### Reflexive. {-}
One can see that this relation is reflexive, since if $\frac{a}{b}$, with $b\neq 0$, is a rational expression of integers that $ab=ab$ and so $\frac{a}{b}$ is equivalent to itself. 

#### Symmetric.{-}
Since equality of integers is a symmetric relationship we have that for any integers $a,b,c,d$ with $ad=bc$ that $bc=ad$ and so this equivalence of rational expressions is symmetric.

#### Transitive. {-}
The more challenging property of equivalence relations to prove is the transitive property. To prove transitivity, we assume that we have three rational expressions, $$\frac{a}{b}, \: \frac{c}{d}, \: \mbox{ and } \: \frac{m}{n},$$ with $b$, $d$, and $n$ all non-zero, with $$\frac{a}{b}\sim \frac{c}{d} \quad \mbox{ and } \quad \frac{c}{d} \sim \frac{m}{n}.$$  From the definition of equivalence of rational expressions we have that $ad=bc$ and $cn=md$. From the properties of multiplication of integers we have that $$(ad)\cdot (cn) = (bc) \cdot (md),$$ or equivalently,
$$ancd = bmcd.$$ This can then be rewritten as $$ (an-bm)\cdot (cd) =0.$$ Since the integers have the property that the product of two non-zero integers is also non-zero, we know that either $an-bm=0$ or $cd=0$. Since $d$ is assumed to be non-zero, $cd=0$ would imply that $c=0$. This would then imply that both $ad$ and $md$ are zero, since $ad=bc$ and $cn=md$. Since $d\neq 0$ we would then have that $a=0$ and $m=0$, and consequently that $an=bm=0$. Therefore, $an-bm=0$ independently of $cd=0$. Therefore, we have that $an=bm$ and that $\frac{a}{b}\sim \frac{m}{n}$, implying that equivalence of rational expressions is transitive.


Since this relation on rational expressions is an equivalence relation we should explore what the equivalence classes of these rational expressions look like.

We will first look at the rational expressions that are equivalent to $\frac{0}{1}$. We see that 

$$\left[\frac{0}{1}\right] = \left\{ \frac{a}{b} \: \middle \vert \: b\neq 0, \: 0\cdot b=1\cdot a\right\} = \left\{\frac{0}{b} \: \middle \vert \: b \neq 0 \right\}$$
Similarly, 
$$\left[ \frac{1}{1}\right] = \left\{ \frac{a}{b} \: \middle \vert \: b\neq 0, \: 1\cdot b = 1 \cdot a\right\} = \left\{ \frac{a}{a} \: \middle \vert \: a \neq 0\right\}.$$

We can follow this process and see that we can embed the integers into this set of equivalence classes using the correspondence of
$$ n\in \mathbb{Z} \quad \leftrightarrow \quad \left[\frac{n}{1} \right] .$$

### Operations

Now that we understand the set of equivalence classes of rational expressions, we need to define our operations of addition and multiplication on this set and verify that these definitions are well-defined.

```{definition}
Let $\left[\frac{a}{b}\right]$ and $\left[\frac{c}{d}\right]$ be two equivalence classes of rational expressions with $a,b,c,d\in \mathbb{Z}$ and $b,d\neq 0$. We define addition and multiplication as 
$$\left[\frac{a}{b}\right] + \left[\frac{c}{d}\right] := \left[\frac{ad+bc}{bd}\right] \quad \mbox{ and } \quad \left[\frac{a}{b}\right] \times \left[\frac{c}{d}\right] := \left[ \frac{ac}{bd} \right].$$
```

Since $b$ and $d$ are non-zero integers, the no non-zero zero divisors property tells us that $bd\neq 0$ and so the operations of addition and multiplication are closed. We can make sure that they operate appropriately on the equivalence classes by making sure that different representations for the original equivalence classes of rational expression generate a sum and product that are elements of the same equivalence class. The process is similar to the same process that was done with the integers. Since we did the proof for addition being well-defined on the integers, we will do the proof that multiplication is well-defined for the rational numbers.

Let $\frac{a}{b}\sim \frac{a'}{b'}$ and $\frac{c}{d} \sim \frac{c'}{d'}$. Using the definition of the equivalence classes, we have that $$a\cdot b' = b \cdot a' \quad \mbox{ and } \quad c \cdot d' = d \cdot c'.$$

Therefore, 
\begin{align*}
    (bd)\cdot (a'c') &= (ba') \cdot (dc') \quad \mbox{ by the commutative property of multiplication} \\
    &= (ab') \cdot (dc') \quad \mbox{ since } ab' = ba' \\
    &= (ab') \cdot (cd') \quad \mbox{ since } cd'=dc' \\
    &= (ac) \cdot (b'd') \quad \mbox{ by the commutative property of multiplication}
\end{align*}
and so we have that $\frac{ac}{bd} \sim \frac{a'c'}{b'd'}$.

So we have constructed a new number system that we call the rational numbers, $\mathbb{Q}$. In order to simplify our notation, we will no longer write each rational number as an equivalence class. We will instead just keep in mind the equivalence class structure and use any of the possible elements of the equivalence class to represent the rational number. We will also often write rational expressions of the form $\frac{n}{1}$ as the corresponding integer $n$. We will also write
$$\frac{-a}{b}, \quad \frac{a}{-b}, \mbox{ and }\quad -\frac{a}{b} $$ as equivalent expressions.


>#### Related Content Standards {-}
>+ (7.NS.2)]Apply and extend previous understandings of multiplication and division and of fractions to multiply and divide rational numbers.
>     a. Understand that multiplication is extended from fractions to rational numbers by requiring that operations continue to satisfy the properties of operations, particularly the distributive property, leading to products such as $(-1)(-1) = 1$ and the rules for multiplying signed numbers. Interpret products of rational	numbers by describing real-world contexts.
>     a. Understand that integers can be divided, provided that the divisor is not zero, and every quotient of integers (with non-zero divisor) is a rational number. If $p$ and $q$ are integers, then $-(p/q) = (-p)/q = p/(-q)$. Interpret quotients of rational numbers by describing real world contexts. 
>     a. Apply properties of operations as strategies to multiply and divide rational numbers.
        
### Order

We can define an order on $\mathbb{Q}$ by first defining the relationship between a rational expression and $0$, built on the order on the integers. We say that for a rational expression $\frac{a}{b}$, with $b\neq 0$ that
$$ \left(\frac{a}{b} < 0 \Leftrightarrow ab<0\right),  \quad \left(\frac{a}{b}=0 \Leftrightarrow ab=0\right), \mbox{ and } \quad \left(\frac{a}{b} >0 \Leftrightarrow ab >0\right).$$
From the no non-zero divisors property of the integers, we know that one, and only one, of the three statements can be true for a single rational expression. If $\frac{c}{d}\sim \frac{a}{b}$, we use that $ad=bc$ and  $$(ad)^2 = (ad)(ad) = (ad)(bc) = (ab)(cd)$$ to see that $ab$ and $cd$ have the same relationship to $0$. So this ordering with respect to $0$ is well-defined.

We can then extend the ordering to the entire set of rational numbers by defining
$$\left(\frac{a}{b} < \frac{c}{d} \Leftrightarrow \left(\frac{a}{b} + \frac{-c}{d}\right) <0 \right), \quad
\left(\frac{a}{b} = \frac{c}{d} \Leftrightarrow \left(\frac{a}{b} + \frac{-c}{d}\right) = 0 \right),$$
$$\mbox{ and } \quad \left(\frac{a}{b} > \frac{c}{d} \Leftrightarrow \left(\frac{a}{b} + \frac{-c}{d}\right) > 0 \right).$$

If we have rational numbers $\frac{a}{b}$, $\frac{m}{n}$, and $\frac{p}{q}$ with 
$$\frac{a}{b} < \frac{m}{n}, \quad \mbox{ and } \quad \frac{m}{n} < \frac{p}{q},$$ then 
$$\frac{m}{n}+ \frac{-a}{b} >0, \quad \frac{p}{q}+ \frac{-m}{n} >0$$ and so 
$$\frac{p}{q} + \frac{-a}{b} = \frac{p}{q}+ \frac{-m}{n} + \frac{m}{n}+ \frac{-a}{b}>0$$
and this relation satisfies the properties of an order and we can extend the properties of inequalities from the integers.

```{theorem}
Let $r,s,t\in \mathbb{Q}$. Then $$(1) \: (r=s) \Leftrightarrow (r+t=s+t),  \quad (2) \: (r<s) \Leftrightarrow (r+t<s+t),$$
$$(3) \: (r=s) \Leftrightarrow (r\times t=s\times t), \mbox{ with } t\neq 0 \quad \mbox{ and } \quad (4) \: (r<s) \Leftrightarrow (r\times t<s\times t), \mbox{ with } t> 0.$$
```

```{proof}
In order to get the idea of the proof we will prove parts (1) and (4) and leave the proofs of the other two parts to the reader.

Let $r=\frac{a}{b}$, $s=\frac{c}{d}$, and $t=\frac{m}{n}$ be rational numbers with $a,b,c,d,m,n\in \mathbb{Z}$ and $b,d,n\neq 0$.
    
Part (1). Then $$(r=s) \Leftrightarrow (ad=bc) \Leftrightarrow \left( (ad)n^2 = (bc) n^2 \right) \Leftrightarrow \left( (an+bm)(dn)=(cn+dm)(bn) \right) $$ since $d$ and $n$ are non-zero. We then see that 
$$\left( (an+bm)(dn)=(cn+dm)(bn) \right) \Leftrightarrow \left(\frac{an+bm}{bn} = \frac{cn+dm}{dn}\right) \Leftrightarrow  (r+t=s+t)$$ by the definition of equality of rational numbers and properties of the operations on the integers. 
    
Part (4). We can assume without any loss of generality that $b,d,n >0$ by changing the signs in the numerators, if needed. Then we know that $$ (r<s) \Leftrightarrow ((s-r)>0 ) \Leftrightarrow \left( \frac{cb-ad}{bd} >0 \right) \Leftrightarrow \left( (cb-ad)(bd) >0\right) \Leftrightarrow \left( (cb-ad) >0\right)$$ since $bd>0$.
    
Similarly, $$\left( r\times t < s\times t\right) \Leftrightarrow \left( \frac{am}{bn} < \frac{cm}{dn} \right) \Leftrightarrow \left( \frac{bcmn - admn}{bdn^2} >0\right). $$ Since $b,d,n >0$, we know that this is equivalent to 
$$\left( (bcmn-admn)>0 \right) \Leftrightarrow \left( (cb-ad)(mn) > 0 \right) \Leftrightarrow \left( (cb-ad) >0 \right)$$ since $t>0$ and $n>0$ implies that $m>0$ and $mn>0$.

Therefore we have that $$(r<s) \Leftrightarrow \left( r\times t < s\times t\right).$$
    
```

>#### Related Content Standards {-}
>+ (7.NS.1)] Apply and extend previous understandings of addition and subtraction to add and subtract rational numbers; represent addition and subtraction on a horizontal or vertical number line diagram.
>     a. Describe situations in which opposite quantities combine to make 0. For example, a hydrogen atom has 0 charge because its two constituents are oppositely charged.
>     a. Understand $p + q$ as the number located a distance $|q|$ from $p$, in the positive or negative direction depending on whether $q$ is positive or negative. Show that a number and its opposite have a sum of $0$ (are additive inverses). Interpret sums of rational numbers by describing real-world contexts.
>     a. Understand subtraction of rational numbers as adding the additive inverse, $p- q = p + (-q)$. Show that the distance between	two rational numbers on the number line is the absolute value of their difference, and apply this principle in real-world contexts.
>     a. Apply properties of operations as strategies to add and subtract rational numbers. 


### Algebraic Properties

Since the rational numbers were constructed from the integers, they maintain many of the corresponding properties of the integers. So if $\frac{a}{b}$, $\frac{m}{n}$, and $\frac{p}{q}$ are elements of $\mathbb{Q}$. 

```{r ratprops, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
addition <- c("$\\frac{a}{b}+\\frac{m}{n} \\in \\mathbb{Q}$", " $\\frac{a}{b}+\\left(\\frac{m}{n}+\\frac{p}{q}\\right) =\\left(\\frac{a}{b}+\\frac{m}{n}\\right)+\\frac{p}{q}$", "$\\frac{a}{b}+\\frac{m}{n}=\\frac{m}{n}+\\frac{a}{b}$", "$\\frac{a}{b}+0=\\frac{a}{b}$", "$\\frac{a}{b} \\cdot \\left(\\frac{m}{n}+\\frac{p}{q}\\right) = \\left(\\frac{a}{b}\\cdot \\frac{m}{n}\\right) + \\left(\\frac{a}{b} \\cdot \\frac{p}{q}\\right)$", "$\\frac{a}{b} + \\left(-\\frac{a}{b}\\right) =0$", "If $\\frac{a}{b},\\frac{m}{n}>0$, then $\\frac{a}{b}\\cdot \\frac{m}{n}>0$.", "$\\frac{a}{b} \\cdot \\frac{m}{n}=0$ if, and only if, $\\frac{a}{b}=0$, $\\frac{m}{n}=0$, or both")
multiplication <- c("$\\frac{a}{b}\\cdot \\frac{m}{n} \\in \\mathbb{Q}$", "$\\frac{a}{b}\\cdot \\left(\\frac{m}{n}\\cdot \\frac{p}{q}\\right) = \\left(\\frac{a}{b} \\cdot \\frac{m}{n}\\right) \\cdot \\frac{p}{q}$", "$\\frac{a}{b}\\cdot \\frac{m}{n} = \\frac{m}{n}\\cdot \\frac{a}{b}$", "$\\frac{a}{b} \\cdot 1 = \\frac{a}{b}$", "$\\left(\\frac{a}{b}+\\frac{m}{n}\\right) \\cdot \\frac{p}{q} = \\left(\\frac{a}{b}\\cdot \\frac{p}{q}\\right) + \\left(\\frac{m}{n}\\cdot \\frac{p}{q}\\right)$", "If $a,b\\neq 0$, $\\frac{a}{b} \\cdot \\left(\\frac{b}{a}\\right) =1$", "If $\\frac{a}{b},\\frac{m}{n}<0$, then $\\frac{a}{b}\\cdot \\frac{m}{n}>0$.", "")
property <- c("Closure", "Associative Property", "Commutative Property", "Identities", "Distributive Property", "Inverses", "", "No zero divisors")
intprops <- data.frame(addition, multiplication, property)
names(intprops) <- c('', '', 'Property')
knitr::kable(intprops, format='pandoc', caption = 'Properties of Rational Numbers')
```



### Properties of Exponents

Using our definition of exponents from Section \@ref(sec:Integers), we will expand the number system for the base from the integers to the rational numbers.

If $\displaystyle{\frac{p}{q}}$ (with $q\neq 0$) is a rational number, for any natural number $n$ we define $$\left(\frac{p}{q}\right)^n := \frac{p^n}{q^n}. $$ Such a definition makes sense because we know from Section \@ref(sec:Integers) that $p^n$ and $q^n$ are both well-defined integers, that $q^n\neq 0$, and the ratio of such integers in a rational number.

We can then extend the number system for the exponents from the natural numbers to the integers for $p,q\neq 0$ by defining
$$\left(\frac{p}{q}\right)^{-n} = \left(\frac{q}{p}\right)^n$$ for all $n\in \mathbb{N}$. In the case that $p=0$, this definition would not be well-defined and so we define $0^n=1$ for all $n\in \mathbb{Z}$.

>#### Related Content Standards {-}
>+ (8.EE.1) Know and apply the properties of integer exponents to generate equivalent numerical expressions.

With the extended definition of exponents we still have for all rational numbers, $a$, that $a^0=1$ and $a^1=a$. We also have from our definition that for $a\neq 0$ and $n\in \mathbb{Z}$, $$a^{-n}=\frac{1}{a^n}=\left(\frac{1}{a}\right)^n.$$

For $m,n \in \mathbb{N}$ and integers $p$ and $q$, with $q\neq 0$, 
$$\left(\frac{p}{q}\right)^m \cdot \left(\frac{p}{q}\right)^n = \left(\frac{p^m}{q^m} \right) \cdot \left(\frac{p^n}{q^n}\right) = \frac{p^m \cdot p^n}{q^m \cdot q^n} = \frac{p^{(m+n)}}{q^{(m+n)}} = \left(\frac{p}{q}\right)^{(m+n)}.$$
This can then be extended to $m,n\in \mathbb{Z}$ with

+ [Case 1:] If $p=0$, then $\frac{p}{q}=0$ and $$0^m\cdot 0^n = 0\cdot 0=0=0^{(m+n)}.$$
+ [Case 2:] If $p\neq 0$, we let $m,n\in \mathbb{Z}$. 
If $m$ and $n$ are both non-negative, then we have the property given above.
If $m$ and $n$ are both negative, then $$\left(\frac{p}{q}\right)^{m} \cdot \left(\frac{p}{q}\right)^{n} = 
\left(\frac{q}{p}\right)^{-m} \cdot \left(\frac{q}{p}\right)^{-n} =
\left(\frac{q}{p}\right)^{(-m-n)} = \left(\frac{p}{q}\right)^{m+n}.$$
If one is negative, $m$, and the other is non-negative, $n$, then 
$$ \left(\frac{p}{q}\right)^{m} \cdot \left(\frac{p}{q}\right)^{n} = \frac{p^m}{q^m} \cdot \frac{q^{-n}}{p^{-n}} = \frac{p^{m+n}}{q^{m+n}} =  \left(\frac{p}{q}\right)^{m+n}. $$

Therefore, we have that for all $a\in \mathbb{Q}$ and for all $m,n\in \mathbb{Z}$ that $a^m\cdot a^n = a^{m+n}$.

One can make similar arguments to extend the remainder of Theorem \ref{thm:exponents-integers} so that $(ab)^n=a^n\cdot b^n$ for each $n\in \mathbb{Z}$ and  $ (a^m)^n = a^{mn}$ for each $m,n\in \mathbb{Z}$.

Since the rational numbers include values between integers, it makes sense to explore the relationships between exponents and the order of the rational numbers given by inequalities. If $a\in \mathbb{Q}$ with $a>1$, then we can write $a$ as $\frac{p}{q}$ for some integers $p$ and $q$ with $0<q<p$, and from Theorem \ref{thm:exponents-integers} we have  $0<q^n<p^n$ for any integer $n\geq 1$. So we can conclude that $a^n>1$. If $n$ is a negative integer, then $0<q^{-n}<p^{-n}$ and so $a^{-n} >1$, and hence $a^n<1$.

If $a$ and $b$ are rational numbers with $0<a<b$, then $\frac{a}{b}\in \mathbb{Q}$ with $0<\frac{a}{b}<1$ so $0<\left(\frac{a}{b}\right)^n<1$ and $0<a^m<b^m$.

We combine all of the results into the following theorem.

```{theorem, exponents-rationals}
Let $a$ and $b$ be rational numbers.

1. $a^0=1$ and $a^1=a$

1. If $a>1$, then $a^n >1$ for all integers $n>0$, and if $a<1$, then $0<a^n<1$ for all integers $n<0$

1. $a^{-n} = \frac{1}{a^n}$, $a^m\cdot a^n = a^{m+n}$, $(ab)^n=a^n\cdot b^n$, and $(a^m)^n = a^{mn}$ for each $m,n\in \mathbb{Z}$

1. If $0<a<b$ and $m\in \mathbb{N}$, then $0<a^m<b^m$.

1. If $a>1$ and $m<n$, then $a^m<a^n$.

```


### Exercises

1. Prove that there exists a rational number between any two rational numbers. (If $a,b\in \mathbb{Q}$ with $a<b$, then there exists a $c\in \mathbb{Q}$ such that $a<c<b$.)

1. Let $x,y\in \mathbb{Q}$ such that $0<x<y$. Prove that $0<\frac{1}{y} < \frac{1}{x}$.

1. Look graphically at $\mathbb{Z}\times (\mathbb{Z}-\{0\})$, which is a different representation of $\widehat{\mathbb{Q}}$ defined at the beginning of this section.
    a. How would you graphically describe the equivalence classes of the relation $$ (a,b) \sim (c,d) \Leftrightarrow ad=bc?$$
    a. How does this interact with the standards in Grades 7 and 8 regarding ratio and direct proportions?

1. Compare the definition of order of rational numbers in this section with other techniques of determining the order of rational numbers (i.e. cross-multiplying, finding lowest common denominator, `butterfly method', etc.).
    a. What types of (mis)conceptions arise with each of these methods?
    a. Is there an order to presenting these various methods for determining the order of rational numbers that improves student understanding?
    
## Representations of Rational Numbers

Each rational number can be represented in several different ways, with the most prominent being as a mixed number, a fraction, a decimal, or a percent. In this section we will discuss how to convert between these different representations and how to use other physical and graphical representations to understand and use rational numbers.


### Decimal Representations

It is sometimes helpful to write rational numbers in a decimal representation, rather than fraction notation. This is particularly true in the comparison of two rational numbers. One process for finding the decimal representation from a fraction is long division, but it is often simpler to use technology to find the decimal representation. (It also greatly reduces the amount of error involved.) However, it is important to have a deeper understanding of the decimal representation of rational numbers to verify that the result obtained from a calculator or computer is correct. For instance, if one uses a calculator to find $\frac{8}{17}$, a calculator with an 8 digit display will give the answer $0.4705882$, and a calculator with a 16 digit display will give the answer $0.470588235294118$. There is no apparent repetition in this decimal, leading to the question of if it repeats or not. It is the repeated use of the division algorithm that allows us to determine why a decimal representation of a rational number always repeats.



>#### Related Content Standards {-}
>+ (7.NS.2) Apply and extend previous understandings of multiplication and division and of fractions to multiply and divide rational numbers.
>     d. Convert a rational number to a decimal using long division; know that the decimal form of a rational number terminates in 0s or eventually repeats.


We will state the division algorithm for integers here, but the proof is in Section \@ref(sec:Integer-Ring).


```{theorem, name="The Division Algorithm for Integers"}
If $a$ and $b$ are integers with $b>0$, then there are unique integers $q$ and $r$ such that $$a=bq+r, \quad \mbox{with } 0\leq r <b.$$
```

We will explain the process of long division and its relationship to the division algorithm to determine the decimal representation of rational numbers using an example.

In order to find the decimal expansion of $\frac{7}{12}$, we will first rewrite the rational expression as $\frac{1}{10} \cdot \frac{70}{12}$. We then note that $70=12\cdot 5 + 10$ in the division algorithm. Thus, 
$$\frac{7}{12} = \frac{1}{10} \cdot \frac{70}{12} = \frac{1}{10} \cdot \left( 5 + \frac{10}{12}\right)= \frac{5}{10} + \frac{1}{10} \cdot \frac{10}{12}.$$ We continue this process with $\frac{10}{12}$ so that we have $$\frac{7}{12}= \frac{5}{10} + \frac{1}{10^2} \left(\frac{100}{12}\right) = \frac{5}{10} + \frac{1}{10^2} \left(8 + \frac{4}{12}\right)= \frac{5}{10} + \frac{8}{10^2} + \frac{1}{10^2} \cdot \frac{4}{12}.$$ We continue the process, so that $$\frac{7}{12} = \frac{5}{10} + \frac{8}{10^2} + \frac{1}{10^3} \cdot \frac{40}{12} = \frac{5}{10} + \frac{8}{10^2} + \frac{3}{10^3} + \frac{1}{10^3} \cdot \frac{4}{12}.$$ We notice that this remainder of $4$ again tells us that we will get the same quotient again and again. Therefore, we see that
$\frac{7}{12} = 0.58\overline{3}$.

As we can see with this process, a decimal will start repeating when a remainder in the division algorithm is repeated. Since the only possible remainders of a quotient by $12$ are in the set $\{0,1,2,3,4,5,6,7,8,9,10,11\}$, we see that any decimal representation of a fraction whose denominator is $12$ will repeat after at most 11 places.

Similarly, the decimal representation for any rational number will eventually repeat. This repetition may be of the digit $0$, in which case we will call it a **finite decimal expansion**, or it may be a different number of digits, in which we will call it a **repeating decimal expansion**. In the case of $\frac{7}{12}$, we see that the repeating happens after a certain number of digits. These types of decimal representations will be called **delayed repeating decimal expansions**. One can determine the type of decimal expansion based on the prime factorization of the denominator, once the fraction is written in its simplified form.


>#### Related Content Standards {-}
>+ (8.NS.1) Know that numbers that are not rational are called irrational. Understand informally that every number has a decimal expansion; for rational numbers show that the decimal expansion repeats eventually, and convert a decimal expansion which repeats eventually into a rational number.


Now that we know that all rational numbers can be represented by decimal expansions that are finite, repeating, or delayed repeating, we will study the process of starting with a decimal representation and determining a fraction representation. We already know that if the decimal representation does not repeat that it does not correspond to a rational number. 

If the number has a finite decimal expansion, then it can be converted to a fraction using the base ten representation. For instance the number $3627.854$ can be rewritten as $\frac{3627854}{1000}$.

Let's look at a decimal expansion that is repeating like $0.\overline{35682}$. This could be written in the form $$0.\overline{35682}= \frac{10^5}{10^5} 0.\overline{35682} = \frac{1}{10^5} \cdot 35682.\overline{35682} = \frac{1}{10^5} \left( 35682 + 0.\overline{35682}\right).$$ We can rearrange this equation so that 
$$10^5 \cdot 0.\overline{35682} - 0.\overline{35682} = 35682$$ or equivalently that 
$$0.\overline{35682} = \frac{35682}{10^5-1}.$$ This same process can be used to find the fractional representation of any rational number with a repeating decimal expansion.

If a rational number has a delayed repeating decimal expansion, then one can rewrite the number in a way to generate a repeating decimal. For instance, 
$$324.51\overline{89} = \frac{10^2}{10^2} \cdot 324.51\overline{89} = \frac{1}{10^2} \left( 32451.\overline{89}\right) = \frac{1}{10^2} \left( 32451+\frac{89}{10^2-1} \right) = \frac{(32451\cdot 99)+89}{9900}.$$

### Physical and Graphical Representations

>#### Related Content Standards {-}
>+ (6.NS.1) Interpret and compute quotients of fractions, and solve word problems involving division of fractions by fractions, e.g., by using visual fraction models and equations to represent the problem. For example, create a story context for $\frac{2}{3} \div \frac{3}{4}$ and use a visual fraction model to show the quotient; use the relationship between multiplication and division to explain that $\frac{2}{3} \div \frac{3}{4} = \frac{8}{9}$ because $\frac{3}{4}$ of $\frac{8}{9}$ is $\frac{2}{3}$. (In general, $\frac{a}{b} \div \frac{c}{d} = \frac{ad}{bc}$.) How much chocolate will each person get if 3 people share $\frac{1}{2}$ lb of chocolate equally? How many $\frac{3}{4}$-cup servings are in $\frac{2}{3}$ of a cup of yogurt? How wide is a rectangular strip of land with length $\frac{3}{4}$ mi and area $\frac{1}{2}$ square mi?       


Consider the following problem from the Singapore Mathematics Curriculum:

> Jenny, Bob, and Paul shared a sum of money. Jenny received $\frac{2}{5}$ of the money, Bob received $\frac{1}{4}$ of the money and Paul received the rest of the money. If Paul received $\$1.50$ more than Bob, how much more money did Jenny receive than Paul.

While this is not a modeling problem, in that this type of situation would never be described in this way, it does help generate fluency in working with different representations of rational numbers. We will walk through one possible solution that does not use variables, but does build the conceptual understanding necessary for successful work in algebra.

The first thing that we notice is that we need a way to represent $\frac{2}{5}$ and $\frac{1}{4}$ of the same whole. As such, we need our whole item to be broken into $20$ equal sized pieces.

```{r, echo=FALSE, fig.align='center', out.width='60%', fig.cap=''}
knitr::include_graphics(
  'tikz/singapore-fraction1.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

We then note that Jenny received $\frac{2}{5}$ of the money, which is equivalent to $\frac{8}{20}$ of the money. So we will mark $8$ of the pieces with a J. Similarly, Bob received $\frac{1}{4}$ of the money and so we will mark $5$ pieces with a B.

```{r, echo=FALSE, fig.align='center', out.width='60%', fig.cap=''}
knitr::include_graphics(
  'tikz/singapore-fraction2.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

We then see that Paul received the equivalent of 7 blocks, while Bob received 5 blocks. Since Paul received $\$1.50$ more than Bob we know that 2 blocks is equivalent to $\$1.50$, making each block represent $\$0.75$. Since Jenny received the equivalent of 1 block more than Paul we see that she received $\$0.75$ more than Paul.

There are many other ways to solve this problem that also develop numerical fluency with rational numbers, and the more ways that students can represent these numbers and move between the representations, the more fluent they will be with the rational number system.

#### Fraction Strips

One common tool used to help students determine equivalent fraction representations is through the use of fraction strips. These are pieces of paper (all the same size), but marked off in different numbers of parts. Then by setting the pieces of paper next to each other, students are able to notice equivalency of fractions. For instance, in the fraction strips below one can see that $\frac{1}{2}$ is equivalent to $\frac{2}{4}$.

```{r, echo=FALSE, fig.align='center', out.width='60%', fig.cap=''}
knitr::include_graphics(
  'tikz/fraction-strip1.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

One can also use fraction strips to represent the operation of division of fractions. For instance if we have the two fraction strips below we can use them to determine that $\frac{1}{2} \div \frac{1}{6} = 3$ because $3$ of the $\frac{1}{6}$ regions can fit inside of the $\frac{1}{2}$ region.

```{r, echo=FALSE, fig.align='center', out.width='60%', fig.cap=''}
knitr::include_graphics(
  'tikz/fraction-strip1.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

#### Pattern Blocks

Another tool to help students work with fractions are pattern blocks. The pattern blocks most useful for many of the fractions that students use are the hexagon, triangle, trapezoid, and rhombus below.

```{r, echo=FALSE, fig.align='center', out.width='60%', fig.cap=''}
knitr::include_graphics(
  'tikz/pattern-blocks.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

Students can see that $6$ triangles, $2$ trapezoids, or $3$ rhombi fit inside of the hexagon. So we can represent $\frac{2}{3}$ as two rhombi, with one hexagon representing the whole. One can then see that fractions whose denominator is $2$, $3$, or $6$ are easily represented using one hexagon as the whole. However, if we allow two hexagons to represent the whole, we can represent fractions with denominators of $2$, $3$, $4$, $6$, or $12$. One can also think about $\frac{1}{2} \div \frac{1}{3}$ as how many rhombi fit inside of one trapezoid. 



#### Area Model

The area model described in Section \@ref(sec:integer-representation) can help students to visualize multiplication of fractions. For instance, one can find the product of $\frac{2}{5}$ and $\frac{2}{3}$ using the following diagram. One can then see that $\frac{4}{15}$ of the squares are shaded. So $\frac{2}{5} \times \frac{2}{3} = \frac{4}{15}$.


```{r, echo=FALSE, fig.align='center', out.width='60%', fig.cap=''}
knitr::include_graphics(
  'tikz/area-model-fractions.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

We are also able to build upon the area model to visualize the distribution property involved in the multiplication of mixed numbers. If we want to find $23\frac{2}{3} \times 32 \frac{5}{7}$, then we have to remember the convention that $23\frac{2}{3} = 23 + \frac{2}{3}$ and $32\frac{5}{7}= 32 + \frac{5}{7}$. (Note how confusing this convention can be for students, particularly since in algebra two numbers next to each other represent multiplication instead of addition.) We can then allow for an abstract area representation (the sizes are not proportional) to more easily use the distributive property.

```{r, echo=FALSE, fig.align='center', out.width='50%', fig.cap=''}
knitr::include_graphics(
  'tikz/area-model-Mixednumbers.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

There are many other tools that can help students understand rational numbers and their operations, but we will leave those for more pedagogically oriented sources.

#### Number Line and Plane

It is also important for students to use rational numbers on a number line (and the associated plane). In particular, the number line can help students understand inequalities, relationships between different numbers, and the absolute value of a number. Some key factors to keep in mind are to not only have fractions between $0$ and $1$, or only mixed numbers outside of that interval. Students need to use a variety of different numbers and representations together to gain fluency.

```{r, echo=FALSE, fig.align='center', out.width='60%', fig.cap=''}
knitr::include_graphics(
  'tikz/number-line-rationals.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

>#### Related Content Standards {-}
>+ (6.NS.7) Understand ordering and absolute value of rational numbers.
>     a. Interpret statements of inequality as statements about the relative position of two numbers on a number line diagram.
>     a. Write, interpret, and explain statements of order for rational numbers in real-world contexts.
>     a. Understand the absolute value of a rational number as its distance from 0 on the number line; interpret absolute value as magnitude for a positive or negative quantity in a real-world situation.
>     a. Distinguish comparisons of absolute value from statements about order.



### Exercises
    
1. Explain, in a way that a middle school student can understand, how to determine the type of decimal expansion for any given fraction representation of a rational number and why it is that way. (i.e. when is the decimal expansion finite? when does it repeat? when is it delayed periodic? what do each of these mean?)
    
1. For each of the following, give an example of a rational number in the form of $\frac{p}{q}$ with the appropriate decimal representation and explain why it works.
    a. has a terminating decimal with 4 decimal places
    a. has a non-zero integer part and a delay of 3 decimal places before a periodic part with period of 3 digits.
    a. has a periodic decimal representation with 5 digits.
        
1. For each of the following, give the representation of the rational number in the form of $\frac{p}{q}$, with $p$ and $q$ relatively prime (no common factors): 
    a. $0.156$
    a. $0.23\overline{4}$
    a. $23.\overline{15}$
  	a. $3.14159$
    a. $2.5\overline{9}$
    a. $2016$
    
  1. Write an algorithm in the programming language of your choice to input a decimal representation of a rational number and output a fraction representation.
    
1. Classify each of the following real numbers as 
      
      + a terminating decimal, 
      + simple periodic decimal (note its period), 
      + delayed periodic decimal (note its delay and period), or
      + non-periodic decimal

      a. $\frac{1}{6}$
      a. $\frac{1}{2}$
      a. $\frac{2}{14}$
      a. $\frac{3}{20}$
      a. $\frac{4}{63}$
      a. $\pi$
      a. $\sqrt{2}$
      a. $\sqrt{5}$
      a. $\frac{18}{40}$
      a. $0.75$
      a. $0.83\overline{45}$
      

## Real Numbers {#sec:reals}

One of the properties of the integers that does not extend to the rational numbers is the least-upper-bound property (\@ref(def:well-ordering)). In order to prove this, we first need to prove a series of small lemmas (a smaller version of a theorem that is used to prove a larger theorem).

```{lemma}
If $n$ is an integer, then $n$ can be written as $2k$ for some integer $k$ or $2j+1$ for some integer $j$, but cannot be written both ways. (In other words, an integer is even or odd, but not both.)
```


```{proof}
We will use the technique of a proof by contradiction and assume that there is an integer $n$ such that $n=2k$ and $n=2j+1$ for some integers $j$ and $k$. Then $2j=2k+1$ and so the equivalent equation $2(j-k)=1$ is true. However, since $2$ does not have a multiplicative inverse in the integers, $j-k$ cannot be an integer. This means that our assumption is false and our lemma is proven.
```


```{lemma}
If $n$ is an integer such that $n^2$ is even, then $n$ is even.
```

```{proof}
We will prove the contrapositive of the statement, which is logically equivalent to the original statement. Let $n$ be an odd integer, i.e. $n=2j+1$ for some integer $j$. Then $n^2=(2j+1)^2=4j^2+4j+1=2(2j^2+2j)+1$ and so $n^2$ is odd. 
```

```{lemma}
There does not exist a rational number whose square is $2$.
```

```{proof}
We will again use the technique of a proof by contradiction. Assume that there is a rational number whose square is $2$. We will denote the rational number by $\frac{a}{b}$ where $a$ and $b$ are not both even. (This can be accomplished by finding an equivalent representation without 2 as a common factor.) Since its square is $2$, we have that $\left(\frac{a}{b}\right)^2 =2$ and equivalently that $a^2 = 2 b^2$. Since $a^2$ is even, the previous lemma tells us that $a$ is even. So there exists an integer $k$ such that $a=2k$ and so $a^2=4k^2$. We then have that $4k^2=2b^2$ and so $b^2=2k^2$. This means that $b^2$ is even and consequently that $b$ is even. This contradicts our original assumption and so no such rational number exits.
```

We will now look at the set 
$$S = \left\{ \frac{a}{b}\in \mathbb{Q} \middle \vert \frac{a}{b} >0 \: \mbox{ and } \: \left(\frac{a}{b}\right)^2 \geq 2\right\}$$ which is bounded below in its definition and does not have a least element. Therefore, the rational numbers do not satisfy the least-upper-bound property.

Since the rational numbers do not satisfy the least-upper-bound property, the rational numbers have "holes" that need to be filled, and so are not complete. Building upon the work of @bertrand1840 and in a parallel vein to @Heine1872, @Dedekind1872 developed a construction of the real numbers from the rational numbers using what are now referred to as Dedekind cuts using equivalence classes of such cuts. Another method of construction of the real numbers from the rational numbers follows the path of @Cauchy1821 and defines the real numbers as equivalence classes of Cauchy sequences of rational numbers. As each of these methods requires the development of a great deal of mathematical machinery, we will refer to other textbooks in analysis for the details. With @Rudin1976 providing an excellent treatment of Dedekind cuts in the Appendix of Chapter 1. Chapter I of @Thurston1956 provides an detailed treatment of the construction of the Cauchy numbers to generate the real numbers through equivalence classes.

Because the real numbers form a complete number system in that it satisfies the least-upper-bound property, for any bounded set of real numbers, $S$, we can define $\sup S$ to the be the least element of the set $\{ x \in \mathbb{R} \vert x\geq a, \forall a\in S\}$ and $\inf S$ is the greatest element of the set $\{x\in \mathbb{R} \vert x\leq a, \forall a\in S\}$. These two attributes of a set play an important role in the development of calculus.

Since $\sqrt{2}$ is not rational, but we know it is a real number because of the completeness of the reals, and so there are real numbers that are not rational. We define these numbers to be **irrational**.

```{definition}
A real number, $a$, is called rational if there exist integers $p$ and $q$ such that $a=\frac{p}{q}$. A real number that is not rational is called irrational.
```

>#### Related Content Standards {-}
>+ (HSN.RN.3) Explain why the sum or product of two rational numbers is rational; that the sum of a rational number and an irrational number is irrational; and that the product of a non-zero rational number and an irrational number is irrational.


We have already shown that the sum and product of two rational numbers are rational, but we would like to explore how rational and irrational numbers interact with each other.

```{theorem}
Let $u\in \mathbb{Q}$ and $v\in (\mathbb{R}-\mathbb{Q})$. Then $u+v\notin \mathbb{Q}$ and if $u\neq 0$, $uv\notin \mathbb{Q}$.
```

```{proof}
When trying to prove that something does not satisfy a certain property, then it is best to use a proof by contradiction or to prove the contrapositive.

Let $u\in \mathbb{Q}$ and so there exist integers $p$ and $q$ such that $u=\frac{p}{q}$. Let $v\in (\mathbb{R}-\mathbb{Q})$. We will now prove that $u+v$ is not rational using a proof by contradiction. If $u+v$ is rational, then there exist integers $m$ and $n$ such that $u+v=\frac{m}{n}$. This means that $$v= \frac{m}{n}-u = \frac{m}{n}-\frac{p}{q} = \frac{mq-np}{nq}$$ making it a rational number, contradicting the assumption that it is irrational.
    
The proof that $uv$ is irrational is very similar.
```

One of the important properties to notice about the rational and irrational numbers is that they are dense in the real numbers in that between any two real numbers one can find both a rational and an irrational number.

```{theorem, name="Density of the Rational Numbers"}
Let $x,y\in \mathbb{R}$ be any two real numbers where $x<y$. Then there exists a rational number $q\in \mathbb{Q}$ such that $x<q<y$.
```

```{proof}
Since $x<y$, we know that $y-x>0$. So there exists a positive natural number $n\in \mathbb{N}$ such that $0<\frac{1}{n}<y-x$. Equivalently, $0<1<ny-nx$ and $nx+1<ny$. From properties of the integers we know that there exists an integer $N\in \mathbb{Z}$ such that $N-1\leq nx<N$, or equivalently $N\leq nx+1 < N+1$. Therefore, $$nx<N\leq nx+1 < ny$$ and so $x<\frac{N}{n} <y$. So $q=\frac{N}{n}$ is a rational number such that $x<q<y$.
```


```{theorem, name="Density of the Irrational Numbers"}
Let $x,y\in \mathbb{R}$ be any two real numbers where $x<y$. Then there exists an irrational number $v\in \mathbb{R}\setminus\mathbb{Q}$ such that $x<v<y$.
```

```{proof}
Since $x$ and $y$ are real numbers, $\frac{x}{\sqrt{2}}$ and $\frac{y}{\sqrt{2}}$ are also real numbers such that $\frac{x}{\sqrt{2}}<\frac{y}{\sqrt{2}}$. By the density of the rational number, there exists a $q\in \mathbb{Q}$ such that $$\frac{x}{\sqrt{2}} < q < \frac{y}{\sqrt{2}},$$ or equivalently $x<q\sqrt{2}<y$. Since $q\in \mathbb{Q}$ and $\sqrt{2}$ is irrational, $q\sqrt{2}$ is irrational and so if $v=q\sqrt{2}$, $v\in \mathbb{R}\setminus\mathbb{Q}$ such that $x<v<y$.
```


Since the operations of addition and multiplication on the real numbers are associative and commutative, for each natural number $n$ we can define $a^n$ recursively and we can prove the following properties using the same techniques as in Section \@ref(sec:Integers), and Section \@ref(sec:rationals) using limit ideas from the definition of the real numbers.

```{theorem, exponents-real-integers}
Let $a,b \in \mathbb{R}$ and let $m,n\in \mathbb{Z}$.

+ $a^0=1$ and $a^1=a$

+ If $a>1$ and $n>0$, then $a^n >1$. If $a>1$ and $n<0$, then $0<a^n<1$.

+ $a^{-n} = \frac{1}{a^n}$

+ $a^m\cdot a^n = a^{m+n}$, $(ab)^n=a^n\cdot b^n$, and $ (a^m)^n = a^{mn}$

+ If $0<a<b$ and $m\in \mathbb{N}$, then $0<a^m<b^m$.

+ If $a>1$ and $m<n$, then $a^m<a^n$. 

```

With these properties, we are able to define define algebraic numbers, which is a number system between the rational numbers and the real numbers.

```{definition}
A real number, $a$, is called \textbf{algebraic} if $$n_0 + n_1 a + n_2 a^2 + \cdots + n_m a^m =0$$ for some integers $n_0, n_1, n_2, \ldots, n_m$.
```

For all rational numbers in the form $\frac{p}{q}$, we know that $$q \cdot \left(\frac{p}{q}\right)^1 + (-p) = 0$$ and so all rational numbers are algebraic numbers. We can also see that $\sqrt{2}$ is an algebraic number since $(\sqrt{2})^2-2=0$. We say that the algebraic numbers are a number system because they include the additive identity, $0$; the multiplicative identity, $1$; and are closed under addition and multiplication.


We will prove in Section \@ref(sec:cardinality) that there are an infinite number of real numbers that are not algebraic. We will call any real number that is not algebraic a **transcendental** number. Two such examples of transcendental numbers are $e$ and $\pi$. 

In the early 1930's, @Gelfond and @Schneider independently solved Hilbert's seventh problem from his 1900 address to the International Congress of Mathematics showing that if $a$ and $b$ are algebraic numbers with $a\neq 0,1$ and $b$ irrational that $a^b$ is a transcendental number. Since we have such a limited knowledge of these transcendental numbers, it is evident that the continuum of the real numbers is a very deep and complicated concept that we are only beginning to understand.

We include a graphical representation of how these different sets of numbers fit together. All of these sets of numbers are subsets of the complex numbers, and a set above another set represents an inclusion relationship.

```{r, echo=FALSE, fig.align='center', out.width='60%', fig.cap=''}
knitr::include_graphics(
  'tikz/NumberTree.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```






### Properties of Exponents {#subsec:real-exponents} 

Let $a>0$ be a real number. For each integer $n>0$, we define $$S_n= \left\{ x\in \mathbb{Q}\middle \vert x>0 \mbox{ and } x^n>a\right\}$$ and so we can define $$\sqrt[n]{a}=a^{\frac{1}{n}} := \inf S_n.$$ 


>#### Related Content Standards {-}
>+ (8.EE.2) Use square root and cube root symbols to represent solutions to equations of the form $x^2=p$ and $x^3=p$, where $p$ is a positive rational number. Evaluate square roots of small perfect squares and cube roots of small perfect cubes. Know that $\sqrt{2}$ is irrational.        


If $a>0$ is a real number, $m,n,p,q$ are integers, $n>0$, and $q>0$ such that $\frac{m}{n}=\frac{p}{q}=r$, we have that $mq=pn$. So $$\left(\left(a^{\frac{1}{n}}\right)^m\right)^{nq} = a^{mq} = a^{pn} = \left( \left( a^{\frac{1}{q}} \right)^p \right)^{nq}.$$ This means that we can define $a^r:=\left(a^m\right)^{\frac{1}{n}}$.

>#### Related Content Standards {-}
>+ (HSN.RN.1) Explain how the definition of the meaning of rational exponents follows from extending the properties of integer exponents to those values, allowing for a notation for radicals in terms of rational exponents.
>+ (HSN.RN.2) Rewrite expressions involving radicals and rational exponents using the properties of exponents.        


We can now prove various exponential properties of real numbers with rational exponents. In order to prove a generalization of Theorem \@ref(thm:exponents-real-integers) to rational exponents we will only need to prove the corresponding properties for rational numbers of the form $\frac{1}{n}$ and it directly follows from $(a^n=b^n) \Leftrightarrow a=b$ for real numbers $a$ and $b$ and positive integers $n$. As such, we will omit the majority of these proofs from this text. However, will include one lemma and proof in order to understand the flavor of the proofs.

```{lemma}
Let $a\in \mathbb{R}^+$. For all $r,s\in \mathbb{Q}$, $a^{r+s}=a^r\cdot a^s$.
```

```{proof}
Let $a\in \mathbb{R}^+$ and let $r,s\in \mathbb{Q}$ such that $r=\frac{m}{n}$ and $s=\frac{p}{q}$ for some integers $m$, $n$, $p$, and $q$.
    
Since $q\neq 0$ and $n\neq 0$,
$$\left(a^{r+s}\right)^{qn} = \left( a^{\frac{mq+pn}{qn}}\right)^{qn} = a^{(mq+pn)}$$ and
$$\left(a^r\cdot a^s\right)^{qn} = \left(a^{\frac{mqn}{n}}\right)\cdot \left(a^{\frac{pqn}{q}}\right) = a^{(mq+pn)}.$$ Therefore,  $a^{r+s}=a^r\cdot a^s$.
```
            

For a real number $y$, define $$S_y= \left\{ a^r \middle \vert r \leq y\right\}.$$ Furthermore, define $$a^y = \sup S_y.$$ Using arguments from analysis and the properties of limits, one can extend the previous properties of exponents to real valued exponents as given in Theorem \@ref(thm:exponents-reals). Due to the mathematical machinery necessary for these proofs, we will refer the reader to an analysis textbook.

```{theorem, exponents-reals}
Let $a$ and $b$ be positive real numbers and let $x,y\in \mathbb{R}$.

+ $a^0=1$ and $a^1=a$

+ $a^x >1$ for all $x>0$, and $0<a^x<1$ for all $x<0$

+ $a^{-x} = \frac{1}{a^x}$

+ $a^x\cdot a^y = a^{x+y}$, $(ab)^x=a^x\cdot b^x$, and $ (a^x)^y = a^{xy}$

+ If $0<a<b$ and $x>0$, then $a^x<b^x$.

+ If $a>1$ and $x<y$, then $a^x<a^y$.

```


### Exercises

1. If $n=2$, $a^\frac{1}{2}$ is often written as $\sqrt{2}$, but for $n=\{3,4,5, \ldots\}$, $a^{\frac{1}{n}}$ is written as $\sqrt[n]{a}$. This difference in notation can cause many problems for students. What are some other notation differences in the secondary curriculum that have similar situations?

1. What can be said about the sum and product of two irrational numbers?    


1. Suppose that $a\in \mathbb{R}$ with $a>0$. If $n$ is an even positive integer, show that there are exactly two real numbers that are solutions to the equation $x^n=a$.
    
1. Show that $\sqrt{2}+\sqrt{3}$ is an algebraic number by finding a polynomial for which it is a zero.
    
1. Write out a detailed proof that $\sqrt{3}$ is irrational using the techniques of the proof that $\sqrt{2}$ is irrational. For which numbers can this process be generalized?
 


## Complex Numbers {#sec:complex}


Unlike all of the previous number system constructions, in order to construct the complex numbers, $\mathbb{C}$, we do not have to work through equivalence classes. Instead, we look at the set 
$$\mathbb{R} \times \mathbb{R} = \left\{ (x,y) \middle \vert x,y \in \mathbb{R}\right\}$$ and define the addition and multiplication operations as 
$$(a,b) + (c,d) = (a+c, b+d) \quad \mbox{ and } \quad (a,b) \cdot (c,d) = (ac-bd, ad+bc).$$

We can see that elements of the form $(a,0)$, $\mathbb{R}\times \{0\}$, act just like the real numbers in that $$(a,0)+(b,0)=(a+b,0) \quad \mbox{ and } \quad (a,0)\cdot (b,0) = (ab,0)$$ and so we can think of these as being the real numbers inside of the complex numbers.

Elements of the form $(0,a)$, $\{0\}\times \mathbb{R}$, operate very differently, particularly with regard to multiplication, since 
$$(0,a)+(0,b)=(0,a+b) \quad \mbox{ and }  (0,a) \cdot (0,b) = (-ab,0).$$

### Rectangular Representation

If we define scalar multiplication on this set by $k \cdot (a,b) = (ka,kb)$, we see that the elements of the set $\mathbb{R} \times \mathbb{R}$ operate as a vector space in that for all $(a,b) \in \mathbb{R}\times \mathbb{R}$, $(a,b) = a \cdot (1,0) + b \cdot (0,1)$. If we then replace $(1,0)$ with its corresponding element of $1$ in $\mathbb{R}$ and label $0,1$ with a new symbol that we call $i$, we have the complex numbers, $$\mathbb{C} = \left\{ a+bi\middle \vert a, b\in \mathbb{R}\right\}$$ with addition and multiplication defined by 
$$(a+bi) + (c+di) = (a+c) + (b+d)i \quad \mbox{ and } \quad (a+bi) \cdot (c+di) = (ac-bd) + (ad+bc)i.$$ 

>#### Related Content Standards {-}
>+ (HSN.CN.1) Know there is a complex number $i$ such that $i^2 = -1$, and every complex number has the form $a + bi$ with $a$ and $b$ real.
>+ (HSN.CN.2) Use the relation $i^2 = -1$ and the commutative, associative, and distributive properties to add, subtract, and multiply complex numbers.        

In the eighth chapter of the greatest work of mathematics in the 18^th^ century, possibly even in the modern era, Leonard Euler [-@Euler] described a geometric, algebraic, and analytic connection between ways of representing complex numbers that enabled many of the mathematical and scientific discoveries of the past 350 years. What Euler recognized is that in addition to representing complex numbers as $a+bi$, one can describe the complex numbers geometrically as points on the plane where the real part of the complex number provides the horizontal component and the imaginary part of the complex number provides the vertical component.

```{r, echo=FALSE, fig.align='center', out.width='60%', fig.cap=''}
knitr::include_graphics(
  'tikz/complex-rectangular.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```


### Complex Conjugation and Modulus


>#### Related Content Standards {-}
>+ (HSN.CN.4) Represent complex numbers on the complex plane in rectangular and polar form (including real and imaginary numbers), and explain why the rectangular and polar forms of a given complex number represent the same number.



Since points on the plane can also be described by the distance from the origin and the angle made with the positive horizontal axis, complex numbers can also be described by a magnitude and an angle. When we combine that with knowledge of trigonometric functions we have that every complex number can be written in the form 
$$a+bi = r(\cos(\theta) + i \sin(\theta))$$

```{r, echo=FALSE, fig.align='center', out.width='60%', fig.cap=''}
knitr::include_graphics(
  'tikz/complex-polar.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```


If we look at the correspondence of $a=r\cos(\theta)$ and $b=r\sin(\theta)$, we can derive formulas to solve for $r$ and $\theta$ in terms of $a$ and $b$, 
$$r= \sqrt{a^2+b^2} \quad \mbox{ and } \quad \theta = \begin{cases} \arctan\left(\frac{b}{a}\right) & \mbox{ if } a\geq 0 \\
\arctan\left(\frac{b}{a}\right)+\pi & \mbox{ if } a<0\\
\end{cases}.$$ 

In addition to this polar representation of a complex number, Euler defined the **complex conjugate** of a number $z=a+bi$ to be $\overline{z}=a-bi$. Geometrically, the complex conjugate of a number is its reflection across the horizontal axis. We also see that if we multiply a complex number $z=a+bi$ by its complex conjugate we have $(z\overline{z}) = a^2+b^2=r^2$, which is the square of the **modulus**, $|z|^2$, or distance from $0$.

Notice that the modulus of a complex number is a real number. This makes it easier to divide by complex numbers. Using the complex conjugate we can write 
$$\frac{z}{w} = \frac{z}{w} \cdot \frac{\overline{w}}{\overline{w}} = \frac{z\overline{w}}{|w|^2} \:  \mbox{ or } \: \frac{a+bi}{c+di} = \frac{(a+bi)(c-di)}{c^2+d^2} = \left(\frac{ac+bd}{c^2+d^2}\right) + \left(\frac{bc-ad}{c^2+d^2}\right) i$$ for complex numbers $z=a+bi$ and $w=c+di$.

This modulus of a complex number generates a distance between two complex numbers, $|z_2-z_1|$, in a similar manner to absolute value measuring the distance between two real numbers, $|x_2-x_1|$.

>#### Related Content Standards {-}
>+ (HSN.CN.3) Find the conjugate of a complex number; use conjugates to find moduli and quotients of complex numbers.
>+ (HSN.CN.5) Represent addition, subtraction, multiplication, and conjugation of complex numbers geometrically on the complex plane; use properties of this representation for computation.
>+ (HSN.CN.6) Calculate the distance between numbers in the complex plane as the modulus of the difference, and the midpoint of a segment as the average of the numbers at its endpoints.



### Euler's Equation

Euler's [-@Euler] greatest discovery that one can use infinite series to find a strong connection between exponential functions and trigonometric functions  and that 
$$e^{x+yi} = e^x \left( \cos(y) + i \sin(y)\right) \quad \mbox{ Euler's Equation,}$$ where $e$ is the (Euler) constant $$e= 1 + \frac{1}{1} + \frac{1}{1\cdot 2} + \frac{1}{1\cdot 2 \cdot 3} + \frac{1}{1\cdot 2\cdot 3\cdot 4} + \cdots$$ 

This polar representation of complex numbers simplifies multiplication and exponents in that 
$$r_1 e^{i\alpha} \cdot r_2 e^{i\beta} = (r_1\cdot r_2) e^{i(\alpha+\beta)} \quad \mbox{ and } \left(r e^{i\theta}\right)^n = r^n e^{in\theta}.$$

Hence we can see that multiplication of complex numbers involves multiplying the moduli of the numbers and adding the angles, sometimes called an amplitwist.


We will explore these properties of the complex numbers in depth in Chapter \@ref(ch:transformations).



### Exercises

1. For each of the following complex numbers, plot the number on the complex plane and write the number in the form $a+bi$ and in the form $re^{i\theta}$. (You may leave $\theta$ as an arctangent of a number for angles that are not the standard angles.)
      
      a. $-14+14i$
      
      a. $\displaystyle{\frac{1+i}{1-\sqrt{3}i}}$
      
      a. $\displaystyle{3e^{\frac{-2\pi}{3}i}}$ 
      
      a. $1+3i$
      
      a. $\displaystyle{\overline{-2+5i}}$
    
1. Use the Maclaurin series representations of $e^x$, $\cos(x)$, and $\sin(x)$ to prove that $$e^{iy}=\cos(y) + i \sin(y).$$
    
1. Use Euler's equation to prove the angle addition identities for $\sin$ and $\cos$.
    
1. Use Euler's equation to prove the double angle identities for $\sin$ and $\cos$.
    
1. Use Euler's equation to show De Moivre's formula
$$\left(\cos(x) + i \sin(x) \right)^n = \cos(nx) + i\sin(nx) $$ is true and use the formula to find identities for $\cos(3x)$ and $\sin(3x)$ in terms of $\cos(x)$ and $\sin(x)$.
    




