# Integral Domains and Polynomials {#ch:rings}

## Properties of the Ring of the Integers {#sec:Integer-Ring}

In this section we spend time better understanding the integral domain of the integers. We will look at some of the divisibility properties that are the underpinnings for our standard long division algorithms and build up to the prime factorization of the integers.

The first key component of the integers that we will look at is the Division Algorithm.

```{theorem, division-algorithm, name="The Division Algorithm for Integers"}
If $a$ and $b$ are integers with $b>0$, then there are unique integers $q$ and $r$ such that $$a=bq+r, \quad \mbox{with } 0\leq r <b.$$
```

We call the integer $q$ in the division algorithm the quotient of $a$ and $b$ and we call $r$ the remainder of the quotient of $a$ and $b$.

>#### Related Content Standards {-}
>+ (4.NBT.6) Find whole-number quotients and remainders with up to four-digit dividends and one-digit divisors, using strategies based on place value, the properties of operations, and/or the relationship between multiplication and division. Illustrate and explain the calculation by using equations, rectangular arrays, and/or area models. 
>+ (5.NBT.6) Find whole-number quotients of whole numbers with up to four-digit dividends and two-digit divisors, using strategies based on place value, the properties of operations, and/or the relationship between multiplication and division. Illustrate and explain the calculation by using equations, rectangular arrays, and/or area models.


```{proof}
Let $a$ and $b$ be integers with $b>0$. We will first prove the existence of integers $q$ and $r$ such that $$a=bq+r, \quad \mbox{with } 0\leq r <b$$ and then prove that these integers are unique.
	
**Existence.** If $a\geq 0$, then we can let
$$S= \left\{ a-kb \vert  \: a-kb\geq 0 \mbox{ and $k$ is a non-negative integer}\right\}.$$
Since $a\geq 0$, we see that $a=a-0\cdot b$ is an element of $S$ and so $S$ is non-empty. We also see that $S \subseteq [0,a]$. 
	
So by the Well-Ordering Property of the Integers, there is a smallest element of $S$ which we will call $r$. Since $r\in S$, there exists a non-negative integer $q$ such that $r=a-qb$. 
	
Since $r$ is the smallest element of $S$ and since $a-(q+1)b<a-qb=r$, we see that $a-(q+1)b<0$, otherwise it would be a smaller element of $S$. 
	
This implies that $r=a-qb<b$. Therefore we have the existence of the $q$ and $r$.
	
If $a<0$, using the above process we can find non-negative integers $q$ and $r$, with $0\leq r<b$, such that $-a=bq+r$. So $a=b(-q)+r$ satisfies the existence statement. 
	
**Uniqueness.** In order to prove the uniqueness of the quotient and remainder, we assume that there are two pairs of integers $(q_1,r_1)$ and $(q_2,r_2)$ such that 
$$a= b q_1 + r_1 \mbox{ and } a= b q_2 + r_2, \mbox{ with } 0\leq r_1<b \mbox{ and } 0\leq r_2 <b.$$
This implies that $b q_1 + r_1 = b q_2 + r_2$ and by rearranging the terms we have that $$b (q_1 - q_2) = r_2-r_1.$$  If $0 \leq r_1 \leq r_2<b$, then $0 \leq b(q_1-q_2)=r_2-r_1<b$. So $q_1-q_2$ is an integer in $[0,1)$ and so must be $0$. Therefore, $q_1=q_2$ and so $r_1=r_2$. Therefore, the quotient and remainder are unique.
```

It is through a repetition of a variation of this division algorithm that we are able to divide integers with the standard long division algorithm. Look at various ways that we can find the quotient of $284$ by $13$. In particular compare the division algorithm in Theorem \@ref(thm:division-algorithm), the standard long division algorithm, the area model, and other models with which you are familiar.



### Composing and Decomposing Integers

Numerical fluency involves the ease at which one is able to break apart and recombined numbers under addition and/or multiplication. For addition, the fluency focuses on the tens using items like ten frames. For multiplication, it is a fluency with multiplication facts up to 12s, with flexibility and reversibility.

>#### Related Content Standards {-}
>+ (4.OA.4) Find all factor pairs for a whole number in the range 1-100. Recognize that a whole number is a multiple of each of its factors. Determine whether a given whole number in the range 1-100 is a multiple of a given one-digit number. Determine whether a given whole number in the range 1-100 is prime or composite.


```{definition}
Let $a$ and $b$ be integers with $b\neq 0$. We say that $b$ divides $a$ if there exists a $q\in \mathbb{Z}$ such that $a=bq$. In this situation we also say that 

+ $a$ is a multiple of $b$

+  $b$ is a divisor (or factor) of $a$

An integer $n>1$ is called prime if its only positive integer divisors are itself and $1$.
```

Note that 1 is not a prime number, since prime numbers are defined to be integers greater than 1. The primary reason behind this is to make the prime factorization of the integers unique.

So the prime numbers less than 100 are 2, 3, 5, 7, 11, 13, 17, 19, 23, 29, 31, 37, 41, 43, 47, 53, 59, 61, 67, 71, 73, 79, 83, 89, and 97.

### Greatest Common Divisor and Least Common Multiple

```{definition}
Let $a$ and $b$ be integers, not both zero. We say that the **greatest common divisor** (also called the **greatest common factor**) of $a$ and $b$, denoted $\mathrm{gcd}(a,b)$, is the natural number $d$ such that

+ $d$ divides both $a$ and $b$, and
+ if $u$ divides both $a$ and $b$ for some integer $u$, then $u\leq d$.

Let $a$ and $b$ be integers, not both zero. We say that the **least common multiple** of $a$ and $b$, denoted $\mathrm{lcm}(a,b)$, is the natural number $m$ such that

+ $m$ is a multiple of both $a$ and $b$, and
+ if $u$ is a multiple of both $a$ and $b$ for some integer $u$, then $u\geq m$.

```

>#### Related Content Standards {-}
>+ (6.NS.4) Find the greatest common factor of two whole numbers less than or equal to 100 and the least common multiple of two whole numbers less than or equal to 12. Use the distributive property to express a sum of two whole numbers 1-100 with a common factor as a multiple of a sum of two whole numbers with no common factor.        


There are many different options for finding the greatest common divisor and least common multiple of two numbers. If the numbers are small, one can find the greatest common divisor by listing out all of the possible factors of each number and choosing the largest one in common. Another option is using the factor trees of the two numbers, based on the prime factorizations of the numbers. Similarly, for small numbers one can list out the multiples of each number until you find a common multiple, or you can use information from the numbers' prime factorizations. When the numbers are much larger, one can develop other techniques to create algorithms for finding the greatest common divisor and least common multiple. Many of these techniques are based on the following theorem.

```{theorem, integers-division-gcd}
If $a,b,q,r$ are integers, with $a$ and $b$ not both zero, and if $$a=bq+r,$$ then $\mathrm{gcd}(a,b)=\mathrm{gcd}(b,r)$.
```

```{proof}
Since $a$ and $b$ are not both zero, we let $d=\mathrm{gcd}(a,b)$. This means that there are integers $m$ and $n$ such that $a=md$ and $b=nd$. Therefore, 
$$r=a-bq=md-(nd)q = d\cdot (m-nq)$$ and so $d$ divides $r$ and $d$ divides $b$. Since $d$ is a common divisor of $b$ and $r$, we see that $$\mathrm{gcd}(a,b)\leq \mathrm{gcd}(b,r).$$ Since $a=bq+r$, any common divisor of $b$ and $r$ must also be a divisor of $a$, we have that $\mathrm{gcd}(b,r)$ is a common divisor of $a$ and $b$ and so $$\mathrm{gcd}(b,r) \leq \mathrm{gcd}(a,b).$$ Therefore, $\mathrm{gcd}(a,b)=\mathrm{gcd}(b,r)$.
```


```{theorem, name="Euclidean Algorithm"}
Let $a$ and $b$ be two positive integers with $a>b$. Then the greatest common divisor of $a$ and $b$ can be found using the following algorithm. 

+ [Step 1] Apply the division algorithm to find integers $q$ and $r$ such that $a=bq+r$ with $0\leq r<b$.
+ [Step 2] If $r=0$, then $b$ is $\mathrm{gcd}(a,b)$.
+ [Step 3] While $r>0$, replace define $a:=b$ and $b:=r$ and apply the division algorithm to the new $a$ and $b$ to generate a new pair of integers $q$ and $r$. Once $r=0$, then the value of $b$ is the greatest common divisor of the original integers $a$ and $b$.

```


In order to better understand the algorithm, we will find the greatest common divisor of $924$ and $260$. We start by letting $a=924$ and $b=260$. From the division algorithm we find that
$$924 = 3 \cdot 260+144.$$ We now let $a=260$ and $b=144$ and use the division algorithm again to find that $$260=1\cdot 144 + 116.$$ Continuing this process we see that $\mathrm{gcd}(924,260)=4$. 
\begin{align*}
144 &=1\cdot 116 + 28 \\
116 &= 4 \cdot 28 + 4 \\
28 &= 7\cdot 4 + 0 
\end{align*}



### Fundamental Theorem of Arithmetic

The most common way to find the greatest common divisor and least common multiple of two numbers is dependent upon the numbers' prime factorizations. This ability to write a number uniquely as a product of primes is called the Fundamental Theorem of Arithmetic, which we will prove in this section.

We say that two numbers are **relatively prime** if their greatest common divisor is $1$.


```{theorem, relatively-prime}
Natural numbers $a$ and $b$ are relatively prime if and only if there exist integers $s$ and $t$ such that $sa+tb=1$.
```

```{proof}
Suppose that $a$ and $b$ are relatively prime. From the proof of the Euclidean algorithm we can see that for integers $a$ and $b$ that there exists integers $s$ and $t$ such that $sa+tb=\mathrm{gcd}(a,b)=1$.
	
Conversely, if there exist integers $s$ and $t$ such that $sa+tb=1$, we know that $d=\mathrm{gcd}(a,b)$ divides $a$ and $b$ and so would also divide $1$. Since the only integers that divide 1 are $1$ and $-1$ and since $\mathrm{gcd}(a,b)\geq 1$, we have that $\mathrm{gcd}(a,b)=1$.
```

```{lemma, prime-lemma}
If $p$ is a prime number and $a$ and $b$ are integers such that $p$ divides $ab$, then $p$ divides $a$ or $p$ divides $b$.
```

```{proof}
In order to prove the Lemma, we will prove that the negation is false. Suppose that $a$ is relatively prime to $p$ and $b$ is relatively prime to $p$. Then from Theorem \@ref(thm:relatively-prime) we have integers $s$, $t$, $u$, and $v$ such that $sa+tp=1$ and $ub+vp=1$. Then 
$$1= 1 \cdot 1 = (sa+tp)\cdot (ub+vp) = (su)(ab)+(sav+tub+tvp)p$$ and so $ab$ and $p$ are relatively prime, implying that $p$ does not divide $ab$. 
```


```{theorem, name="Fundamental Theorem of Arithmetic"}
Every integer $n>1$ can be factored as a product of prime numbers, and except for the order of the factors, this prime factorization is unique. 
```


```{proof}
Let $n>1$ be an integer. If $n$ is prime, we have a prime factorization of $n$. If $n$ is not prime, then there are integers $n_1$ and $n_2$ such that $n=n_1 n_2$ and we can assume without loss of generality that $1<n_1\leq n_2<n$. If either of these integers are composite, then they also can be factored into factors strictly smaller than them. This process can be continued, but it cannot be continued indefinitely since $n$ is a finite number. At the conclusion of the process, $n$ will have a prime factorization. We now only need to show uniqueness of this factorization.
	
Assume that $n$ has two distinct prime factorizations,
$$n=p_1 p_2 \cdots p_j \quad \mbox{and} \quad n=q_1 q_2 \cdots q_k$$ and we can assume without loss of generality that $j\leq k$.
	
Since $p_1$ divides $n$, a generalization of Lemma \@ref(lem:prime-lemma) gives us that $p_1$ must divide one of the $q_i$. By reordering the primes we can label this $q_i$ as $q_1$. Since both $p_1$ and $q_1$ are prime, they must be equal. We can continue this process with $p_2 \cdots p_j$ and $q_2\cdots q_k$ to find that $p_2=q_2$ and continuing the process to equate each of the $p_i$ with a $q_i$. If some of the $q_i$ remain after this process, these would be factors of $1$ and therefore could not be prime. Therefore, $j=k$ and we see that the two prime factorizations are just a reordering of each other.
```





### Exercises

1. Let $n\in \mathbb{Z}$. What are the possible values for the remainder of $n^2$ with a divisor of $3$? Justify your result.
		
1. Find the $q$ and $r$ from the Division Algorithm for the given values of $a$ and $b$.

      a. $a=0$, $b=14$
      a. $a=-42$, $b=14$
      a. $a=9$, $b=14$

1. Use the Division Algorithm to prove that for positive integers $m$ and $n$, with $m > n$, then exactly one of the following statements must be true:

      + There is a positive integer $k$ such that $m = kn$
      + There is a positive integer $k$ such that $kn < m < (k + 1)n$

	
1. Use the Euclidean Algorithm to find $\mathrm{gcd}(4368,31050)$.

1. Create a computer program or spreadsheet based on the Euclidean Algorithm that you can use to find the greatest common divisor of two integers.
		
1. Prove that for all integers that $\mathrm{gcd}(a,b)\cdot \mathrm{lcm}(a,b) = |ab|$. (Hint: Use the prime factorizations of $a$ and $b$.)
		


## Polynomial Rings

Another set of examples of integral domains in the secondary curriculum are polynomial rings. We let $$\mathbb{R}[x]= \left\{ a(x)=a_0 + a_1 x + a_2 x^2 + \cdots + a_n x^n =\sum_{j=0}^n a_j x^j  \middle \vert a_j \in \mathbb{R} \right\}$$

The concept of a polynomial can be generalized to not just having real number coefficients. For instance, we will sometimes require that the polynomials have integer coefficients or possibly coefficients that are complex numbers. So we will generalize the definition of a polynomial to be able to be used in these cases.


```{definition}
Let $R$ be a ring. We define a polynomial $f(x)$ with coefficients in $R$ to be an infinite formal sum $$\sum_{j=0}^\infty a_j x^j = a_0 + a_1 x+ a_2 x^2 + \cdots + a_n x^n + \cdots$$ where $a_j\in R$ for all $j$ and $a_j =0$ for all but finitely many values of $j$.
	
We denote the set of such polynomials as $R[x]$.
```

Since the $a_j=0$ for all but finitely many values of $j$, by the well-ordering principle, there is a largest value of $j$ for which $a_j \neq 0$. We call this largest value of $j$ the **degree** of the polynomial. If all of the coefficients are $0$, we say the polynomial is degree zero.

So we can define the function "$\mathrm{degree}$" that maps the non-zero polynomials to the natural numbers by
$$\mathrm{degree}\left(\sum_{j=0}^\infty a_j x^j\right) \mbox{ is the largest value of $j$ for which $a_j$ is non-zero}.$$ We call the $a_j$ that corresponds to this value of $j$ the **leading coefficient**.

>#### Related Content Standards {-}
>+ (HSA.APR.1) Understand that polynomials form a system analogous to the integers, namely, they are closed under the operations of addition, subtraction, and multiplication; add, subtract, and multiply polynomials. 


If $a(x)=\sum_{j=0}^n a_j x^j$ and $b(x)=\sum_{j=0}^m b_j x^j$ are two elements of $R[x]$, with $n\leq m$, then we define 
$$a(x)+b(x) := \sum_{j=0}^m (a_j+b_j) x^j$$ where $a_j:=0$ for $n<j\leq m$. From this definition, we see that addition is closed. We also see that addition is associative from the associativity of addition on $R$. Since $R$ is a ring, addition is commutative on $R$ and so addition is commutative on $R[x]$. We also see that the zero polynomial, $0(x)=0$, is an additive identity for the set. So $(R[x],+)$ is an Abelian group.

Multiplication on $R[x]$ is more challenging to write in a compact form. We will assume that the indeterminate, $x$, commutes multiplicatively with every element in $R$, i.e. $ax=xa, \: \forall a\in R$ (the set of all elements of a ring that commute with every element of the ring is called the center of the ring). We can then define multiplication as

\begin{align*}
    a(x) \cdot b(x) &:= \left( \sum_{j=0}^n a_j x^j \right) \cdot \left(\sum_{k=0}^m b_k x^k\right) 
    = \sum_{j=0}^n \left( a_j x^j \left(\sum_{k=0}^m b_k x^k\right) \right) \\
    &= \sum_{j=0}^n \sum_{k=0}^m \left(a_j x^j b_k x^k\right) 
    = \sum_{j=0}^n \sum_{k=0}^m \left(a_j b_k\right) x^{(j+k)}
\end{align*}
and we see that $R[x]$ is closed under multiplication.

Unless $R$ is a commutative ring, we must maintain the order of the coefficients. The associativity of multiplication and distributive properties are inherited from $R[x]$. If the ring $R$ has a multiplicative identity, then $R[x]$ also has a multiplicative identity of the polynomial of the multiplicative identity. Finally, if $R$ is a commutative ring, then multiplication on $R[x]$ is commutative.


```{theorem}
The set $R[x]$ of polynomials with coefficients from a ring $R$ with the binary operations of polynomial addition and multiplication is a ring. 
```


One of the main activities that we do with polynomials in the secondary curriculum is to treat the polynomials as functions and evaluate them for various values. 

The following theorem gives us the mathematical justification for such evaluations.


In this theorem we use a field $F$ and a subfield $E$, with is a subset of $F$ that also satisfies the field properties inherited from $F$.  The reason that we use the field $E$ and the subfield $F$ is that we often have polynomials with whose coefficients are from one field and the point at which we are evaluating from a larger field. For instance, we often look at quadratic polynomials with real coefficients and determine that they have complex zeros. In this instance, the subfield would be $\mathbb{R}$ and the larger field would be $\mathbb{C}$.

```{theorem}
Let $F$ be a subfield of a field $E$ and let $\alpha$ be any element of $E$. The map $\phi_\alpha: F[x] \rightarrow E$ defined by $$\phi_\alpha(a_0 + a_1 x + a_2 x^2 + \cdots + a_n x^n) = a_0 + a_1 \alpha + a_2 \alpha^2 + \cdots + a_n \alpha^n$$ is a ring homomorphism from $F[x]$ into $E$.
```

We go ahead and include the proof of this theorem below to demonstrate how these evaluation homomorphisms work and to demonstrate certain properties of polynomial functions. 

```{proof}
Let $F$ be a subfield of a field $E$ and let $\alpha$ be any element of $E$. We also let $a(x)$ and $b(x)$ be generic elements of $F[x]$ and label them as 
$$a(x) = \sum_{j=0}^\infty a_j x^j \mbox{ and } b(x)=\sum_{j=0}^\infty b_j x^j$$ where the $a_j$ and the $b_j$ are zero for all but finitely many values of $j$.
	
Then $\phi_\alpha \left(a(x)\right) = \sum_{j=0}^\infty a_j \alpha^j$ and $\phi_\alpha\left(\left(b(x)\right)\right) = \sum_{j=0}^\infty b_j \alpha^j$, which are elements of $E$ since addition and multiplication on $E$ are closed and only finitely many elements of each of the sums are non-zero.
	
Using the definition of addition of two polynomials, we have that $a(x)+b(x)$ is a polynomial that is written as $$a(x)+b(x) = \sum_{j=0}^\infty (a_j+b_j) x^j.$$
Therefore, 
$$\phi_\alpha \left(a(x)+b(x)\right) = \sum_{j=0}^\infty (a_j+b_j) \alpha^j$$ and by rearranging the finite number of terms using the distributive property and that both addition and multiplication are commutative in $E$ we have that 
$$\phi_\alpha \left(a(x)+b(x)\right) = \sum_{j=0}^\infty a_j \alpha^j + \sum_{j=0}^\infty b_j \alpha^j = \phi_\alpha \left(a(x)\right) + \phi_\alpha \left(b(x)\right). $$
	
Similarly, one finds that 
$$\phi_\alpha \left(a(x) \cdot b(x)\right) = \phi_\alpha (a(x)) \cdot \phi_\alpha(b(x)).$$ 
```

You should keep in mind that the operations are using the same symbol for operations in two different rings. For the case of $a(x)+b(x)$, the addition is in the polynomial ring $F[x]$, while $\phi_\alpha \left(a(x)\right) + \phi_\alpha \left(b(x)\right)$ is addition in $E$.

We say that an element $a \in E$ is a **zero of a polynomial** $p(x)\in F[x]$ if $$\phi_a ( p(x))=0.$$

We would really like to have the property that for all $a,b \in R$, $a$ and $b$ are the only zeros of the polynomial $(x-a)(x-b)$. 
In order for that to be true, we need $R[x]$ to be an integral domain. However, we first need some properties related to the degree of polynomials.

```{theorem}
Let $R$ be an integral domain and let $a(x)$ and $b(x)$ be elements of $R[x]$. Then
$$\mathrm{degree}(a(x)+b(x))\leq \mathrm{max}\left( \mathrm{degree}(a(x)), \mathrm{degree}(b(x))\right) \mbox{, and}$$
$$\mathrm{degree}\left( a(x) b(x)\right) = \mathrm{degree}(a(x))+\mathrm{degree}(b(x)).$$
```

```{proof}
Let $a(x)=\sum_{j=0}^n a_j x^j \in R[x]$ with degree $n$ and $b(x)=\sum_{j=0}^m b_j x^j\in R[x]$ with degree $m$. We can assume without any loss of generality that $n\leq m$. 

If $n<m$, $a(x)+b(x) = \sum_{j=0}^m (a_j+b_j) x^j$, where $a_j=0$ for $n<j\leq m$, we see that $\mathrm{degree} (a(x)+b(x)) =m$. If $n=m$ the sum also has degree $m$, unless the two leading coefficients are additive inverses. Therefore, $$\mathrm{degree}(a(x)+b(x))\leq \mathrm{max}\left( \mathrm{degree}(a(x)), \mathrm{degree}(b(x))\right).$$ 

Since 
$$a(x) \cdot b(x)  = \sum_{j=0}^n \sum_{k=0}^m \left(a_j b_k\right) x^{(j+k)}$$ we see that the leading term of $a(x)\cdot b(x)$ is $a_n b_m$. This is non-zero because $R$ is an integral domain. So the degree of $a(x) \cdot b(x)$ is $m+n$. And we have that $$\mathrm{degree}\left( a(x) b(x)\right) = \mathrm{degree}(a(x))+\mathrm{degree}(b(x)).$$
```

We are now ready to prove that $R$ being an integral domain implies the $R[x]$ is an integral domain.

```{theorem, Polynomial-Integral-Domain}
If $R$ is an integral domain, then $R[x]$ is an integral domain.
```

```{proof}
Assume that there are two polynomials, $a(x)=\sum_{j=0}^n a_j x^j$ and $b(x)=\sum_{j=0}^m b_j x^j$ in $R[x]$, such that $a_n\neq 0$ and $b_m\neq 0$. Then the product of the polynomials has degree of $n+m$. If $n+m>0$, then the product cannot be the zero polynomial. If $n+m=0$, then $n$ and $m$ are both zero and $a(x)=a_0$, $b(x)=b_0$, and $a(x)\cdot b(x)=a_0 b_0$. Since $a_0\neq 0$ and $b_0\neq 0$, since $R$ is an integral domain, the product is non-zero. 
```

### Exercises


1. In $\mathbb{Z}_{12}$, what are the zeros of the polynomial $6-5x+x^2$?

1. Use the various models of multiplication presented in Section \ref{sec:integer-representation} to add and multiply the following pairs of polynomials:

      a. $a(x)=-2x+3$ and $b(x)=x-1$
      a. $a(x)=2x^2-4x+1$ and $b(x)=x^3-2$
      a. Which models transition well to polynomials and which do not?



## Properties of Polynomial Rings

The properties of irreducible (prime) elements and factorizations is very important in the study of the integers and is dependent upon the integral domain properties. We have also seen that certain polynomial rings are also integral domains. In this section we will see how the factorization of integers into prime factors can be extended to the ring of polynomials. Throughout this section we will let $S$ be a general integral domain (you can think of it as $\mathbb{Z}$, $\mathbb{Q}$, $\mathbb{R}$, or $\mathbb{C}$) and let $S[x]$ be the ring of polynomials with coefficients from $S$.

### Polynomial Division

We begin this generalization by finding a generalization of the division algorithm. We see from the Division Algorithm for Integers that we need to have some type of ordering for the elements of our integral domain. We will call such an ordering a Euclidean function.

```{definition}
Let $R$ be an integral domain. A **Euclidean function** on $R$ is a function $f$ from $R\setminus \{0\}$ to the non-negative integers satisfying the fundamental division-with-remainder property.

+ If $a$ and $b$ are in $R$ and $b$ is non-zero, then there exist $q$ and $r$ in $R$ such that $a=bq+r$ and either $r=0$ or $f(r)<f(b)$.

An integral domain for which a Euclidean function exists is called a Euclidean domain.
```

Since $S$ is an integral domain, we see that $S[x]$ is also an integral domain (Theorem \@ref(thm:Polynomial-Integral-Domain)). We can now use the degree function of the polynomial as a Euclidean function and state a Division Algorithm for Polynomials that makes $S[x]$ a Euclidean domain, when $S$ is a field.



```{theorem, Division-Algorithm-Poly, name="Division Algorithm for Polynomials"}
Let $S$ be a field. Suppose that $a(x)$ and $b(x)$ are polynomials in $S[x]$ and that $b(x)$ is not the zero polynomial. Then there are unique $q(x)$ and $r(x)$ in $S[x]$ such that 
$$a(x)=b(x)\cdot q(x) + r(x)$$ and $r(x)$ is the zero polynomial or the degree of $r(x)$ is less than the degree of $b(x)$.
```

The reason that we need $S$ to be a field is to allow for division of the coefficients in the division algorithm. For example in $\mathbb{Z}[x]$, if $a(x)=x^2$ and $b(x)=3x$, the $r(x)$ in the division algorithm would need to be $3x$.

```{proof}
Let $S$ be a field and let $a(x)$ and $b(x)$ be polynomials, with $b(x)$ not the zero polynomial.
	
Let 
$$T= \left\{ a(x)-b(x)\cdot s(x) | s(x) \in S[x]\right\}$$ and from the well-ordering property we can choose an $r(x)$ in $T$ with minimal degree. So $$a(x)=b(x) q(x) + r(x)$$ for some $q(x)\in S[x]$.
	
Let $m=\mathrm{degree}(b(x))$ and $n=\mathrm{degree}(r(x))$, and let $b_m$ be the leading coefficient of $b(x)$ and $r_n$ be the leading coefficient of $r(x)$. If $m\geq n$, 
$$a(x)-b(x)\left(q(x)- \frac{r_n}{b_m} x^{m-n}\right)= r(x)-\frac{r_n}{b_m} x^{m-n} b(x) $$ which is an element of $T$ with lower degree than $r(x)$, contradicting the choice of $r(x)$ to be an element of $T$ with minimal degree. Therefore, $\mathrm{degree}(r(x))<\mathrm{degree}(b(x))$.
	
For the uniqueness of $q(x)$ and $r(x)$, we assume that there are two pairs of such functions such that 
$$a(x)=b(x) q_1(x)+r_1(x) \quad \mbox{and} \quad a(x)=b(x) q_2(x)+r_2(x)$$ with the degrees of $r_1(x)$ and $r_2(x)$ being less than the degree of $b(x)$. Subtracting the two equations and rearranging terms gives us $$b(x) \left( q_1(x)-q_2(x) \right) = r_2(x)-r_1(x).$$ Since the degree of the polynomial on the right side of the equation is less than the degree of $b(x)$, $(q_1(x)-q_2(x))$ must be the zero polynomial. So $q_1(x)= q_2(x)$ and thus implying that $r_1(x)=r_2(x)$. And we have proven the uniqueness of the $q(x)$ and $r(x)$.
```


>#### Related Content Standards {-}
>+ (HSA.APR.6) 	Rewrite simple rational expressions in different forms; write $\frac{a(x)}{b(x)}$ in the form $q(x)+\frac{r(x)}{b(x)}$, where $a(x)$, $b(x)$, $q(x)$, and $r(x)$ are polynomials with the degree of $r(x)$ less than the degree of $b(x)$, using inspection, long division, or, for the more complicated examples, a computer algebra system. 
	

Take some time and work through some examples of long division of polynomials and compare these with the procedures in the long division of integers. It is through these connections between the two processes that students are able to understand and retain the knowledge of the long division algorithm for polynomials.

We can now extend the properties of factors and prime (or irreducible) elements of the integers to the polynomial ring.

```{definition}
Let $S$ be an ring and $S[x]$ be the corresponding ring of polynomials. Let $a(x)$ and $b(x)$ be polynomials in $S[x]$ with $b(x)$ not the zero polynomial. We say that $b(x)$ divides $a(x)$ if there exists a $q(x)$ in $S[x]$ such that $a(x)=b(x)\cdot q(x)$. In this situation we also say that

+ $a(x)$ is a multiple of $b(x)$,
+ $b(x)$ is a divisor of $a(x)$, and 
+ $b(x)$ is a factor of $a(x)$.

```


We can now state the prove the Remainder Theorem and Factor Theorem for polynomials.

```{theorem, name="Remainder Theorem"}
Let $S$ be a field ($\mathbb{Q}, \mathbb{R}, \mathbb{C}$) and suppose that $a(x)$ is a polynomial of degree greater than or equal to $1$ in $S[x]$, and let $c\in S$. Then $a(c):=\phi_c(a(x))$ is the remainder obtained by dividing $a(x)$ by $x-c$, where $\phi_c$ is the evaluation function.
```

```{proof}
Applying the Division Algorithm to $a(x)$ and $x-c$ we see that there is a unique $q(x)$ and $r(x)$ such that $a(x)=q(x)\cdot (x-c)+r(x)$ with the degree of $r(x)$ less than the degree of $x-c$. Therefore, $r(x)$ is a constant polynomial that we will call $r$. Since the evaluation function is a ring homomorphism we have that $$a(c):= \phi_c(a(x))=\phi_c(q(x))\cdot \phi_c(x-c) + \phi_c(r).$$
	
Since $\phi_c(x-c)=c-c=0$ and since $0\cdot \phi_c(q(x))=0$, we have that $a(c)$ is the remainder.
```

This leads us directly to the Factor Theorem since $a(x)=q(x)\cdot (x-c) + a(c)$.

```{theorem, name="Factor Theorem"}
Let $S$ be a field and suppose that $a(x)$ is a polynomial of degree at least $1$. Then $x-c$ is a factor of $a(x)$ if and only if $a(c)=0$.
```



>#### Related Content Standards {-}
>+ (HSA.APR.2) 	Know and apply the Remainder Theorem: For a polynomial $p(x)$ and a number $a$, the remainder on division by $x-a$ is $p(a)$, so $p(a)=0$ if and only if $(x-a)$ is a factor of $p(x)$.        



### GCF and LCM for Polynomials

```{definition}
Let $a(x)$ and $b(x)$ be in $S[x]$, not both the zero polynomial. We say that a **greatest common divisor** (also called the **greatest common factor**) of $a(x)$ and $b(x)$ is a polynomial $d(x)$ such that

+ $d(x)$ divides both $a(x)$ and $b(x)$, and
+ if $u(x)$ divides both $a(x)$ and $b(x)$ for some polynomial $u(x)$, then $u(x)$ divides $d(x)$.

```

Note that there is not a single greatest common divisor of two polynomials since constant multiples of a greatest common divisor may also be greatest common divisors. In the situation that $S$ is a field, we can multiply a greatest common divisor by the multiplicative inverse of its leading coefficient and we now have a monic polynomial (a polynomial with leading coefficient of the multiplicative identity) that is a greatest common divisor. This monic polynomial is unique and so we define $\mathrm{gcd}(a(x),b(x))$ to be the **unique monic greatest common divisor of $a(x)$ and $b(x)$**.

Following the line of proofs of the integers, we can prove a theorem analogous to Theorem \@ref(thm:integers-division-gcd) for polynomials.

```{theorem, polynomial-division-gcd}
If $a(x)$, $b(x)$, $q(x)$, and $r(x)$ are polynomials in $S[x]$ for a field $S$, with $a(x)$ and $b(x)$ not both the zero polynomial, and if $$a(x)=b(x)q(x)+r(x),$$ then $\mathrm{gcd}(a(x),b(x))=\mathrm{gcd}(b(x),r(x))$.
```

This theorem is the basis for a generalization of the Euclidean Algorithm to the polynomials.


```{theorem, name="Euclidean Algorithm for Polynomials"}
Let $a(x)$ and $b(x)$ be two polynomials in $S[x]$ for a field $S$, with $\mathrm{degree}(a(x))>\mathrm{degree}(b(x))$. Then the greatest common divisor of $a(x)$ and $b(x)$ can be found using the following algorithm. 
	
+ [Step 1] Apply the division algorithm to find polynomials $q(x)$ and $r(x)$ such that $a(x)=b(x)q(x)+r(x)$ with $0\leq \mathrm{degree}(r(x))<\mathrm{degree}(b(x))$.
+ [Step 2] If $r(x)$ is the zero polynomial, then is $\mathrm{gcd}(a,b)$ is $b(x)$ times the multiplicative inverse in $S$ of its leading coefficient.
+ [Step 3] While $r(x)$ is not the zero polynomial, replace $a(x):=b(x)$ and $b(x):=r(x)$ and apply the division algorithm to the new $a(x)$ and $b(x)$ to generate a new pair of polynomials $q(x)$ and $r(x)$. Once $r(x)$ is the zero polynomial, the value of $b(x)$ is a greatest common divisor of the original polynomials $a(x)$ and $b(x)$ and  $\mathrm{gcd}(a,b)$ is this polynomial times the multiplicative inverse in $S$ of its leading coefficient. 

```

In order to better understand the Euclidean Algorithm for Polynomials we will use the algorithm to find the unique monic greatest common divisor of $a(x)=2x^5-x^4-41x^3+59x^2-43x+60$ and $b(x)=x^3+2x^2+x+2$.


The first step is to use the division algorithm to find the quotient and remainder of $a(x)$ divided by $b(x)$. One of the most efficient methods of finding this quotient and remainder (other than using a computer algebra system) is through long division of polynomials.


```{r, echo=FALSE, fig.align='center', out.width='75%', fig.cap=''}
knitr::include_graphics(
  'tikz/polynomial-division.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```



Therefore,
$$a(x)=b(x) \cdot (2x^2-5x-33) + (126x^2 + 126)$$ and so we know that $\mathrm{gcd}(a(x),b(x))= \mathrm{gcd}(x^3+2x^2+x+2, 126x^2+126)$. Using long division of polynomials we can find the new quotient and remainder for these two polynomials.

```{r, echo=FALSE, fig.align='center', out.width='50%', fig.cap=''}
knitr::include_graphics(
  'tikz/polynomial-division2.png',
  auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE),
  dpi = NULL,
  error = getOption("knitr.graphics.error", TRUE)
  )
```

So, $$x^3+2x^2+x+2 = (126x^2+126)\left(\frac{1}{126}x+\frac{1}{63}\right)+ 0$$ and we see that $126x^2+126$ is a greatest common divisor of $a(x)$ and $b(x)$. Therefore, the unique monic greatest common divisor of $a(x)$ and $b(x)$ is $x^2+1$.


### Irreducible Polynomials

We now turn our attention to studying the analogy to prime numbers in the polynomial ring.


```{definition}
An element in an integral domain, $D$, is called a **unit** of $D$ if it has a multiplicative inverse. A non-zero element, $p$, of $D$ is called an \emph{irreducible} of $D$ if any factorization $p=ab$ in $D$ either $a$ or $b$ is a unit.
```

When $S$ is a field, we see that the units of $S[x]$ are the constant polynomials since the degree of the product of two polynomials in $S[x]$ is equal to the sum of the degrees of the factors.

Notice that since $\sqrt{2}$ is irrational that $x^2-2$ is irreducible in $\mathbb{Q}[x]$ but has the factorization of $x^2-2=(x-\sqrt{2})(x+\sqrt{2})$ in $\mathbb{R}[x]$ and $\mathbb{C}[x]$. This means that the set of irreducible polynomials depends upon the field from which the coefficients reside.

The following is an essential theorem in understanding factorization of polynomials and is one of the main consequences of the concept of irreducible elements. However, the proof of the theorem requires building up significant machinery in abstract algebra and so we will not include it here.

```{theorem}
Let $p(x)$ ba an irreducible polynomial in $S[x]$. If $p(x)$ divides $a(x)b(x)$ for $a(x),b(x)\in S[x]$, then either $p(x)$ divides $a(x)$ or $p(x)$ divides $b(x)$.
```

With this property involving the factorization of polynomials that we use on a regular basis we can also prove a property similar to that of the Fundamental Theorem of Arithmetic for integers.

```{theorem, polynomial-factorization}
Let $S$ be a field. Then every polynomial of positive degree in $S[x]$ can be represented as a product
$$ c p_1(x) p_2(x) \cdots p_k(x)$$ where $c\in S$ and the $p_i(x)$ are monic irreducible polynomials in $S[x]$. This representation is unique up to reordering the irreducible polynomials.
```


The proof of this theorem is very similar to the method used to prove the Fundamental Theorem of Arithmetic, and so we will not include the details here. However, this theorem demonstrates the importance for understanding the irreducible polynomials in $\mathbb{R}[x]$ and $\mathbb{C}[x]$ in the process of understanding the properties of polynomials with real and complex coefficients.

The main result in this area is the Fundamental Theorem of Algebra.

```{theorem, name="The Fundamental Theorem of Algebra"}
The irreducible monic polynomials in $\mathbb{C}[x]$ are of the form $x-c$.
```

There are many different methods to prove this result, but the primary method uses the machinery of complex calculus. 


>#### Related Content Standards {-}
>+ (HSN.CN.9) Know the Fundamental Theorem of Algebra; show that it is true for quadratic polynomials.
	 



### Exercises

1. Find a prime factorization of the following polynomials in $\mathbb{Z}[x]$, $\mathbb{Q}[x]$, $\mathbb{R}[x]$, and $\mathbb{C}[x]$.

      a. $4x^3-2x^2+2x-1$
      a. $x^3-1$
      a. $6x^2-13x-5$
      a. $x^5-2x^4-2x^3+4x^2-3x+6$

1. Explain how the domain of the variable $x$ affects the number of solutions to the equation $$(x^4-1)(4x^4-9x^2+2)=0.$$
		

1. Find the unique **monic** prime factorization of the polynomials in $\mathbb{Q}[x]$, $\mathbb{R}[x]$, and $\mathbb{C}[x]$.

      a. $x^4+4x^2+4$
      a. $7x^4-28$
      a. $2x^4-5x^2-3$
      a. $9x^4+440 x^2-49$
      a. $x^{106} -x^{100}$
      a. $4x^4+12x^3-86x^2-92x-48$
      a. $x^4-x^2-2$
      a. $x^2-\pi$
      a. $2x^4-x^3-19x^2+2x+30$
      a. $a x^2 + bx + c  $ based on the various possible values of $a$, $b$, and $c$ with $a,b,c \in \mathbb{Z}$.



1. Prove Theorem \@ref(thm:polynomial-division-gcd) using the proof of Theorem \@ref(thm:integers-division-gcd) as an outline.

1. Prove that if $f$ is a Euclidean function on an integral domain $R$, then $f(a)\leq f(b)$ for all non-zero $a$ and $b$ in $R$.
		
1. Find the value $c$ so that $(x-3)$ is a factor of the polynomial $p(x)$. $$p(x)=cx^3-15x-68$$
		
1. Find the monic greatest common divisor of the polynomials $a(x)=4x^5+12x^4+20x^3+16x^2+16x+4$ and $b(x)=6x^5+12x^4+18x^3+12x^2+12x$.
		

## Rational Expressions

In the same way that one can extend the integers to the field of rational numbers by defining addition and multiplication on pairs of integers, one can extend the polynomial ring by defining addition and multiplication of pairs of polynomials and then defining a set of equivalence classes on this set.

>#### Related Content Standards {-}
>+ (HSA.APR.7)] Understand that rational expressions form a system analogous to the rational numbers, closed under addition, subtraction, multiplication, and division by a non-zero rational expression; add, subtract, multiply, and divide rational expressions. 
	

Let $F$ be a field and let $F[x]$ be the integral domain of polynomials with coefficients from $F$. We define an equivalence relation on the set of rational expressions of polynomials as $\frac{p(x)}{q(x)}$ is equivalent to $\frac{a(x)}{b(x)}$ if and only if $p(x) \cdot b(x)=a(x)\cdot q(x)$.

Define $F(x)$ to be the set of equivalence classes of quotients of polynomials $\left[\frac{p(x)}{q(x)}\right]$ with $q(x)$ not being the zero polynomial under this equivalence relation. We then proceed with dropping the equivalence class notation to simplify the notation.  We define the following operations on $F(x)$.

$$\frac{p(x)}{q(x)} + \frac{f(x)}{g(x)} = \frac{p(x)\cdot g(x) + q(x) \cdot f(x)}{q(x)\cdot g(x)} \quad \mbox{ and } \quad \frac{p(x)}{q(x)} \cdot \frac{f(x)}{g(x)} = \frac{p(x) \cdot f(x)}{q(x)\cdot g(x)}$$

	
We call the set $F(x)$ the set of rational expressions with coefficients from $F$ and we label the corresponding set with operations as $\langle F(x),+,\cdot\rangle$, but usually refer to it as just $F(x)$.


We see that the operations are well-defined because the polynomials are well-defined and integral domain and therefore closed under addition and multiplication and do not have any zero divisors. Thus if neither $q(x)$ nor $g(x)$ are the zero polynomial, then neither is their product. This leads us to our next result that is analogous to the rational numbers.

```{theorem}
Let $F$ be a field. Then $\langle F(x),+,\cdot \rangle$ is a field.
``` 

```{proof}
We leave the proof as an exercise to help the reader become more familiar with this field.
```

This implies that the rational expressions with real valued coefficients, $\mathbb{R}(x)$ forms a field very similar to the rational numbers. We will explore this field in greater depth in Section \@ref(sec:rational-functions).

### Exercises

1. Prove that $\mathbb{R}(x)$ is a field.


	

